---
title: "Preliminary Sample-Size Analysis for CLAPS"
author: "Pablo Bernabeu and CLAPS collaborators"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
format:
  pdf:
    geometry: "margin=1in"
    linestretch: 1.25
    fig-cap-location: top
    fig-pos: "htbp"
    colorlinks: true
    code-overflow: wrap
    include-in-header:
      text: |
        \definecolor{subtitlegray}{gray}{0.35}
        \setkomafont{date}{\normalfont\normalsize\color{subtitlegray}}
        % Microtypographic character protrusion and font expansion for an
        % evener, less rivered right margin under justification.
        \usepackage{microtype}
        % Discourage widows and orphans (a lone line stranded at the top or
        % bottom of a page), which a technical report with many short
        % paragraphs is otherwise prone to.
        \widowpenalty=10000
        \clubpenalty=10000
        % Allow a little more stretch between words before over-full lines are
        % risked, so justification does not force awkward hyphenation.
        \setlength{\emergencystretch}{3em}
        % Wrap long lines inside code blocks at the margin (breakanywhere lets a
        % long path or token break even without a space) rather than letting
        % them run past the right edge.
        \usepackage{fvextra}
        \fvset{breaklines=true,breakanywhere=true}
        % Show the "Abstract" label, which KOMA suppresses by default. The
        % report follows APA 7 elsewhere, where the abstract is labelled.
        \KOMAoptions{abstract=on}
        % APA 7 caption style: bold label, italic title, applied to figures and tables.
        \usepackage{caption}
        \DeclareCaptionFormat{apa7}{#1\\[5pt]#3}
        \DeclareCaptionFormat{apa7par}{#1\par\vspace{5pt}#3}
        \captionsetup{format=apa7, labelsep=none, labelfont=bf, textfont=it,
          justification=raggedright, singlelinecheck=false, skip=6pt}
        \captionsetup[longtable]{format=apa7par, labelsep=none, labelfont=bf,
          textfont=it, justification=raggedright, singlelinecheck=false, skip=6pt}
        % Top-align figures that land on their own float page (avoids vertical centring).
        \makeatletter
        \setlength{\@fptop}{0pt}
        \setlength{\@fpsep}{12pt plus 1fil}
        \setlength{\@fpbot}{0pt plus 1fil}
        \makeatother
bibliography:
  - ../references.bib
csl: apa.csl
link-citations: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)

# Tidy per-cell Bayes-factor exceedance from the literature-anchored run,
# aggregated by scripts/aggregate_literature.R (one row per language x N x
# n_verbs x hypothesis). p_bf_primary is the proportion of replicates with BF10 >= 10
# (the preregistered detection threshold). Prose about which languages and
# sample sizes this covers is derived below rather than written by hand, since
# an earlier summary was aggregated before one language had finished and the
# report went on asserting that gap after the cells had landed.
exc <- read_csv("../outputs/design_summary_literature/bf_exceedance_literature.csv",
                show_col_types = FALSE) |>
  filter(n_verbs %in% c(40, 72))   # 12 and 20 verbs are well below the design's count
lit_languages <- sort(unique(exc$language))
lit_language_text <- if (length(lit_languages) > 2) {
  paste(paste(head(lit_languages, -1), collapse = ", "), "and", tail(lit_languages, 1))
} else paste(lit_languages, collapse = " and ")
lit_n_min <- min(exc$n_participants)
lit_n_max <- max(exc$n_participants)

focal <- c(H1a = "H1a_semantics_positive",
           H1b = "H1b_active_interaction_negative")

# Binding focal power = the smaller of the two focal exceedance probabilities,
# since the recommendation requires BOTH H1a and H1b to clear 0.80.
foc <- exc |>
  filter(hypothesis %in% focal) |>
  mutate(H = names(focal)[match(hypothesis, focal)]) |>
  select(language, n_participants, n_verbs, reps, H, p_bf_primary) |>
  pivot_wider(names_from = H, values_from = p_bf_primary) |>
  mutate(binding = pmin(H1a, H1b))

# Restrict the figure/recommendation to cells with (near) the full 50 replicates.
MIN_REPS <- 45L
foc_full <- filter(foc, reps >= MIN_REPS)

pct <- function(x) ifelse(is.na(x), "-", sprintf("%.0f%%", 100 * x))

# Recommendation: smallest N at which the binding focal power reaches 0.80,
# per language x verb count.
rec <- foc_full |>
  filter(binding >= 0.80) |>
  group_by(language, n_verbs) |>
  summarise(rec_N = min(n_participants), .groups = "drop")

# Convergence robustness for the cell cited in the text (English, 72 verbs, N = 50):
# binding focal power computed over all fits vs over the converged subset only.
cc <- exc |>
  filter(language == "English", n_verbs == 72, n_participants == 50,
         hypothesis %in% focal) |>
  summarise(all = min(p_bf_primary), conv = min(p_bf_primary_conv),
            n_conv = max(n_conv))

# The public report is deliberately independent of participant-level pilot data.
# It uses only disclosure-reviewed derivatives that are already represented in
# the published PDF: a raster figure and aggregate sample-composition counts.
pilot_composition <- read_csv(
  "../outputs/design_summary_pilot/pilot_sample_composition.csv",
  show_col_types = FALSE
)
pilot_figure <- "../outputs/figures/pilot_acceptability_by_gender_report.png"

# Pilot-grounded (v2) summaries, aggregated on the cluster by
# scripts/aggregate_pilot.R: per-cell joint power, fitted pilot
# parameters and the analytic detection ceilings. Reading them here keeps
# every simulation result in the report programmatic.
jp <- read_csv("../outputs/design_summary_pilot/joint_power_pilot.csv",
               show_col_types = FALSE)
pp <- read_csv("../outputs/design_summary_pilot/pilot_params_ceilings.csv",
               show_col_types = FALSE) |>
  mutate(language = factor(language, levels = c("English", "Turkish", "Norwegian"))) |>
  arrange(language)

pct0  <- function(x) sprintf("%.0f%%", 100 * x)
pct1  <- function(x) sprintf("%.1f%%", 100 * x)
num2  <- function(x) sprintf("%.2f", x)
mnum2 <- function(x) sprintf("$%.2f$", x)   # math mode, for a proper minus sign
ppv   <- function(lang, col) pp[[col]][pp$language == lang]

# Cells from the v2 grid are shown and cited only once they hold a minimum
# number of replicates; smaller cells are still accumulating.
MIN_REPS_V2 <- 10L
jp_a  <- filter(jp, mode == "assurance", reps >= MIN_REPS_V2)
jval  <- function(lang, N, col) jp_a[[col]][jp_a$language == lang & jp_a$n_participants == N]
jrow  <- function(lang, N) filter(jp_a, language == lang, n_participants == N)
jcount <- function(lang, N) {
  r <- jrow(lang, N)
  as.integer(round(r$p_joint * r$reps))
}
jratio <- function(lang, N) {
  r <- jrow(lang, N)
  sprintf("%d/%d", jcount(lang, N), r$reps)
}
jmcse <- function(lang, N) {
  r <- jrow(lang, N)
  sqrt(r$p_joint * (1 - r$p_joint) / r$reps)
}
max_n <- function(lang) max(jp_a$n_participants[jp_a$language == lang])
nor_cross     <- min(filter(jp_a, language == "Norwegian", p_joint >= 0.80)$n_participants)
tur_max_joint <- max(filter(jp_a, language == "Turkish")$p_joint)
# Smallest sample at which the interaction alone is adequately powered. The test is
# the exact one-sided 95% lower bound, not the point estimate: an earlier version
# took the first cell whose observed rate reached 0.80 and so reported Norwegian as
# adequate at 50 participants on the strength of 22 successes out of 24, a cell whose
# lower bound is 0.76. A sample size is only recommendable if the simulation is
# precise enough to support the recommendation.
lower95 <- function(k, n) if (n == 0) NA_real_ else stats::binom.test(k, n, alternative = "greater")$conf.int[1]
h1b_cross <- function(lang, col = "p_h1b", target = 0.80) {
  d <- filter(jp_a, language == lang)
  ok <- vapply(seq_len(nrow(d)), function(i)
    isTRUE(lower95(round(d[[col]][i] * d$reps[i]), d$reps[i]) >= target), logical(1))
  if (!any(ok)) NA_integer_ else as.integer(min(d$n_participants[ok]))
}

# The interaction alone (H1b), which is the confirmatory prediction. Rates come with
# an exact interval, because every one of them is a proportion over a few dozen to a
# few hundred replicates and none is known to the precision a bare percentage implies.
h1bv  <- function(lang, N, col = "p_h1b") jval(lang, N, col)
h1bci <- function(lang, N, col = "p_h1b") {
  r <- jrow(lang, N); k <- round(r[[col]] * r$reps)
  b <- stats::binom.test(k, r$reps)$conf.int
  sprintf("[%.2f, %.2f]", b[1], b[2])
}

# Pooled arm, combined across sample sizes at or below the cap. The arm is flat in N,
# which is what a verb-limited design predicts and what the cells show, so combining
# them is the best available estimate; cell by cell the pooled arm holds only 16 to 30
# replicates and no single cell settles anything.
pooled_combined <- function(col, max_n = 100) {
  d <- pooledp[pooledp$n_participants <= max_n, ]
  k <- sum(round(d[[col]] * d$reps)); n <- sum(d$reps)
  b <- stats::binom.test(k, n)$conf.int
  list(p = k / n, k = k, n = n, lo = b[1], hi = b[2],
       txt = sprintf("%s [%.2f, %.2f]", pct0(k / n), b[1], b[2]))
}

# A directional Bayes factor of k is met exactly when the posterior probability of the
# predicted sign reaches k/(k+1). Stating the equivalent one-sided tail probability
# keeps the reader from reading these thresholds as the point-null Bayes factors that
# the conventional verbal labels describe.
bf_sign  <- function(k) k / (k + 1)
bf_z     <- function(k) qnorm(bf_sign(k))
bf_onep  <- function(k) 1 - pnorm(bf_z(k))
bf_for_p <- function(p) { z <- qnorm(1 - p); s <- pnorm(z); s / (1 - s) }

sg_max        <- function(lang) max(filter(jp, mode == "safeguard", reps >= MIN_REPS_V2,
                                           language == lang)$p_joint)

# Pooled expectations, derived from the per-language pilot parameters: the
# verb-information floor with all three languages' verbs pooled, and the
# ceiling implied by the between-language spread of the interaction (the
# quantity that shrinks with neither verbs nor participants).
pooled_floor       <- mean(pp$tau_verb_intercept) / sqrt(sum(pp$sxx))
pooled_floor_per_sd <- pooled_floor * mean(pp$affectedness_sd)
pooled_het_ceiling <- pnorm(abs(mean(pp$interaction_mean)) /
                             (sd(pp$interaction_mean) / sqrt(nrow(pp))) - qnorm(10 / 11))

# Literature cross-check reference cells: the full verb set and the reduced
# 40-verb set at the smallest sample, and the secondary (H2) detection range
# at the full verb set.
lit50    <- filter(foc_full, n_verbs == 72, n_participants == 50)
lit40_50 <- filter(foc_full, language == "English", n_verbs == 40,
                   n_participants == 50)$binding
h2_72    <- filter(exc, hypothesis == "H2a_pseudo_positive", n_verbs == 72,
                   reps >= MIN_REPS)$p_bf_primary

# Pooled cross-linguistic results, aggregated by scripts/aggregate_pooled.R.
# The pooled arm may still be pre-first-cell, so the report degrades to the
# forward-looking wording when the file is absent or has no usable cells.
pooled_csv <- "../outputs/design_summary_pilot/pooled_power_pilot.csv"
pooledp <- if (file.exists(pooled_csv)) read_csv(pooled_csv, show_col_types = FALSE) else NULL
# Pooled joint power at the smallest per-language sample, by threshold, for the
# recommendation. Degrades to a neutral phrase before the first cell lands.
pool50 <- if (!is.null(pooledp) && nrow(pooledp) && 50 %in% pooledp$n_participants) {
  pooledp[pooledp$n_participants == 50, ]
} else NULL
pooled_pct <- function(col) if (!is.null(pool50)) pct0(pool50[[col]]) else "not yet estimated"

pooled_interim <- if (!is.null(pooledp) && nrow(pooledp) && 50 %in% pooledp$n_participants) {
  r <- pooledp[pooledp$n_participants == 50, ]
  sprintf(paste("The first pooled cells have returned, and they separate the two evidence",
                "thresholds sharply. At 50 participants per language, across the %d replicates",
                "available so far, the joint power is %s at the strong-evidence threshold but",
                "%s at a Bayes factor of 3, with median Bayes factors of %.0f and %.0f for the",
                "affectedness slope and the interaction. Pooling therefore makes moderate",
                "evidence for both predictions close to certain within the cap, while strong",
                "evidence for both together falls well short of the conventional target.",
                "Restricting the estimate to fits that pass every convergence check moves it",
                "by a few percentage points in either direction and does not change that",
                "conclusion. These cells carry a Monte-Carlo error of about %s, and the larger",
                "samples are still computing, so the figure will firm up but is unlikely to",
                "reverse."),
          r$reps, pct0(r$p_joint), pct0(r$p_joint_bf3), r$median_bf_h1a, r$median_bf_h1b,
          pct0(sqrt(r$p_joint * (1 - r$p_joint) / r$reps)))
} else "First results are expected within days."
```

`\begin{abstract}\normalsize\noindent`{=latex}
CLAPS asks whether the acceptability of a passive sentence depends on how strongly the
event affects the person or thing it happens to, and whether that pattern holds across
languages. This report asks how large a sample is needed to answer that convincingly. The
answer turns on an unusual feature of the design. Affectedness is a property of the verb,
not of the participant, so precision is governed by how many verbs are tested rather than by
how many people rate them. Past a point, recruiting more participants stops helping.

At the `r ppv("English", "n_verbs")` verbs used here, English's theoretical
infinite-participant ceiling is `r pct0(ppv("English", "ceiling_joint_bf10"))`, only just
above the conventional 80% target, and its finite-sample simulations remain below target.
Turkish's ceiling is itself below 80%.
Norwegian is the exception, reaching target at about `r nor_cross` participants. Earlier
simulations were more optimistic for a correctable reason. They assumed less variation
between verbs than the pilot data show, and omitted the variation in how individual verbs
behave across sentence types. Measured rather than assumed, both push the required samples
up sharply.

Two remedies fit within what the project can realistically collect, 80 participants per
language at the current verb count. Analysing the three languages together roughly triples
the verb-level information and lets every language contribute, Turkish included. Adopting a
moderate rather than a strong standard of evidence, as the project proposed at the outset,
brings English and Norwegian to target individually and makes the combined test
near-certain. Turkish meets neither standard alone and is better reported as an estimate
than as a test. Simulations continue, but cannot change this picture, since every
per-language sample at or below the agreed cap is simulated to full replication and the
cells still accumulating all lie above it.
`\end{abstract}`{=latex}

`\vspace{6pt}`{=latex}

This note reports preliminary sample-size guidance for CLAPS (A Crosslinguistic Approach
to Passive Semantics). The guidance comes from a Bayes-factor design analysis. A design analysis of this kind
simulates the planned model many times under plausible true effects,
then asks how often a given sample size would yield compelling evidence for the focal
predictions.

Two versions of the analysis are run. The primary one is data-grounded. It fits the planned model to each language's real pilot data and simulates new studies
from that fit, so its results reflect the effects actually observed in the CLAPS pilot.
The collaborators asked the design analysis to use that pilot as its basis. A single pilot's point estimate is noisy
and tends to overstate power, the concern that Albers and Lakens (2018) raise for
pilot-based power analyses. The data-grounded analysis avoids that point estimate in two
ways. First, on each simulated study it draws the focal effects from the pilot posterior, the
range of effect sizes the pilot data still leave plausible, so that the reported
probability of a decisive Bayes factor reflects the pilot's uncertainty about the size of
the effect. This yields an assurance, a power averaged over the
plausible sizes of the effect, rather than a power computed at one assumed size. Second, it also
evaluates a deliberately conservative lower bound on the effect, as a safeguard against an
optimistic pilot. A second version, for comparison, is based instead on the pooled effect sizes from the
five previously published studies. It provides an independent cross-check
against the broader evidence base.

The two versions agree on the central structural point. The power to detect the
affectedness effect is governed by the number of verbs in the materials rather than by the
number of participants, because affectedness is a property of the verb. They diverge on
whether a modest sample is enough. On the published-literature effects, a per-language
sample of at most 50 participants would suffice at the full verb set. On the pilot effects,
and with the verb-level variability measured from the pilot rather than assumed, the
picture is less favourable. Norwegian reaches the target at about `r nor_cross`
participants, but English and Turkish do not reach it at any simulated sample size. For those two languages,
the binding constraint is the number of verbs, and in Turkish also the strength of the
effect itself, so adding participants alone cannot bring the design to the usual power
target. Because the data-grounded analysis is the agreed primary basis, this is the finding
that should guide the design. The report quantifies these limits analytically, explains why
the two analyses disagree, and sets out design options that remain feasible for a volunteer
consortium. The recommended option is a pooled cross-linguistic analysis, and the section on it below
gives its rationale and the first results.

The data-grounded analysis is substantially complete. Results are in for
all three languages up to `r max_n("Norwegian")` participants, for English up to
`r max_n("English")` and for Turkish up to `r max_n("Turkish")`, and the largest samples
are still computing. Each simulated
condition rests on `r min(jp_a$reps)` to `r max(jp_a$reps)` studies, so the individual
figures carry a Monte-Carlo error of several percentage points, that is, the uncertainty
that comes from basing each figure on a limited number of simulated studies.
The 400-participant simulations now confirm the English plateau, the levelling-off of
power short of the target, so it no longer rests on extrapolation. The report should be read as a decision-relevant interim pass rather than a
final recommendation.

# Background and Motivation

The question CLAPS addresses is whether the acceptability of passive and related sentence
types tracks the semantic affectedness of the patient, and whether that relationship
differs by sentence type. Affectedness, the degree to which a patient or theme undergoes a
change of state or is causally impinged upon by the action, is a core dimension of
argument structure [@dowty1991; @beavers2011]. It has long been argued to be the semantic
heart of the passive, which carries the meaning that the surface subject is in a state or
circumstance characterised by the agent having acted upon it [@pinkerLebeauxFrost1987].
Across languages, adult acceptability-judgement and comprehension studies support the view
that passive acceptability is semantically structured rather than purely syntactic
[@ambridgeBidgoodPineRowlandFreudenthal2016; @aryawibawaAmbridge2018;
@darmasetiyawanAmbridge2022; @liuAmbridge2021; @bidgoodPineRowlandAmbridge2020;
@ambridgeArnonBekman2023].

The cross-linguistic evidence base to date spans English, Indonesian, Balinese, Hebrew and
Mandarin. A Bayesian meta-analytic synthesis finds a consistent pattern across them
[@ambridgeArnonBekman2023]. Affectedness raises acceptability for both passives and actives,
more strongly for passives, and does not raise it for pseudo-passives. The
affectedness constraint has not yet been tested in Turkish or Norwegian. CLAPS extends the
paradigm to these two typologically and morphologically distinct languages. No prior
language-specific effect-size estimates exist for them, so the CLAPS pilot supplies the
first such estimates. The data-grounded analysis uses these pilot estimates as its primary
basis, and the cross-linguistic literature provides an independent comparison, which turns
out to be the more optimistic of the two.

# The Acceptability Model and the Focal Predictions

Each acceptability judgement is a discrete rating from 1 to 7. These ratings are modelled
with a Bayesian cumulative-logit mixed-effects model, an ordinal model that respects the
discreteness, ordering and bounded range of rating data while leaving the underlying scale
free to vary across conditions [@burknerVuorre2019]. Treating the same ratings as
continuous, under a Gaussian likelihood, biases the regression coefficients and inflates
error rates where floor or ceiling effects are present [@liddellKruschke2018]. Recent
empirical work shows that modelling them cumulatively recovers psycholinguistic effects
more accurately [@verissimo2021; @taylorRousseletScheepersSereno2023].

The expected pattern is already visible in the pilot data, which come from the sample whose
composition by language is given in @tbl-pilot in Appendix A. @fig-pilot shows the raw
acceptability ratings against agent semantic affectedness, separately by sentence type,
referent gender and language. The affectedness slope is positive and steepest for the
passive in English and Norwegian, and shallower for the active and the pseudo-passive. The
trends are very similar for male and female referents within each language. This is
reassuring, because the focal effects should not be driven by a referent-gender imbalance.

```{r fig-pilot}
#| fig-cap: "Raw Pilot Acceptability by Agent Semantic Affectedness, Sentence Type, Referent Gender and Language"
#| fig-width: 6.5
#| fig-pos: "H"
#| fig-align: center
#| out-width: "6.1in"

knitr::include_graphics(pilot_figure)
```

`\vspace{-8pt}`{=latex}

*Note.* Points are individual pilot ratings, jittered. Lines are per-sentence-type linear
trends with 95% confidence bands. The pilot data are used for calibration only. Norwegian
has no pseudo-passive.

`\vspace{12pt}`{=latex}

The model predicts acceptability from sentence type, from verb affectedness, and from
their interaction. The passive is the reference level, and the active and the
pseudo-passive are the comparisons. The focal predictions concern the affectedness slope in the
passive and how it changes across sentence types. The primary prediction (H1b) is
one-tailed and holds that affectedness raises passive acceptability more than it raises
active acceptability, so the active-by-affectedness interaction is negative. A supporting
prediction (H1a) holds that affectedness raises acceptability in the passive itself, so
the passive affectedness slope is positive. A sample size is judged adequate only when
both of these predictions reach the evidence threshold. A secondary, two-tailed prediction
(H2) concerns the pseudo-passive-by-affectedness interaction. It does not gate the
recommendation and does not apply to Norwegian, which has no pseudo-passive.

# Design-Analysis Method

The design analysis estimates statistical power by simulation. Each simulated dataset is
drawn from the same cumulative-logit model that will be used for analysis, under assumed
values for the effects of interest. Each dataset is then analysed in the same way. The
proportion of simulated studies that yield decisive evidence estimates the power at that
sample size. The whole procedure stays within the Bayesian framework. This matters because
the confirmatory test is a Bayes factor rather than a significance test, and the power to
return a decisive Bayes factor is not the same quantity as the power to return a significant
*p*-value.

A central feature of the design is that affectedness is a property of the verb rather than
of the individual trial. Each verb carries a single affectedness value, held constant
across participants and sentence types, exactly as in the real materials. The number of verbs, rather than the number of trials,
therefore determines the precision with which the affectedness effect can be estimated. The
verb count is treated as a design dimension alongside the participant count.

The two analyses differ in where they take the assumed true effects and the affectedness
values. The data-grounded analysis, which is primary, fits the planned model to each
language's pilot data. It then takes the population effects, the random-effect covariance
and the ordinal thresholds from that fit, and it reads each verb's actual affectedness
score from the pilot data. The literature-anchored cross-check instead sets the
population effects to values from the cross-linguistic meta-analysis. On the standardised
affectedness predictor, the passive affectedness slope is set near the pooled value of
about 0.23 logits per standard deviation (a logit is a unit of the model's underlying scale, not a raw point on the 1 to 7 rating), with a negative active-by-affectedness
interaction of about $-0.15$ and a small positive pseudo-passive-by-affectedness
interaction. In the cross-check, affectedness is spread across a plausible range rather than
read from the pilot, and the by-participant and by-verb random-effect standard deviations
are set to realistic magnitudes. The referent-gender covariate is held out of both
preliminary runs and is examined separately.

The analysis model uses the maximal random-effects structure consistent with the design,
meaning the most complex structure the design supports, so any effect that can vary across
participants or verbs is allowed to do so. It has correlated by-participant slopes for
sentence type, affectedness and their interaction, together with by-verb intercepts and
sentence-type slopes [@barrLevyScheepersTily2013].
Because affectedness does not vary within a verb, a by-verb affectedness slope cannot be
estimated and is omitted, so affectedness enters as a population-level effect, one estimated
for the sample as a whole rather than allowed to vary by verb. The priors
are weakly to moderately informative and centred at zero on the focal slopes, so that the
directional tests are not pre-loaded. A normal prior of standard deviation 0.5 is placed on
the affectedness slope and on the active interaction, and a slightly wider normal prior of
0.6 on the pseudo-passive interaction. The sentence-type main effects take a wider normal
prior. Student-*t* priors are placed on the ordinal thresholds and on the by-participant and
by-verb standard deviations, and an LKJ prior, the standard prior for a correlation matrix,
on the correlations among those random effects
[@gelman2006; @lewandowskiKurowickaJoe2009].

Evidence for each prediction is summarised by a directional Savage-Dickey Bayes factor, a
number that weighs how strongly the data favour the predicted direction of the effect over
there being no effect, obtained by sampling the prior alongside the posterior
[@wagenmakersLodewyckxKuriyalGrasman2010]. A simulated study counts as a detection when its
Bayes factor reaches 10, the preregistered threshold for strong evidence. Under the
zero-centred prior, this is not a stringent bar. The Bayes factor used here is directional,
meaning the posterior odds that the coefficient carries the predicted sign divided by the
same odds under the prior, so under a symmetric prior a value of 10 is odds of 10 to 1 on the
direction. That is a posterior probability of 0.91, or about 1.34 standard errors from zero,
whereas a one-sided *p* < .05 requires 1.64, so the Bayesian criterion is if anything the
less demanding of the two. It is not the point-null Bayes factor of a default two-sided test,
which is a different quantity on a different scale. Power is the
proportion of simulated studies that reach it. The recommended sample size is the smallest
at which both focal predictions reach a power of 80%.

Each replicate is fitted with four
Markov chains of 3,000 iterations after 1,000 warm-up iterations, at the same target
acceptance rate of 0.99 and maximum tree depth of 12 that the confirmatory analysis will
use. The proposal specifies a longer run for that single confirmatory fit, at sixteen chains
of 5,000 iterations after 2,000 warm-up iterations. The shorter chains add Monte-Carlo noise to each simulated Bayes
factor rather than shifting it systematically, so the power figures below are, if anything,
mildly conservative.

The literature-anchored grid crosses the three languages with participant counts from 50 to
120. The output read here covers `r lit_language_text` from `r lit_n_min` to `r lit_n_max`
participants at verb counts of 40 and 72. Each completed cell holds 50 replicates, which
gives a Monte-Carlo standard error of about five to six percentage
points near a power of 80%, so estimates close to the threshold are not perfectly ordered
across adjacent sample sizes. The central code for the data-generating process, the model
and the Bayes factor is reproduced in Appendix B. The full operating characteristics of this
run, cell by cell, are reported in @tbl-opchar in Appendix C, and Appendix D maps the whole
repository, including which file implements each of the two simulation engines described
above and how the pooled analysis reuses one of them.

# Results

## Data-grounded results

The data-grounded analysis is the primary basis for the recommendation, and it is now
substantially complete. In its assurance form it draws the focal effects afresh from each
language's pilot posterior on every simulated study, so the reported power carries the
pilot's own uncertainty about the size of those effects. @tbl-pilot-joint reports the joint
power, the proportion of simulated studies in which both focal predictions clear the
Bayes-factor threshold together, at the pilot's real verb count.

```{r tbl-pilot-joint}
#| tbl-cap: "Preliminary Data-Grounded Joint Power by Participant Count and Language, Assurance Variant"

pj <- tidyr::crossing(n_participants = sort(unique(jp_a$n_participants)),
                      language = c("English", "Turkish", "Norwegian")) |>
  left_join(select(jp_a, language, n_participants, p_joint),
            by = c("language", "n_participants")) |>
  mutate(cell = ifelse(is.na(p_joint), "-", pct1(p_joint))) |>
  select(-p_joint) |>
  pivot_wider(names_from = language, values_from = cell)
kable(pj[, c("n_participants", "English", "Turkish", "Norwegian")],
      format = "latex", booktabs = TRUE, longtable = TRUE,
      align = "rrrr",
      col.names = c("Participants", "English", "Turkish", "Norwegian")) |>
  kableExtra::kable_styling(
    latex_options = "repeat_header",
    repeat_header_method = "replace",
    repeat_header_text = "\\textbf{Table 1 (continued)}"
  )
```

`\vspace{-8pt}`{=latex}

*Note.* Joint power is the proportion of studies in which both focal predictions clear a
Bayes factor of 10 together, at 72 verbs and 66 for Norwegian. Cells below `r MIN_REPS_V2`
replicates are still accumulating and shown blank. The rest hold `r min(jp_a$reps)` to
`r max(jp_a$reps)` studies each. Percentages are shown to one decimal place to preserve the
exact Monte-Carlo summaries. Norwegian at 130 participants sits a little above the
approximate infinite-participant ceiling in @tbl-ceilings, at `r jratio("Norwegian", 130)` =
`r pct1(jval("Norwegian", 130, "p_joint"))` against a Monte-Carlo standard error of
`r pct1(jmcse("Norwegian", 130))`, which is sampling variation rather than a disagreement
between the two.

`\vspace{12pt}`{=latex}

The pattern differs sharply by language. Norwegian reaches the 80% target at about
`r nor_cross` participants and is comfortably powered beyond it. English and Turkish do not
reach it at any simulated sample size. For English, the analytic ceiling is
`r pct0(ppv("English", "ceiling_joint_bf10"))`, but that is the limiting value with
unlimited participants, not the power of a finite study; its simulated joint power remains
below 80%. It is `r pct1(jval("English", 300, "p_joint"))` by 300 participants and
`r pct1(jval("English", 400, "p_joint"))` at 400. The reason is
structural. The joint requirement is held back by the affectedness main effect (H1a), whose
detection rate levels off at roughly three in four, while the interaction (H1b) climbs
comfortably past it. Because affectedness is a property of the verb, its main effect is
estimated from the 72 verbs, so adding participants does not sharpen it. The precision of
that effect, and with it the power to detect it, is set by the number of verbs. For Turkish,
both focal effects are weak in the pilot. The joint power never exceeds
`r pct1(tur_max_joint)` at any simulated sample size, and the present design does not bring
it to target.

A conservative safeguard variant, which fixes each focal effect at a cautious lower bound of
its pilot posterior, is lower again. Its highest joint power over the samples run is
`r pct0(sg_max("Norwegian"))` for Norwegian, `r pct0(sg_max("English"))` for English and
`r pct0(sg_max("Turkish"))` for Turkish. This is the design's power under a deliberately
pessimistic reading of the pilot.

The pilot posterior itself puts the probability that both focal effects have their predicted sign at `r pct1(ppv("English", "sign_prob_joint"))` for English,
`r pct1(ppv("Norwegian", "sign_prob_joint"))` for Norwegian and
`r pct1(ppv("Turkish", "sign_prob_joint"))` for Turkish, so the pilot points the right way
in all three languages. The shortfall
for English and Turkish is therefore a matter of precision rather than direction. Because
affectedness varies only
between verbs, the standard error of its effect cannot fall below the by-verb variability
divided by the spread of the affectedness scores, however many participants respond. This
is the stimulus-sampling limit long recognised in psycholinguistics [@clark1973] and
quantified for designs with crossed random effects by Westfall et al. [-@westfallKennyJudd2014;
see also @juddWestfallKenny2017; @brysbaertStevens2018]. The limit can be computed directly
from the fitted pilot models, with no simulation involved. @tbl-ceilings reports the
resulting upper bounds, the detection rates that would remain even with unlimited
participants at the current verb counts.

```{r tbl-ceilings}
#| tbl-cap: "Analytic Upper Bounds on Detection with Unlimited Participants, at the Current Verb Counts"

fmt_verbs <- function(v) ifelse(is.na(v), "> 1,200", format(v, big.mark = ","))
ce <- pp |> transmute(
  Language        = as.character(language),
  `H1a ceiling`   = pct0(ceiling_h1a_bf10),
  `H1b ceiling`   = pct0(ceiling_h1b_bf10),
  `Joint ceiling` = pct0(ceiling_joint_bf10),
  `Verbs for 80%` = fmt_verbs(verbs_for_80),
  `Verbs for 90%` = fmt_verbs(verbs_for_90)
)
kable(ce, align = "lrrrrr")
```

`\vspace{-8pt}`{=latex}

*Note.* Ceilings are the probability of a Bayes factor of at least 10 with unlimited
participants, from the `r format(ppv("English", "ndraws"), big.mark = ",")` pilot posterior
draws and the fitted by-verb covariance at 72 verbs and 66 for Norwegian. The last columns
give the verbs needed for an 80% and 90% joint ceiling. A ceiling just above target, as for
English, still leaves it out of reach, since any finite sample sits below its ceiling.

`\vspace{12pt}`{=latex}

These bounds agree closely with the simulations. English's simulated joint power climbs to
`r pct1(jval("English", 300, "p_joint"))` by 300 participants against a ceiling of
`r pct0(ppv("English", "ceiling_joint_bf10"))`. Turkish never exceeds
`r pct1(tur_max_joint)` against a ceiling of `r pct0(ppv("Turkish", "ceiling_joint_bf10"))`.
Norwegian's simulated `r pct1(jval("Norwegian", 100, "p_joint"))` at 100 participants
matches its ceiling of `r pct0(ppv("Norwegian", "ceiling_joint_bf10"))` within Monte-Carlo
error. The agreement between two entirely different routes to the same
quantities, stochastic simulation and direct calculation, is a strong check that the
pipeline is sound. A further check makes the point starkly for English. The pilot's own posterior standard deviation for the affectedness slope, which measures how precisely the pilot pins that slope down, is `r num2(ppv("English", "slope_post_sd"))`, already close to the verb-count floor of `r num2(ppv("English", "se_floor_h1a"))`. A replication with the same 72 verbs would therefore have little room to
estimate that slope more precisely than the pilot already has, whatever its sample size.
The residual gap reflects the floor formula's own approximation, since it ignores some of
the covariance in the by-verb random effects, rather than a real target that more
participants could close.

The ceilings also show what would and would not help. Relaxing the evidence criterion from
a Bayes factor of 10 to 6 or 3 lifts the English joint ceiling to
`r pct0(ppv("English", "ceiling_joint_bf6"))` and `r pct0(ppv("English", "ceiling_joint_bf3"))`,
but lifts Turkish only to `r pct0(ppv("Turkish", "ceiling_joint_bf6"))` and
`r pct0(ppv("Turkish", "ceiling_joint_bf3"))`, so a weaker threshold cannot rescue the
Turkish test. Adding verbs raises every ceiling, because the precision floor shrinks with
the square root of the verb count. English would reach a joint ceiling of 90% at about
`r ppv("English", "verbs_for_90")` verbs, and Norwegian at about
`r ppv("Norwegian", "verbs_for_90")`. Turkish would need about
`r ppv("Turkish", "verbs_for_80")` verbs merely to reach a ceiling of 80%, which is not
realistic within a single language, so for Turkish the remedy lies elsewhere, as taken up
under Paths Forward below.

These results are preliminary in one remaining respect, since each condition rests on
`r min(jp_a$reps)` to `r max(jp_a$reps)` simulated studies and so carries a Monte-Carlo
error of a few percentage points at the better-populated sample sizes and more at the
sparser ones. The
English plateau itself, though, is now confirmed rather than extrapolated. At 400
participants, with `r jval("English", 400, "reps")` replicates in, the joint power sits at
`r pct1(jval("English", 400, "p_joint"))` (`r jratio("English", 400)`), compared with
`r pct1(jval("English", 300, "p_joint"))` (`r jratio("English", 300)`) at 300 participants.
These sparse estimates are not literally flat, but their decline is compatible with
Monte-Carlo variation and, critically, there is no sustained rise towards the target.
Overturning the plateau would
have required a sustained rise at 400 and 500, and the 400-participant data already rule
that out. The largest cells still computing will fill in the remaining sample sizes, but by that
test they can no longer overturn the plateau.

## Literature-anchored cross-check

The grid is read as a surface over participant count and verb count, summarised by the
binding focal power, the smaller of the two focal detection probabilities, since the
recommendation requires both to clear 80%. This cross-check is the more optimistic of the
two analyses, because it fixes the effects at the pooled literature values and carries no
uncertainty about them. @fig-surface
shows this surface for the two languages whose cells are complete. Cells outlined in gold
meet the target, and blank cells are those whose replicates are still accumulating. The
pattern is clear. Power increases steeply with the number of verbs and only weakly with the
number of participants.

```{r fig-surface}
#| fig-cap: "Binding Focal Power by Participant Count and Verb Count, for Each Language"
#| fig-width: 7.0
#| fig-height: 2.5
#| fig-pos: "H"

# Languages and sample sizes are taken from the data rather than listed here,
# so the panel follows whatever the aggregation covers.
grid_df <- tidyr::crossing(
  language = sort(unique(foc_full$language)),
  n_participants = sort(unique(foc_full$n_participants)),
  n_verbs = c(40, 72)
) |>
  left_join(select(foc_full, language, n_participants, n_verbs, binding),
            by = c("language", "n_participants", "n_verbs")) |>
  mutate(
    nv = factor(n_verbs, levels = c(40, 72)),
    np = factor(n_participants, levels = c(50, 60, 70, 80, 90, 100, 120)),
    lab = ifelse(is.na(binding), "", sprintf("%.0f", 100 * binding)),
    txtcol = ifelse(!is.na(binding) & binding >= 0.55, "white", "grey15")
  )

ggplot(grid_df, aes(np, nv, fill = binding)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_tile(data = filter(grid_df, !is.na(binding), binding >= 0.80),
            color = "#E6A100", linewidth = 1.1, fill = NA) +
  geom_text(aes(label = lab, color = txtcol), size = 3.1, fontface = "bold") +
  scale_color_identity() +
  scale_fill_gradientn(
    colours = c("#f7fbff", "#c6dbef", "#6baed6", "#2171b5", "#08306b"),
    limits = c(0, 1), na.value = "grey92",
    name = "Binding focal\npower",
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_wrap(~ language) +
  labs(x = expression(paste("Participants per language (", italic(N), ")")),
       y = "Number of verbs") +
  guides(fill = guide_colourbar(barheight = unit(2.8, "cm"),
                                barwidth = unit(0.4, "cm"))) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(t = 8, r = 4, b = 4, l = 4),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

`\vspace{-8pt}`{=latex}

*Note.* Each cell gives the binding focal power as a percentage, with darker shading for
higher power. Cells outlined in gold meet the 80% target. Blank cells are still accumulating.

`\vspace{12pt}`{=latex}

The two axes are not equally important. At 40 verbs in English, the binding power is
`r pct0(lit40_50)` at 50 participants and reaches the 80% target only between about 60 and
90 participants, within Monte-Carlo noise. At 72 verbs, it is already
`r pct0(min(lit50$binding))` to `r pct0(max(lit50$binding))` at 50 participants. Increasing the number of verbs from 40 to 72 raises power more than increasing
the number of participants does, which follows from affectedness being a verb-level
property. @fig-curves separates the two focal predictions and shows how each approaches the
target as the sample grows, at each verb count.

```{r fig-curves}
#| fig-cap: "Detection Probability for the Focal Predictions, with Wilson 95% Intervals, by Verb Count and Language"
#| fig-width: 6.8
#| fig-height: 2.8
#| fig-pos: "H"

z <- 1.96
curve_df <- foc_full |>
  select(language, n_participants, n_verbs, reps, H1a, H1b) |>
  tidyr::pivot_longer(c(H1a, H1b), names_to = "Prediction", values_to = "p") |>
  mutate(
    den = 1 + z^2 / reps,
    ctr = (p + z^2 / (2 * reps)) / den,
    hw  = z * sqrt(p * (1 - p) / reps + z^2 / (4 * reps^2)) / den,
    lo  = pmax(0, ctr - hw), hi = pmin(1, ctr + hw),
    nv  = factor(paste(n_verbs, "verbs"), levels = paste(c(40, 72), "verbs"))
  )

pd <- position_dodge(width = 4)
ggplot(curve_df, aes(n_participants, p, colour = Prediction, shape = Prediction)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey55") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 3, linewidth = 0.4, position = pd) +
  geom_line(linewidth = 0.7, position = pd) +
  geom_point(size = 2, position = pd, fill = "white", stroke = 0.7) +
  facet_grid(nv ~ language) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = c(H1a = "#0072B2", H1b = "#D55E00")) +
  scale_shape_manual(values = c(H1a = 21, H1b = 24)) +
  labs(x = expression(paste("Participants per language (", italic(N), ")")),
       y = "Detection probability", colour = "Prediction", shape = "Prediction") +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

`\vspace{-8pt}`{=latex}

*Note.* Points are detection probabilities over the simulated studies in each complete cell,
with Wilson 95% intervals reflecting Monte-Carlo uncertainty. The dashed line marks the 80%
target. Points are dodged so the two predictions do not overlap.

`\vspace{12pt}`{=latex}

In the literature-anchored output, which covers `r lit_language_text`, the binding
focal power at the full 72-verb set is already between
`r pct0(min(lit50$binding))` and `r pct0(max(lit50$binding))` at 50 participants, the
smallest sample simulated, in every language included in this cross-check. On the
literature effects, the adequate per-language sample is therefore at most 50 participants at
the full verb set. This is the optimistic reading. The data-grounded results above, which
condition on the weaker pilot effects, do not reach the target for English or Turkish at any
sample size. Because 50 is the smallest sample in the main grid, the analysis places
the adequate sample at or below 50 without determining it more precisely. Pinning it down further, below 50 participants, is not pursued here, since the
recommendation rests on the data-grounded analysis rather than on this cross-check.

## Why the Two Analyses Disagree

The gap between the two analyses calls for an explanation, because the literature-anchored
grid reports `r pct0(min(lit50$binding))` to `r pct0(max(lit50$binding))` at 50
participants on the same 72 verbs at which the data-grounded joint power never reaches 80%.
Expressed per standard deviation of affectedness, the pilot's passive slope is
`r num2(ppv("English", "slope_per_sd"))` logits for English,
`r num2(ppv("Norwegian", "slope_per_sd"))` for Norwegian and
`r num2(ppv("Turkish", "slope_per_sd"))` for Turkish, against a pooled literature value of
about 0.23. The Turkish active interaction, at
`r mnum2(ppv("Turkish", "interaction_per_sd"))`, matches the literature's $-0.15$ almost
exactly, and the English interaction, at `r mnum2(ppv("English", "interaction_per_sd"))`, is
almost three times the literature value, so the CLAPS pilot effects are not unusually
small.

The disagreement comes instead from the noise side of the simulation, in two specific
assumptions. First, the literature-anchored simulator assumed a by-verb intercept standard
deviation of 0.5, a measure of how much verbs differ in their baseline acceptability, where the fitted pilot models put it at
`r num2(ppv("English", "tau_verb_intercept"))` for English,
`r num2(ppv("Turkish", "tau_verb_intercept"))` for Turkish and
`r num2(ppv("Norwegian", "tau_verb_intercept"))` for Norwegian. The precision floor on the
affectedness slope scales directly with this quantity, so assuming 0.5 where the fitted
value is `r num2(ppv("English", "tau_verb_intercept"))` overstates the attainable evidence
considerably. Second, the literature-anchored simulator included no by-verb sentence-type
slopes at all, allowing no verb-to-verb variation in how sentence type shifts the ratings, where the fitted pilots put their standard deviation at
`r num2(ppv("English", "tau_verb_active"))` for English,
`r num2(ppv("Turkish", "tau_verb_active"))` for Turkish and
`r num2(ppv("Norwegian", "tau_verb_active"))` for Norwegian. That omission left the
interaction test with no verb-level noise floor in the simulation, which is why the earlier
grid found the interaction so easy to detect. Both assumptions were reasonable defaults
before the pilot models existed. The pilot has now measured the quantities they stood in
for, and the data-grounded analysis uses the measured values. The earlier optimism, in
short, came less from the assumed effect sizes than from understating how much verbs vary
[@clark1973; @westfallKennyJudd2014].

## Effect-Size Qualification

One assumption deserves particular care, because it bears directly on the absolute sample
size. The numbers shown here condition on assumed effects set to the cross-linguistic
pooled posterior means from a Bayesian meta-analytic synthesis of the passive-affectedness
literature, which is a central estimate rather than an optimistic one
[@ambridgeArnonBekman2023]. On the simulator's internal predictor scale, these appear as
larger nominal values, but expressed per standard deviation of affectedness, they are about
0.23 and $-0.15$ logits, matching the pooled literature values. The larger nominal numbers
therefore reflect the scaling rather than a larger effect. Two cautions nonetheless apply. The first is the one the previous section sets out. The
fitted pilot models put the verb-level variability well above the values this grid assumed,
and that difference, more than any difference in effect size, is what made the grid
optimistic. The pilot effects themselves are broadly in line with the pooled literature
values once both are expressed per standard deviation of affectedness, though the Turkish
terms sit at the weak end, with the interaction at about
`r mnum2(ppv("Turkish", "interaction_per_sd"))` in those units. The second is that the literature is itself
subject to publication bias and to the winner's curse, the tendency of published estimates,
especially from small and low-powered studies, to overstate true magnitudes
[@gelmanCarlin2014; @ioannidis2008; @button2013]. Design analyses anchored to such estimates
can therefore be optimistic about power [@vasishth2018; @albersLakens2018].

These cautions are the reason the data-grounded analysis, rather than this cross-check, is
the primary basis for the recommendation. It draws the focal effects from each language's
own pilot posterior, and it also evaluates a conservative lower bound. It therefore
conditions on the effects observed in CLAPS and carries their uncertainty through to the
reported power. Its results are reported above. They agree with this cross-check for
Norwegian, which is powered at about 100 participants, and they diverge from it for English
and Turkish, where the weaker pilot effects hold the joint power below target at every
sample size run.

A separate point concerns the scale on which the prior sits, rather than the size of the
assumed effect. The literature-anchored grid behind the figures expresses affectedness on
the simulator's internal predictor, whose standard deviation is about 0.29. The real confirmatory analysis and the data-grounded analysis both place it on the analysis
pipeline's Gelman scaling, which divides a predictor by twice its standard deviation and so
leaves it with a standard deviation of 0.5, and the same weakly informative prior is applied
throughout. A Bayes factor reflects the width of the prior
relative to the coefficient it constrains. Holding the prior fixed while the predictor scale
differs therefore leaves the literature-anchored grid mildly optimistic about the
focal-slope Bayes factor, by a factor of about 1.7, relative to the analysis that will
actually be run. The qualitative conclusion is unaffected, because the verb count dominates
the result, the focal Bayes factors at the full verb set lie far above the threshold, and
the data-grounded analysis already uses the correct scale. A scale-corrected version of the
cross-check grid is being computed to confirm the per-language sample sizes directly. Until
it completes, the figures here should be read as a mild upper bound on the Bayes-factor
evidence.

For an effect that lives at the level of the participant, a weaker true effect simply needs
a larger sample, and the rough guide is that a true effect around two-thirds of the assumed
magnitude roughly doubles the sample needed. The affectedness main effect lives at the level
of the verb, however, where its precision is set by the number of verbs, so a weaker true
effect cannot be recovered by recruiting more
participants. This is exactly what the data-grounded results show. At the full verb count
and the pilot's own effects, the affectedness power for English settles below the target and
does not climb with sample size, and for Turkish neither focal effect is strong enough to
bring the joint power to target. The effective levers are therefore the number of verbs, and
for Turkish the strength of the interaction, not the participant count. Where feasible, the
design should still be planned against a deliberately conservative, or safeguard, effect
size rather than a point estimate [@perugini2014], and the data-grounded analysis builds
this in directly through its assurance and safeguard variants. Discounting the
literature-anchored grid towards weaker effects makes the same point: re-evaluated at 75%
and 60% of the assumed effect magnitude, with the 60% level approximating the weaker effects
seen in the pilot, it needs larger samples. The data-grounded analysis conditions on those
pilot effects directly, so the primary results above already reflect that discounting.

## Convergence and Computation

Convergence is sound across the grid, with no divergent transitions anywhere, which are the
warning sign of an unreliable fit. At the smaller verb count, essentially every fit meets the
strict criteria, namely an R-hat below 1.01, which checks that the separate chains have
settled on the same answer, together with bulk and tail effective sample sizes of at least
400, which measure how much independent information the draws carry. At 72 verbs, the maximal
model occasionally falls just short of these criteria, in about one fit in ten at the smaller
samples and up to one in five at the largest, where *N* is 80 to 90. It falls short on the
R-hat or effective-sample-size thresholds rather than on divergences, so it reflects how
efficiently the sampler is exploring rather than a fault in the model's specification.
Restricting the power calculation to the fully converged fits leaves the estimates
essentially unchanged. For English at 72 verbs and 50 participants, the binding focal power
is `r pct(cc$all)` over all 50 fits and `r pct(cc$conv)` over the `r cc$n_conv` converged
fits, so the recommendation does not depend on the difference.

## Secondary Prediction

The pseudo-passive-by-affectedness interaction reaches a detection probability of about
`r pct0(min(h2_72))` to `r pct0(max(h2_72))` at 72 verbs. This remains below the focal
predictions and reflects its smaller
assumed magnitude, and the opposite-direction test is near zero throughout, as expected.
This prediction is secondary and two-tailed, and it does not constrain the sample-size
recommendation. Full operating characteristics for all four predictions, including the
convergence rates, are reported in Appendix C.

# Paths Forward

The analytic bounds turn the shortfall into a set of concrete design choices. Four options
follow, in roughly increasing order of departure from the current plan. Two of them have
since been closed by the project, and are kept here because the reasoning that ruled them
out is part of the record.

The constraints are now these. The design is capped at 80 participants per language, with
100 the absolute maximum any team would be asked for. The verb pool cannot grow beyond 72:
the limit is the supply of suitable verbs in each language, not the length of a session, so
the third option below is unavailable rather than merely expensive. Collection must be at a
single sample size fixed and announced in advance, because the teams have to take a stated
number to their ethics committees, which rules out the fourth option. Many of the teams have
not run a study of this type before, and a design they cannot explain is not a design they
can run.

The first option changes nothing about the materials and revisits the adequacy rule. The
design's primary prediction is H1b, the active-by-affectedness interaction, with H1a as a
supporting prediction. Gating each language on the primary prediction alone, with the passive
slope reported as estimation, brings English to 80% on the interaction at
`r h1b_cross("English")` participants and Norwegian at `r h1b_cross("Norwegian")`, at the
current verb counts and at the strong threshold. Those crossing points are the smallest
samples at which the *lower bound* of the simulated rate reaches 80%, not the first cell whose
point estimate happens to; with a few dozen replicates per cell the two can differ by a whole
sample size. The joint gate on both predictions is what no feasible English sample can
satisfy. This option does not help Turkish, whose interaction ceiling is
`r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` even on its own.

The second option makes the cross-linguistic claim the primary confirmatory test. The
theoretical claim CLAPS addresses is explicitly cross-linguistic, that affectedness
structures passive acceptability across languages, and the meta-analytic synthesis of the
earlier studies frames the evidence in exactly those terms [@ambridgeArnonBekman2023]. A
hierarchical model fitted to all three languages jointly tests that claim directly. It also
multiplies the verb-level information, roughly 210 verb-by-language units in place of 72,
which divides the precision floor by about the square root of three. On the pilot effects,
every language's focal terms point the predicted way, so the pooled test starts from a
favourable position. Turkish then contributes fully to the confirmatory result, and its
language-specific effects are reported as estimates with credible intervals, which reflects
honestly what a 72-verb single-language design can support. The pooled design analysis runs on the same machinery used here, and it is now built. Its rationale and
the first results are set out below.

The third option expands the verb pool without lengthening any participant's session. Verbs,
not participants, are the scarce resource, and a planned-missingness design assigns each
participant a random subset of a larger pool [@grahamTaylorOlchowskiCumsille2006;
@littleRhemtulla2013]. Each volunteer still rates about 72 items, so the 45-minute session
is unchanged, while the pool grows to, say, 120 to 150 verbs. At about
`r ppv("English", "verbs_for_90")` verbs, the English joint ceiling reaches 90%, and the
pooled cross-linguistic test gains in proportion. The cost falls on materials, namely
affectedness norms and translations for the new verbs, rather than on participant time.
This option is closed. It would have required the 72-verb limit to be a constraint on the
session, which planned missingness respects, rather than on the pool. It is the pool: the
teams are already struggling to find 72 suitable verbs in most languages, so there is no
larger set to sample from. That matters beyond this one option, because the verb count is
what every ceiling in this report is a ceiling on. With the pool fixed, those ceilings are
now permanent features of the design rather than quantities a bigger materials budget could
move, and Turkish, which would need about `r ppv("Turkish", "verbs_for_80")` verbs to reach
80% on its own, is out of reach for good.

The fourth option spends volunteer capacity more efficiently rather than demanding more of
it. A sequential design analyses the data in waves and stops as soon as the Bayes factor reaches
a preset bound or the sample reaches an agreed maximum. This typically needs markedly fewer
participants than a fixed-sample design with the same operating characteristics, meaning the
same chances of reaching a decisive result
[@schonbrodtWagenmakersZehetleitnerPerugini2017; @stefanGronauSchonbrodtWagenmakers2019].
Optional stopping of this kind poses no problem for Bayes-factor inference [@rouder2014].

This option is also closed, and for a reason worth recording, since the statistical objection
to it is the one that does not apply. Nothing in the Bayes-factor framework penalises
optional stopping. The obstacle is operational: the collecting teams have to put a specific
number of participants in front of an ethics committee, and a protocol whose sample size
depends on interim results is one that many of them could not explain or defend, however
sound it is. A design that cannot be administered by the people running it is not available,
whatever its operating characteristics. The sample size below is therefore a single fixed
number, announced in advance.

The recommendation is the first two options taken together, and nothing else: the pooled
cross-linguistic analysis is the confirmatory test, it is gated on the interaction alone, its
threshold is a Bayes factor of 6, and every team collects 80 participants.

Gating on the interaction alone is what makes the design work, and the size of the gain is
worth being precise about. It is not that the effects were too small. Requiring both focal
predictions in the same study multiplies two probabilities that are each around 0.85, and the
product is what no feasible sample could reach. Dropping the conjunction lifts the English
ceiling from `r pct0(ppv("English", "ceiling_joint_bf10"))` to
`r pct1(ppv("English", "ceiling_h1b_bf10"))` and the Norwegian ceiling from
`r pct0(ppv("Norwegian", "ceiling_joint_bf10"))` to
`r pct1(ppv("Norwegian", "ceiling_h1b_bf10"))`. The conjunction was never doing theoretical
work in any case. The claim CLAPS tests is that affectedness matters *more* for passives than
for actives, which is the interaction; the passive-only slope is a check that the interaction
means what it is supposed to mean, and it belongs in the report as an estimate rather than in
the gate as a second hurdle. It is reported in full below, with its own detection rate, as a
prespecified positive control.

The threshold needs a plainer justification than its name gives it, because the Bayes factor
used here is directional and does not mean what the conventional verbal labels mean. A
directional Bayes factor of $k$ is reached exactly when the posterior probability of the
predicted sign reaches $k/(k+1)$, so a Bayes factor of 3 is a posterior sign probability of
`r num2(bf_sign(3))`, which is `r sprintf("%.2f", bf_z(3))` standard errors from zero and
corresponds to a one-sided tail probability of `r sprintf("%.2f", bf_onep(3))`. A criterion
that lenient will not carry a confirmatory claim through review, whatever it is called, and
an earlier draft of this report recommended it. A Bayes factor of 6 is
`r sprintf("%.2f", bf_onep(6))` on the same scale and 10 is `r sprintf("%.3f", bf_onep(10))`;
matching a one-sided 5% test would need about `r sprintf("%.0f", bf_for_p(0.05))`. The
threshold is set at 6 because it is the strongest criterion the design can meet at the cap
while leaving the confirmatory test adequately powered, and because stating its tail
equivalent alongside it is more honest than borrowing a label from a different quantity.

At that threshold the pooled test detects the interaction in
`r pooled_combined("p_h1b_bf6")$txt` of simulated studies, combining the pooled cells at or
below the cap. Cell by cell the pooled arm carries only 16 to 30 replicates and no single
cell settles anything; combined, and given a design in which power is flat in sample size,
that is the best estimate available. At a Bayes factor of 10 the same figure is
`r pooled_combined("p_h1b")$txt`, which is why 10 is not the recommendation.

Each language is then reported against the same criterion, as an estimate with its full
posterior rather than as a second gate. English reaches
`r pct0(h1bv("English", 80, "p_h1b_bf6"))` `r h1bci("English", 80, "p_h1b_bf6")` on the
interaction at 80 participants and Norwegian `r pct0(h1bv("Norwegian", 80, "p_h1b_bf6"))`
`r h1bci("Norwegian", 80, "p_h1b_bf6")`, so a team publishing a single-language study in
either can state a conventional power figure for the confirmatory prediction and defend it.
Turkish reaches `r pct0(h1bv("Turkish", 80, "p_h1b_bf6"))`
`r h1bci("Turkish", 80, "p_h1b_bf6")`, and its ceiling on the interaction is
`r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` at a Bayes factor of 10 however many
participants are recruited. Turkish contributes fully to the pooled confirmatory result and
is reported as estimation on its own. That is not a shortfall to be fixed by recruitment; it
is what 72 verbs buy in a language whose interaction is the weakest of the three. Turkish is
also the one language whose passive-only slope is the better-detected of its two focal
predictions, `r pct0(jval("Turkish", 80, "p_h1a"))` against
`r pct0(h1bv("Turkish", 80))` at a Bayes factor of 10, which is worth saying plainly to
anyone reading the Turkish results on their own.

Eighty participants is the recommendation, and 100 would not change it. Between those two
samples the simulations differ by less than their own resolution, and the analytic bounds say
the most that 20 further participants could add is a few points, because the binding
information is in the verbs. The case for 80 is that it is the number the teams can actually
collect.

## What Is Settled and What Is Still Open

The design analysis varies several things at once, and it is worth separating the questions
the results have answered from those that remain a matter of judgement. Three findings are
settled and no further simulation will overturn them. Power for the affectedness effect is
governed by the verb count rather than the participant count, so recruitment alone cannot
rescue a language that is short of target. English and Turkish do not reach the conventional
target on both focal predictions at any feasible per-language sample, whereas Norwegian
does. Pooling the three languages multiplies the verb-level information the estimate rests
on and is the only remedy that stays inside the agreed cap without changing the materials.

The choices that follow from this are set out in @tbl-choices. Each row is a decision for the
collaborators, with what the evidence says about it and what, if anything, is still
outstanding.

<!-- This table is tall enough to break across a page, which strands a single row
     under the caption and repeats the header. The break below keeps it whole. If the
     text above is substantially rewritten, check the rendered page and adjust. -->

`\newpage`{=latex}

| Choice | Decision | What settled it |
|:-----------------|:----------------------|:-----------------------------------------------|
| Adequacy rule | The interaction alone | The conjunction of both focal predictions is what no feasible sample reaches; dropping it lifts the English ceiling from `r pct0(ppv("English", "ceiling_joint_bf10"))` to `r pct0(ppv("English", "ceiling_h1b_bf10"))`. The passive-only slope is reported as a positive control |
| Evidence threshold | Bayes factor of 6 | The strongest criterion the design meets at the cap. A directional Bayes factor of 3 is a one-sided tail probability of `r sprintf("%.2f", bf_onep(3))`, too lenient to carry a confirmatory claim; 10 leaves the pooled test at `r pct0(pooled_combined("p_h1b")$p)` |
| Sample size | 80 per language | Between 80 and 100 the simulations differ by less than their own resolution, and the verb count binds. 80 is what the teams can collect |
| Confirmatory test | Pooled cross-linguistic | The theoretical claim is cross-linguistic, and the verb-level information is greatest there |
| Turkish | Estimation only | Its ceiling on the interaction is `r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` at a Bayes factor of 10 whatever the sample. Settled; further cells will not change it |
| Verb pool | `r ppv("English", "n_verbs")` verbs, fixed | Closed by the project: the limit is the supply of suitable verbs, not session length, so the pool cannot be enlarged |
| Stopping rule | Fixed sample, announced in advance | Closed by the project: teams must take a stated sample size to their ethics committees. The statistical objection to sequential testing does not apply, but the operational one does |

: Design decisions, with what settled each. {#tbl-choices}

`\vspace{12pt}`{=latex}

None of these now waits on further computation. Two were closed by the project rather than by
the simulations, and are recorded as such. The rest follow from results already in hand, and
the simulations have said everything they can about them.

One caveat belongs with the whole table. Every rate in this report is computed over all
fitted cells, whether or not they pass the convergence screen, which is the honest choice
because a real analysis would not discard a study for imperfect sampling. The proportion
passing that screen falls as the sample grows, so rates at different sample sizes are
averages over somewhat different mixtures of clean and untidy fits. On the screened subset
the ordering of adjacent sample sizes sometimes reverses. This does not affect the
recommendation, which rests on a criterion the design clears with room to spare and on a
sample size chosen for feasibility, but it is a reason to read small differences between
adjacent cells as noise rather than as structure.

## The Pooled Analysis: Rationale and Expectations

Following the recommendation above, the pooled cross-linguistic design analysis has been
implemented and validated. Each simulated study draws every
language's effects from that language's own pilot posterior, exactly as in the per-language
assurance runs, and simulates the three datasets. It then analyses them together with the
cross-language model used in the earlier synthesis work, which carries correlated random
slopes by participant, by verb and by language. The evidence criteria are unchanged,
namely directional Savage-Dickey Bayes factors on the pooled focal coefficients under the
same zero-centred priors. Under the 80-participant-per-language cap set by the project
lead, the pooled grid covers samples from 50 to 100, and all simulation re-runs are capped
at 100 participants and 72 verbs, with the larger runs cancelled. The decision arm, at 70, 80 and 100 participants, carries 120 per cell, which
brings the Monte-Carlo error near a power of 80% down to about four percentage points.
The pooled analytic ceiling given below is computed from the fitted per-language models
rather than from a single pooled fit. A fit of the full pooled model to the combined real pilot data, as a check that the model
settles on real rather than simulated data, remains outstanding.
`r pooled_interim`

The rationale is theoretical before it is statistical. The hypothesis CLAPS tests is
cross-linguistic, and the evidence for it has always been read across languages, most
explicitly in the meta-analytic synthesis of the five earlier studies
[@ambridgeArnonBekman2023]. The per-language analyses
treat each language as a self-contained confirmatory study, which is exactly why each
inherits its own verb-count ceiling. The pooled model instead tests the general claim
directly. Its population-level affectedness slope and interaction are the cross-linguistic
effects, and its by-language random slopes let each language depart from them, so
generality is tested rather than assumed.

Statistically, pooling eases the binding constraint identified above. The precision floor
on the affectedness effect is set by the verb-level information, and the pooled analysis
draws on about 210 verb-by-language units instead of 72, which shrinks the floor by roughly
the square root of three. The pilot slopes point the predicted way in all three languages,
at `r num2(ppv("English", "slope_per_sd"))`, `r num2(ppv("Turkish", "slope_per_sd"))` and
`r num2(ppv("Norwegian", "slope_per_sd"))` logits per one within-language standard deviation
of affectedness for the main effect and
`r mnum2(ppv("English", "interaction_per_sd"))`,
`r mnum2(ppv("Turkish", "interaction_per_sd"))` and
`r mnum2(ppv("Norwegian", "interaction_per_sd"))` on the same scale for the interaction,
so their average, which leans on the more precisely estimated languages, sits comfortably
clear of zero. Partial pooling also shares strength in the other direction, since each language's
estimate borrows from the others in proportion to its own uncertainty, and Turkish gains
most from that. The fitted coefficients on the model's Gelman-scaled predictor are about
twice these per-standard-deviation values because that predictor has a standard deviation
of approximately 0.5.

The expected differences from the per-language results follow from this. First, the
ceilings should effectively disappear as constraints. With the pooled effects near the
averages above, and a precision floor of about `r num2(pooled_floor_per_sd)` logits per
standard deviation, both pooled
focal ceilings sit near 100%. The pooled test should therefore be limited by ordinary sampling noise rather than by
a structural cap, and the joint power should cross 80% within the swept range, plausibly
between about 70 and 130 participants per language. Second, the limiting quantity changes. It is no longer the verb count but the variation
between languages, especially in the interaction, whose pilot values span
`r mnum2(min(pp$interaction_per_sd))` to `r mnum2(max(pp$interaction_per_sd))` per standard
deviation. That variation
does not shrink with more verbs or more participants, and with only three languages its variance is only weakly pinned down by the data and rests partly on the prior [@gelman2006], so it is the number to
watch in the results and in sensitivity checks. Under the pilot values, it still leaves the
pooled ceilings near `r pct0(pooled_het_ceiling)` or higher. Third, the interpretation changes. A
strong pooled Bayes factor alongside a wide Turkish-specific interval is a coherent and
likely outcome, and it is the same pattern the cross-linguistic literature shows. The
per-language conclusions above stand unchanged. Norwegian is confirmable on its own, and
the English and Turkish per-language joint tests remain capped whatever the pooled result.

If the simulations return materially weaker power than these expectations, the likely cause
is the between-language variance component rather than the focal effects, and the pooled
test's sensitivity to that component's prior will be examined before any recommendation is
drawn from it.

# Limitations and Status

The status differs across the two analyses. The literature-anchored cross-check is complete
for English and Turkish and points to a modest sample at the full verb set. The
data-grounded analysis, which is the primary basis, is substantially complete. Norwegian is
finished and is adequately powered at about 100 participants. English and Turkish fall short
of the joint power target at every sample size run so far, and only their largest cells are
still computing. The finding that participants alone cannot bring English and Turkish to
target rests on the affectedness effect being verb-limited. The 400-participant cells now
confirm that for English at `r jval("English", 400, "reps")` replicates, and the analytic
ceilings bound it from above, so the running cells are expected to confirm the plateau
rather than change its direction.

Two aspects of the simulation record qualify how far these figures can be pressed. Each
assurance replicate reuses one of the first 24 to 40 saved posterior draws rather than draws
spread evenly across the whole posterior. Checked against the full posterior, those blocks
are representative for English and Norwegian and sit slightly below the mean for the Turkish
affectedness slope, so the simulated Turkish figures are, if anything, mildly pessimistic.
The analytic ceilings use all 8,000 draws and are unaffected, and the pooled
cross-linguistic design analysis and the decision arm sample their draw indices randomly
over the full pilot posteriors, so the limitation does not reach them. Replication itself is
complete, since the original single-language grid finished all 720 of its cells and the low
counts noticed mid-run were cells still computing.

Each data-grounded condition rests on `r min(jp_a$reps)` to `r max(jp_a$reps)` simulated
studies, and each literature-anchored cell on 50, so the figures carry a Monte-Carlo error
of several percentage points near a power of 80%. Individual cells close to the threshold should be read as near target rather
than pinned, and the conclusions rest on the pattern across sample sizes rather than on any
single cell. On the literature effects, the adequate sample at 72 verbs is reported as at most
50, since the main sweep does not go below 50. It is not pinned down further, as the
recommendation rests on the data-grounded analysis rather than on this cross-check.

This is an a-priori power analysis, a simulation under assumed effects to size the study, and
it is distinct from the confirmatory analysis of the real ratings that will be run once data
collection is complete. The analysis code and configuration are openly available, and the
report rebuilds from the committed summary tables without the raw data, which remain private
to CLAPS until the project reaches a more advanced stage. The report will be finalised once
the data-grounded runs complete and the largest English and Turkish samples are in.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix A}\\[3pt]
{\bfseries Composition of the Pilot Sample}
\end{center}
```

This appendix describes the pilot sample from which the affectedness ratings and the
descriptive trends in @fig-pilot are drawn, and against which the data-grounded analysis
fits its models. The counts are read directly from the harmonised pilot data.

```{r tbl-pilot}
#| tbl-cap: "Composition of the Harmonised Pilot Sample, by Language"

comp <- pilot_composition
kable(comp, align = "lrrrrr", format.args = list(big.mark = ","))
```

`\vspace{-8pt}`{=latex}

*Note.* Counts are read from the harmonised pilot data. Items are verb-by-agent-gender
combinations, so the item count is twice the verb count. Norwegian lacks a pseudo-passive,
hence two sentence types rather than three.

`\vspace{12pt}`{=latex}

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix B}\\[3pt]
{\bfseries Key Code for the Simulation and the Models}
\end{center}
```

This appendix reproduces the central code, lightly abridged, from the analysis scripts. The
code shown is the data-generating process for the literature-anchored cross-check, in
`R/06_simulate_design.R`, together with the analysis model in `R/04_model_formulas.R` and
the Bayes factor in `R/05_hypothesis_tests.R`. The data-grounded analysis uses the same
analysis model and Bayes factor, but draws its data-generating effects, random-effect
covariance and thresholds from the fitted pilot models, and reads each verb's affectedness
from the pilot data. Its assurance and safeguard variants are in
`R/10_simulate_from_pilot.R` and `R/10_simulate_from_pilot_v2.R`.

Each simulated dataset is drawn from the same cumulative-logit model that is later fitted.
In the cross-check, affectedness is drawn once per verb across a plausible range, which makes
it a verb-level predictor, and the ordinal response follows from the cumulative-logit linear
predictor.

```r
# DGP for the literature-anchored cross-check, abridged from R/06_simulate_design.R.
verb_semantics <- runif(n_verbs, min = -0.5, max = 0.5)   # one value per verb

# For participant pid, verb vid and sentence type st:
sem   <- verb_semantics[[vid]]
b_int <- if (st == "Active")              beta_active_interaction * sem
         else if (st == "Pseudo_Passive") beta_pseudo_interaction * sem
         else 0
eta   <- part_intercepts[pid] + part_slopes[pid] * sem +
         verb_intercepts[vid] + beta_semantics * sem + b_int

cum_probs <- plogis(thresholds - eta)        # seven ordered categories
probs     <- diff(c(0, cum_probs, 1))
resp      <- sample(seq_len(7), size = 1, prob = probs)
```

The analysis model is the maximal random-effects structure consistent with the design,
fitted with the cumulative-logit family.

```r
# Analysis model, from R/04_model_formulas.R.
brms::bf(
  Response ~ S_Type * Semantics_scaled +
    (1 + S_Type * Semantics_scaled | Participant) +
    (1 + S_Type | Verb),
  family = brms::cumulative(link = "logit", threshold = "flexible")
)
```

Evidence for each prediction is a directional Savage-Dickey Bayes factor, with the prior
sampled alongside the posterior and the posterior mass taken in the predicted direction.

```r
# Directional Savage-Dickey Bayes factor, from R/05_hypothesis_tests.R.
# post_vals and prior_vals are draws of the focal coefficient; the negative
# direction is shown, as for the primary prediction H1b.
post_prob  <- mean(post_vals  < 0)
prior_prob <- mean(prior_vals < 0)
bf_10      <- (post_prob / (1 - post_prob)) / (prior_prob / (1 - prior_prob))
```

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix C}\\[3pt]
{\bfseries Full Bayes-Factor Operating Characteristics}
\end{center}
```

This appendix reports the operating characteristics for all four predictions in every available
cell of the corrected literature-anchored run with at least ten replicates, namely the
probability of exceeding the primary and secondary Bayes-factor thresholds, the median Bayes
factor and the convergence rate, set out in @tbl-opchar, which covers `r lit_language_text`.

```{r tbl-opchar}
#| tbl-cap: "Probability of Exceeding the Primary and Secondary Bayes-Factor Thresholds, Median Bayes Factor and Convergence Rate, by Design Condition and Prediction"

hyp_lab <- c(H1a_semantics_positive = "H1a", H1b_active_interaction_negative = "H1b",
             H2a_pseudo_positive = "H2a", H2b_pseudo_negative = "H2b")
opc <- exc |>
  filter(reps >= 10) |>
  mutate(Hyp = factor(hyp_lab[hypothesis], levels = c("H1a", "H1b", "H2a", "H2b"))) |>
  transmute(Language = language, Verbs = n_verbs, N = n_participants, Hyp,
            `P(BF>=10)` = sprintf("%.0f%%", 100 * p_bf_primary),
            `P(BF>=3)`  = sprintf("%.0f%%", 100 * p_bf_secondary),
            `Median BF` = signif(median_bf, 2),
            Conv        = sprintf("%.0f%%", 100 * p_convergence_ok),
            Reps        = reps) |>
  arrange(Language, Verbs, N, Hyp)
kable(opc, booktabs = TRUE, longtable = TRUE, align = "lrrlrrrrr",
      col.names = c("Language", "Verbs", "N", "Prediction", "P(BF>=10)",
                    "P(BF>=3)", "Median BF", "Convergence", "Reps")) |>
  kableExtra::kable_styling(font_size = 10, latex_options = c("repeat_header"))
```

`\vspace{-18pt}`{=latex}

*Note.* The probability columns give the proportion of simulated studies reaching a Bayes
factor of 10 and of 3, Median BF the median across them to two significant figures,
Convergence the proportion of fits meeting the criteria, and Reps the number of studies in
the cell. Cells with fewer than ten are omitted as still accumulating, and Norwegian has no
pseudo-passive, hence half as many rows.

`\vspace{12pt}`{=latex}

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix D}\\[3pt]
{\bfseries Guide to the Analysis Code}
\end{center}
```

This appendix maps the analysis code, which is openly available at
<https://github.com/pablobernabeu/CLAPS-analysis>. It is meant to make the repository
navigable for a reader who wants to check how a number in this report was produced, or to
reuse part of the machinery. The raw pilot data are not published, since they are
participant-level records held privately until the project is further advanced, so the
repository ships the simulation and analysis code together with the aggregated summaries
the report reads.

Three entry points cover most purposes. A reader checking a reported number wants
`outputs/`, which holds the aggregated summaries and is where every figure in this report
comes from. A reader checking how a simulated study is built and analysed wants
[`R/06_simulate_design.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design.R)
for the literature-anchored generator,
[`R/10_simulate_from_pilot_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot_v2.R)
for the pilot-grounded one, and
[`R/05_hypothesis_tests.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/05_hypothesis_tests.R)
for the Bayes factor. A reader intending to rerun the analysis wants `scripts/`, whose files
are numbered in the order they are meant to run, together with the matching submitters in
`hpc/`. @tbl-repo sets out what each top-level directory holds.

`\vspace{10pt}`{=latex}

| Location | What it holds |
|:------------------|:-----------------------------------------------------------|
| `R/` | The building blocks: data validation and preprocessing (`01`, `02`), priors and model formulas (`03`, `04`), the Bayes factor (`05`), the two simulation engines (`06`, and `10` with `11`), convergence diagnostics (`07`) and summarising (`08`) |
| `scripts/` | Runnable steps, numbered in running order, from harmonising the pilot (`00`) through fitting the pilot models, generating a design grid, running one cell, to aggregating results (`aggregate_pilot.R`, `aggregate_pooled.R`, `aggregate_literature.R`) |
| `hpc/` | One SLURM submitter per arm. Each takes a grid file and an output directory, and runs one grid row per array task |
| `config/` | The design grids, one row per simulated condition, plus the model and prior configuration |
| `outputs/` | The aggregated summaries this report reads, and the figure it displays |
| `tests/` | Unit tests for the validation, prior, diagnostic and Bayes-factor logic |
| `reports/` | This report, its bibliography and citation style |

: Layout of the public analysis repository. {#tbl-repo}

`\vspace{12pt}`{=latex}

The unit of work is the cell, meaning one simulated condition: a language, a participant
count, a verb count and a replicate index. A grid file in `config/` lists the cells, an
array task runs one row, and each cell is saved as a single `.rds` file holding its
summary, its Bayes factors and its convergence diagnostics. The aggregation scripts read a
directory of those files and reduce them to the tables in `outputs/`. Reading one saved
cell is therefore the quickest way to see exactly what the pipeline computes per simulated
study.

Two conventions are worth knowing before reading the code. Names carrying `pilot` refer to
the data-grounded analysis that conditions on the fitted pilot models and forms the primary
basis of this report, while names carrying `literature` refer to the cross-check that
conditions on the published effect sizes. Separately, the suffix `_v2` on a simulation
engine marks the extended version rather than a correction, so both files are live and the
pair should be read together.

Three files in `R/` generate simulated studies, `06`, `10` and `11`, and between them they
implement two engines rather than three. The engines differ in where the true effects come
from, and the second of them is written at two scopes.

The literature-anchored engine is
[`R/06_simulate_design.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design.R).
It invents a study from assumed effect sizes taken from the published estimates, drawing
participants, verbs and affectedness values from scratch. Because the parameters are
stipulated rather than estimated, extending it from one language to several is only a matter
of adding a language level to the same recipe, which is why this engine covers both scopes
in one file: `simulate_claps_data()` for a single language and
`simulate_claps_data_multilanguage()` beside it.

The pilot-grounded engine is
[`R/10_simulate_from_pilot.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot.R).
It takes the true effects from a model already fitted to the pilot data, so a simulated study
inherits that language's estimated slopes, thresholds and variance components together with
its real verbs and their measured affectedness.
[`R/10_simulate_from_pilot_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot_v2.R)
loads that file and adds the assurance and safeguard variants, which draw the effects from
the pilot posterior rather than fixing them at its mean.

Extending this second engine across languages is not a matter of adding a level, because each
language carries its own fitted model. A pooled study has to be assembled from three separate
sets of parameters, simulated language by language and then combined before a single
cross-language model is fitted to the result. That orchestration is what
[`R/11_simulate_pooled_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/11_simulate_pooled_v2.R)
adds. It loads the `_v2` file, calls its per-language simulator once for each language, binds
the three datasets and fits the pooled model, so it is the same engine at a wider scope rather
than a third method.

The files therefore stack rather than compete, each loading the one below it and adding to it:
`11` on `10_v2` on `10`, and separately
[`R/06_simulate_design_gelman.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design_gelman.R)
on `06`. That last file is the narrowest of them. It changes only the scale on which the
analysed predictor is expressed, leaving the simulated responses identical, so that the
literature-anchored cells are analysed on the same footing as the confirmatory model.

Both engines end in the same place. Each defines a cell runner, `run_design_cell()` in `06`
and `run_databased_cell_v2()` and `run_pooled_cell_v2()` in `10_v2` and `11`, which simulates
one dataset, fits it with
[`R/04_model_formulas.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/04_model_formulas.R)
and the priors in
[`R/03_define_priors.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/03_define_priors.R),
computes the Bayes factors with
[`R/05_hypothesis_tests.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/05_hypothesis_tests.R)
and records the diagnostics from
[`R/07_extract_diagnostics.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/07_extract_diagnostics.R),
then writes the `.rds` for that cell. Only the source of the true effects differs; the model,
the evidence criterion and the convergence standard are shared, which is what makes the two
analyses comparable. @tbl-engines summarises where each engine's effects come from and which
file implements it at each scope.

`\vspace{10pt}`{=latex}

| Engine | Where the effects come from | Single language | All three pooled |
|:----------------------|:--------------------------------|:---------------------|:---------------------|
| Literature-anchored | Published effect sizes, stipulated | `06` | `06`, the same file |
| Pilot-grounded | Estimated from the fitted pilot models | `10`, extended by `10_v2` | `11`, which loads `10_v2` |

: The two simulation engines and the files implementing them. {#tbl-engines}

`\vspace{12pt}`{=latex}

Following one arm end to end is the quickest way to see how the pieces connect. For the
pooled analysis reported here, the grid is `config/design_grid_pooled_v2.csv`, one row per
simulated study;
[`hpc/submit_pooled_v2_array.sh`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/hpc/submit_pooled_v2_array.sh)
submits one array task per row;
[`scripts/07_pooled_cell_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/scripts/07_pooled_cell_v2.R)
runs a single row by calling `run_pooled_cell_v2()`; and
[`scripts/aggregate_pooled.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/scripts/aggregate_pooled.R)
reduces the resulting `.rds` files to `pooled_power_pilot.csv` in
`outputs/design_summary_pilot/`, which is the table this report reads for every pooled
figure it quotes.
