---
title: "Preliminary Sample-Size Analysis for CLAPS"
author: "Pablo Bernabeu and CLAPS collaborators"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
format:
  pdf:
    geometry: "margin=1in"
    linestretch: 1.25
    fig-cap-location: top
    fig-pos: "htbp"
    colorlinks: true
    code-overflow: wrap
    include-in-header:
      text: |
        \definecolor{subtitlegray}{gray}{0.35}
        \setkomafont{date}{\normalfont\normalsize\color{subtitlegray}}
        % Microtypographic character protrusion and font expansion for an
        % evener, less rivered right margin under justification.
        \usepackage{microtype}
        % Discourage widows and orphans (a lone line stranded at the top or
        % bottom of a page), which a technical report with many short
        % paragraphs is otherwise prone to.
        \widowpenalty=10000
        \clubpenalty=10000
        % Allow a little more stretch between words before over-full lines are
        % risked, so justification does not force awkward hyphenation.
        \setlength{\emergencystretch}{3em}
        % Wrap long lines inside code blocks at the margin (breakanywhere lets a
        % long path or token break even without a space) rather than letting
        % them run past the right edge.
        \usepackage{fvextra}
        \fvset{breaklines=true,breakanywhere=true}
        % Show the "Abstract" label, which KOMA suppresses by default. The
        % report follows APA 7 elsewhere, where the abstract is labelled.
        \KOMAoptions{abstract=on}
        % APA 7 caption style: bold label, italic title, applied to figures and tables.
        \usepackage{caption}
        \DeclareCaptionFormat{apa7}{#1\\[5pt]#3}
        \DeclareCaptionFormat{apa7par}{#1\par\vspace{5pt}#3}
        \captionsetup{format=apa7, labelsep=none, labelfont=bf, textfont=it,
          justification=raggedright, singlelinecheck=false, skip=6pt}
        \captionsetup[longtable]{format=apa7par, labelsep=none, labelfont=bf,
          textfont=it, justification=raggedright, singlelinecheck=false, skip=6pt}
        % Top-align figures that land on their own float page (avoids vertical centring).
        \makeatletter
        \setlength{\@fptop}{0pt}
        \setlength{\@fpsep}{12pt plus 1fil}
        \setlength{\@fpbot}{0pt plus 1fil}
        \makeatother
bibliography:
  - ../references.bib
csl: apa.csl
link-citations: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)

# Tidy per-cell Bayes-factor exceedance from the literature-anchored run,
# aggregated by scripts/aggregate_literature.R (one row per language x N x
# n_verbs x hypothesis). p_bf_primary is the proportion of replicates with BF10 >= 10
# (the preregistered detection threshold). Prose about which languages and
# sample sizes this covers is derived below rather than written by hand, since
# an earlier summary was aggregated before one language had finished and the
# report went on asserting that gap after the cells had landed.
exc <- read_csv("../outputs/design_summary_literature/bf_exceedance_literature.csv",
                show_col_types = FALSE) |>
  filter(n_verbs %in% c(40, 72))   # 12 and 20 verbs are well below the design's count
lit_languages <- sort(unique(exc$language))
lit_language_text <- if (length(lit_languages) > 2) {
  paste(paste(head(lit_languages, -1), collapse = ", "), "and", tail(lit_languages, 1))
} else paste(lit_languages, collapse = " and ")
lit_n_min <- min(exc$n_participants)
lit_n_max <- max(exc$n_participants)

focal <- c(H1a = "H1a_semantics_positive",
           H1b = "H1b_active_interaction_negative")

# Binding focal power = the smaller of the two focal exceedance probabilities, the
# quantity the design as first specified would have been judged on, when both focal
# predictions had to clear 0.80 in the same study.
foc <- exc |>
  filter(hypothesis %in% focal) |>
  mutate(H = names(focal)[match(hypothesis, focal)]) |>
  select(language, n_participants, n_verbs, reps, H, p_bf_primary) |>
  pivot_wider(names_from = H, values_from = p_bf_primary) |>
  mutate(binding = pmin(H1a, H1b))

# Restrict the figure/recommendation to cells with (near) the full 50 replicates.
MIN_REPS <- 45L
foc_full <- filter(foc, reps >= MIN_REPS)

pct <- function(x) ifelse(is.na(x), "-", sprintf("%.0f%%", 100 * x))

# Crossing point under the joint rule the design originally specified: smallest N at
# which the binding focal power reaches 0.80, per language x verb count.
rec <- foc_full |>
  filter(binding >= 0.80) |>
  group_by(language, n_verbs) |>
  summarise(rec_N = min(n_participants), .groups = "drop")

# Convergence robustness for the cell cited in the text (English, 72 verbs, N = 50):
# binding focal power computed over all fits vs over the converged subset only.
cc <- exc |>
  filter(language == "English", n_verbs == 72, n_participants == 50,
         hypothesis %in% focal) |>
  summarise(all = min(p_bf_primary), conv = min(p_bf_primary_conv),
            n_conv = max(n_conv))

# The public report is deliberately independent of participant-level pilot data.
# It uses only disclosure-reviewed derivatives that are already represented in
# the published PDF: a raster figure and aggregate sample-composition counts.
pilot_composition <- read_csv(
  "../outputs/design_summary_pilot/pilot_sample_composition.csv",
  show_col_types = FALSE
)
pilot_figure <- "../outputs/figures/pilot_acceptability_by_gender_report.png"

# Pilot-grounded (v2) summaries, aggregated on the cluster by
# scripts/aggregate_pilot.R: per-cell joint power, fitted pilot
# parameters and the analytic detection ceilings. Reading them here keeps
# every simulation result in the report programmatic.
jp <- read_csv("../outputs/design_summary_pilot/joint_power_pilot.csv",
               show_col_types = FALSE)
pp <- read_csv("../outputs/design_summary_pilot/pilot_params_ceilings.csv",
               show_col_types = FALSE) |>
  mutate(language = factor(language, levels = c("English", "Turkish", "Norwegian"))) |>
  arrange(language)

pct0  <- function(x) sprintf("%.0f%%", 100 * x)
pct1  <- function(x) sprintf("%.1f%%", 100 * x)
num2  <- function(x) sprintf("%.2f", x)
mnum2 <- function(x) sprintf("$%.2f$", x)   # math mode, for a proper minus sign
ppv   <- function(lang, col) pp[[col]][pp$language == lang]

# Cells from the v2 grid are shown and cited only once they hold a minimum number of
# replicates; thinner cells are suppressed rather than reported at a precision they do
# not support. Nothing further is being simulated.
MIN_REPS_V2 <- 10L
jp_a  <- filter(jp, mode == "assurance", reps >= MIN_REPS_V2)
jval  <- function(lang, N, col) jp_a[[col]][jp_a$language == lang & jp_a$n_participants == N]
jrow  <- function(lang, N) filter(jp_a, language == lang, n_participants == N)
jcount <- function(lang, N) {
  r <- jrow(lang, N)
  as.integer(round(r$p_joint * r$reps))
}
jratio <- function(lang, N) {
  r <- jrow(lang, N)
  sprintf("%d/%d", jcount(lang, N), r$reps)
}
jmcse <- function(lang, N) {
  r <- jrow(lang, N)
  sqrt(r$p_joint * (1 - r$p_joint) / r$reps)
}
max_n <- function(lang) max(jp_a$n_participants[jp_a$language == lang])
nor_cross     <- min(filter(jp_a, language == "Norwegian", p_joint >= 0.80)$n_participants)
tur_max_joint <- max(filter(jp_a, language == "Turkish")$p_joint)
# Smallest sample at which the interaction alone is adequately powered. The test is
# the exact one-sided 95% lower bound, not the point estimate: an earlier version
# took the first cell whose observed rate reached 0.80 and so reported Norwegian as
# adequate at 50 participants on the strength of 22 successes out of 24, a cell whose
# lower bound is 0.76. A sample size is only recommendable if the simulation is
# precise enough to support the recommendation.
lower95 <- function(k, n) if (n == 0) NA_real_ else stats::binom.test(k, n, alternative = "greater")$conf.int[1]
h1b_cross <- function(lang, col = "p_h1b", target = 0.80) {
  d <- filter(jp_a, language == lang)
  ok <- vapply(seq_len(nrow(d)), function(i)
    isTRUE(lower95(round(d[[col]][i] * d$reps[i]), d$reps[i]) >= target), logical(1))
  if (!any(ok)) NA_integer_ else as.integer(min(d$n_participants[ok]))
}

# The interaction alone (H1b), which is the confirmatory prediction. Rates come with
# an exact interval, because every one of them is a proportion over a few dozen to a
# few hundred replicates and none is known to the precision a bare percentage implies.
h1bv  <- function(lang, N, col = "p_h1b") jval(lang, N, col)
h1bci <- function(lang, N, col = "p_h1b") {
  r <- jrow(lang, N); k <- round(r[[col]] * r$reps)
  b <- stats::binom.test(k, r$reps)$conf.int
  sprintf("[%.2f, %.2f]", b[1], b[2])
}

# Pooled arm, combined across sample sizes at or below the cap. The arm is flat in N,
# which is what a verb-limited design predicts and what the cells show, so combining
# them is the best available estimate; cell by cell the pooled arm holds only a few dozen
# replicates and no single cell settles anything.
pooled_combined <- function(col, max_n = 100) {
  d <- pooledp[pooledp$n_participants <= max_n, ]
  k <- sum(round(d[[col]] * d$reps)); n <- sum(d$reps)
  b <- stats::binom.test(k, n)$conf.int
  list(p = k / n, k = k, n = n, lo = b[1], hi = b[2],
       txt = sprintf("%s [%.2f, %.2f]", pct0(k / n), b[1], b[2]))
}

# A directional Bayes factor of k is met exactly when the posterior probability of the
# predicted sign reaches k/(k+1). Stating the equivalent one-sided tail probability
# keeps the reader from reading these thresholds as the point-null Bayes factors that
# the conventional verbal labels describe.
bf_sign  <- function(k) k / (k + 1)
bf_z     <- function(k) qnorm(bf_sign(k))
bf_onep  <- function(k) 1 - pnorm(bf_z(k))
bf_for_p <- function(p) { z <- qnorm(1 - p); s <- pnorm(z); s / (1 - s) }

sg_max        <- function(lang) max(filter(jp, mode == "safeguard", reps >= MIN_REPS_V2,
                                           language == lang)$p_joint)

# Pooled expectations, derived from the per-language pilot parameters: the
# verb-information floor with all three languages' verbs pooled, and the
# ceiling implied by the between-language spread of the interaction (the
# quantity that shrinks with neither verbs nor participants).
pooled_floor       <- mean(pp$tau_verb_intercept) / sqrt(sum(pp$sxx))
pooled_floor_per_sd <- pooled_floor * mean(pp$affectedness_sd)
pooled_het_ceiling <- pnorm(abs(mean(pp$interaction_mean)) /
                             (sd(pp$interaction_mean) / sqrt(nrow(pp))) - qnorm(10 / 11))

# Literature cross-check reference cells: the full verb set and the reduced
# 40-verb set at the smallest sample, and the secondary (H2) detection range
# at the full verb set.
lit50    <- filter(foc_full, n_verbs == 72, n_participants == 50)
lit40_50 <- filter(foc_full, language == "English", n_verbs == 40,
                   n_participants == 50)$binding
h2_72    <- filter(exc, hypothesis == "H2a_pseudo_positive", n_verbs == 72,
                   reps >= MIN_REPS)$p_bf_primary

# How much simulation the report rests on, and how it is distributed across the
# arms. Derived, because these counts move whenever a cell lands, and because the
# distribution is the point: the arm carrying the recommendation is the smallest.
lit_cells   <- unique(exc[, c("language", "n_verbs", "n_participants", "reps")])
n_assurance <- sum(jp$reps[jp$mode == "assurance"])
n_safeguard <- sum(jp$reps[jp$mode == "safeguard"])
n_lit       <- sum(lit_cells$reps)

# Pooled cross-linguistic results, aggregated by scripts/aggregate_pooled.R.
# The pooled arm may still be pre-first-cell, so the report degrades to the
# forward-looking wording when the file is absent or has no usable cells.
pooled_csv <- "../outputs/design_summary_pilot/pooled_power_pilot.csv"
pooledp <- if (file.exists(pooled_csv)) read_csv(pooled_csv, show_col_types = FALSE) else NULL
# Pooled joint power at the smallest per-language sample, by threshold, for the
# recommendation. Degrades to a neutral phrase before the first cell lands.
pool50 <- if (!is.null(pooledp) && nrow(pooledp) && 50 %in% pooledp$n_participants) {
  pooledp[pooledp$n_participants == 50, ]
} else NULL
# What the pooled evidence implies about the true detection rate, and what the run
# considered and not pursued would have shown. A uniform prior on the rate gives a Beta
# posterior; the predictive draws a rate from that posterior and then the successes
# among the new cells. Derived rather than asserted, so these figures move with the
# data instead of going stale the way a written-out number would. The seed fixes
# the predictive so the report renders identically twice.
pooled_post <- function(col = "p_h1b_bf6", max_n = 100) {
  d <- pooledp[pooledp$n_participants <= max_n, ]
  k <- sum(round(d[[col]] * d$reps)); n <- sum(d$reps)
  list(k = k, n = n, a = 1 + k, b = 1 + n - k)
}
pp_atleast <- function(x, ...) { z <- pooled_post(...); 1 - pbeta(x, z$a, z$b) }
pp_below   <- function(x, ...) { z <- pooled_post(...); pbeta(x, z$a, z$b) }

# 380 cells were submitted; about one in fifteen has been lost in this arm, so the
# predictive assumes 355 usable.
N_NEW_SUBMITTED <- 380L
N_NEW_USABLE    <- 355L
pooled_predictive <- local({
  z  <- pooled_post()
  set.seed(20260819)
  th <- rbeta(200000, z$a, z$b)
  new <- rbinom(200000, N_NEW_USABLE, th) / N_NEW_USABLE
  ck  <- rbinom(200000, N_NEW_USABLE, th) + z$k
  cn  <- N_NEW_USABLE + z$n
  list(
    p_within_5  = mean(abs(new - z$k / z$n) <= 0.05),
    p_beyond_10 = mean(abs(new - z$k / z$n) >  0.10),
    p_settles   = mean(qbeta(0.025, 1 + ck, 1 + cn - ck) >= 0.80)
  )
})

pooled_interim <- if (!is.null(pooledp) && nrow(pooledp) && 50 %in% pooledp$n_participants) {
  h1b6 <- pooled_combined("p_h1b_bf6")
  h1b10 <- pooled_combined("p_h1b")
  sprintf(paste("On the criterion this report recommends, the",
                "interaction alone at a Bayes factor of 6, the pooled test succeeds in %s of",
                "simulated studies, combining the %d replicates at or below the cap. At a Bayes",
                "factor of 10 the same figure is %s, which is why the stronger threshold is not",
                "recommended. No single pooled cell settles anything on its own: each holds",
                "between %d and %d replicates, and their Monte-Carlo errors run to about nine",
                "percentage points, which is why they are combined here. Restricting the",
                "estimate to fits that pass every convergence check moves it by a few points in",
                "either direction and does not change the conclusion."),
          h1b6$txt, h1b6$n, h1b10$txt, min(pooledp$reps), max(pooledp$reps))
} else "First results are expected within days."
```

`\begin{abstract}\normalsize\noindent`{=latex} CLAPS asks whether the acceptability of a
passive sentence depends on how strongly the event affects the person or thing it happens
to, and whether that pattern holds across languages. This report asks how large a sample is
needed to answer that convincingly. The answer turns on an unusual feature of the design.
Affectedness is a property of the verb, not of the participant, so precision is governed by
how many verbs are tested. Past a point, recruiting more participants stops helping.

At the `r ppv("English", "n_verbs")` verbs used here, that limit binds hardest on the design
as first specified, which asked for both focal predictions in the same study. English's
ceiling under that rule is `r pct0(ppv("English", "ceiling_joint_bf10"))` however many people
are recruited, held down by the affectedness slope for passives, which tops out at
`r pct0(ppv("English", "ceiling_h1a_bf10"))`. Judged on the active-by-affectedness
interaction alone, the prediction the theory turns on, the ceilings are
`r pct1(ppv("English", "ceiling_h1b_bf10"))` for English and
`r pct1(ppv("Norwegian", "ceiling_h1b_bf10"))` for Norwegian. Turkish is the exception under
every rule, its interaction ceiling being `r pct0(ppv("Turkish", "ceiling_h1b_bf10"))`.
Earlier, more optimistic simulations had assumed less variation between verbs than the pilot
data show.

The recommendation is a single confirmatory test: the three languages analysed together,
judged on the interaction alone, at a Bayes factor of 6, with 80 participants per language
and the current verb pool. That test succeeds in `r pooled_combined("p_h1b_bf6")$txt` of
simulated studies, with English at `r pct0(h1bv("English", 80, "p_h1b_bf6"))` and Norwegian
at `r pct0(h1bv("Norwegian", 80, "p_h1b_bf6"))` individually. The threshold matters because
this Bayes factor is directional: 3 corresponds to a one-sided tail probability of
`r sprintf("%.2f", bf_onep(3))`, too lenient for a confirmatory claim, while 6 is
`r sprintf("%.2f", bf_onep(6))` and is the strongest the design meets within the cap. Turkish
contributes to the combined test and is reported on its own as estimation. The verb pool
cannot grow and the sample is fixed in advance.
`\end{abstract}`{=latex}

<!-- The body starts on its own page. Left to flow, it followed the abstract down the
     title page and finished hard against the bottom margin. -->

`\newpage`{=latex}

This note reports preliminary sample-size guidance for CLAPS (A Crosslinguistic Approach
to Passive Semantics). The guidance comes from a Bayes-factor design analysis. A design analysis of this kind
simulates the planned model many times under plausible true effects,
then asks how often a given sample size would yield compelling evidence for the focal
predictions.

Two versions of the analysis are run. The primary one is data-grounded. It fits the planned
model to each language's real pilot data and simulates new studies from that fit, so its
results reflect the effects actually observed in the CLAPS pilot. The collaborators asked
the design analysis to use that pilot as its basis. A single pilot's point estimate is noisy
and tends to overstate power, the concern that Albers and Lakens (2018) raise for
pilot-based power analyses. The data-grounded analysis avoids that point estimate in two
ways. First, on each simulated study it draws the focal effects from the pilot posterior,
the range of effect sizes the pilot data still leave plausible, so that the reported
probability of a decisive Bayes factor reflects the pilot's uncertainty about the size of
the effect. This yields an assurance, a power averaged over the plausible sizes of the
effect, not a power computed at one assumed size. Second, it also evaluates a deliberately
conservative lower bound on the effect, as a safeguard against an optimistic pilot. A second
version, for comparison, is based instead on the pooled effect sizes from the five
previously published studies. It provides an independent cross-check against the broader
evidence base.

The two versions agree on the central structural point. The power to detect the affectedness
effect is governed by the number of verbs in the materials, because affectedness is a
property of the verb. They diverge on whether a modest sample is enough. On the
published-literature effects, a per-language sample of at most 50 participants would suffice
at the full verb set. On the pilot effects, and with the verb-level variability measured
from the pilot instead of assumed, the picture is less favourable. Norwegian reaches the
target at about `r nor_cross` participants, but English and Turkish do not reach it at any
simulated sample size. For those two languages, the binding constraint is the number of
verbs, and in Turkish also the strength of the effect itself, so adding participants alone
cannot bring the design to the usual power target. Because the data-grounded analysis is the
agreed primary basis, this is the finding that should guide the design. The report
quantifies these limits analytically, explains why the two analyses disagree, and sets out
design options that remain feasible for a volunteer consortium. The recommended option is a
pooled cross-linguistic analysis, and the section on it below gives its rationale and the
first results.

The data-grounded analysis is substantially complete. Results are in for all three languages
up to `r max_n("Norwegian")` participants, for English up to `r max_n("English")` and for
Turkish up to `r max_n("Turkish")`. Nothing further is being simulated: the project has
capped recruitment at 100 participants, so the cells above that are of record interest only.
Each simulated condition rests on `r min(jp_a$reps)` to `r max(jp_a$reps)` studies, so the
individual figures carry a Monte-Carlo error of several percentage points, that is, the
uncertainty that comes from basing each figure on a limited number of simulated studies. The
400-participant simulations now confirm the English plateau, the levelling-off of power
short of the target, so it no longer rests on extrapolation. The recommendation below is
final on this evidence. The one result still outstanding is the refit described under the
pooled model, which bears on the pooled figures and not on the sample size.

# Background and Motivation

The question CLAPS addresses is whether the acceptability of passive and related sentence
types tracks the semantic affectedness of the patient, and whether that relationship
differs by sentence type. Affectedness, the degree to which a patient or theme undergoes a
change of state or is causally impinged upon by the action, is a core dimension of
argument structure [@dowty1991; @beavers2011]. It has long been argued to be the semantic
heart of the passive, which carries the meaning that the surface subject is in a state or
circumstance characterised by the agent having acted upon it [@pinkerLebeauxFrost1987].
Across languages, adult acceptability-judgement and comprehension studies support the view
that passive acceptability is semantically structured rather than purely syntactic
[@ambridgeBidgoodPineRowlandFreudenthal2016; @aryawibawaAmbridge2018;
@darmasetiyawanAmbridge2022; @liuAmbridge2021; @bidgoodPineRowlandAmbridge2020;
@ambridgeArnonBekman2023].

The cross-linguistic evidence base to date spans English, Indonesian, Balinese, Hebrew and
Mandarin. A Bayesian meta-analytic synthesis finds a consistent pattern across them
[@ambridgeArnonBekman2023]. Affectedness raises acceptability for both passives and actives,
more strongly for passives, and does not raise it for pseudo-passives. The
affectedness constraint has not yet been tested in Turkish or Norwegian. CLAPS extends the
paradigm to these two typologically and morphologically distinct languages. No prior
language-specific effect-size estimates exist for them, so the CLAPS pilot supplies the
first such estimates. The data-grounded analysis uses these pilot estimates as its primary
basis, and the cross-linguistic literature provides an independent comparison, which turns
out to be the more optimistic of the two.

# The Acceptability Model and the Focal Predictions

Each acceptability judgement is a discrete rating from 1 to 7. These ratings are modelled
with a Bayesian cumulative-logit mixed-effects model, an ordinal model that respects the
discreteness, ordering and bounded range of rating data while leaving the underlying scale
free to vary across conditions [@burknerVuorre2019]. Treating the same ratings as
continuous, under a Gaussian likelihood, biases the regression coefficients and inflates
error rates where floor or ceiling effects are present [@liddellKruschke2018]. Recent
empirical work shows that modelling them cumulatively recovers psycholinguistic effects
more accurately [@verissimo2021; @taylorRousseletScheepersSereno2023].

The expected pattern is already visible in the pilot data, drawn from the sample whose
composition by language is given in @tbl-pilot in Appendix A. @fig-pilot shows the raw
acceptability ratings against agent semantic affectedness, separately by sentence type,
referent gender and language. The affectedness slope is positive and steepest for the
passive in English and Norwegian, and shallower for the active and the pseudo-passive. The
trends are very similar for male and female referents within each language. This is
reassuring, because the focal effects should not be driven by a referent-gender imbalance.

```{r fig-pilot}
#| fig-cap: "Raw Pilot Acceptability by Agent Semantic Affectedness, Sentence Type, Referent Gender and Language"
#| fig-width: 6.5
#| fig-pos: "H"
#| fig-align: center
#| out-width: "6.1in"

knitr::include_graphics(pilot_figure)
```

`\vspace{-8pt}`{=latex}

*Note.* Panels are by language across rows and by referent gender across columns. Points
are individual pilot ratings, jittered, and lines are per-sentence-type linear trends with
95% confidence bands. Affectedness is z-scored within language. The pilot data are used for
calibration only. Norwegian has no pseudo-passive.

`\vspace{12pt}`{=latex}

The model predicts acceptability from sentence type, from verb affectedness, and from their
interaction. The passive is the reference level, and the active and the pseudo-passive are
the comparisons. The focal predictions concern the affectedness slope in the passive and how
it changes across sentence types. The primary prediction (H1b) is one-tailed and holds that
affectedness raises passive acceptability more than it raises active acceptability, so the
active-by-affectedness interaction is negative. A supporting prediction (H1a) holds that
affectedness raises acceptability in the passive itself, so the passive affectedness slope
is positive. The confirmatory decision rests on H1b alone. H1a is a prespecified positive
control: it checks that the interaction means what it is meant to mean, and is reported as
an estimate. An earlier version of the design required both predictions in the same study,
and the results below are what led to dropping that requirement. A secondary, two-tailed
prediction (H2) concerns the pseudo-passive-by-affectedness interaction. It does not gate
the recommendation and does not apply to Norwegian, which has no pseudo-passive.

# Design-Analysis Method

The design analysis estimates statistical power by simulation. Each simulated dataset is
drawn from the same cumulative-logit model that will be used for analysis, under assumed
values for the effects of interest. Each dataset is then analysed in the same way. The
proportion of simulated studies that yield decisive evidence estimates the power at that
sample size. The whole procedure stays within the Bayesian framework. This matters because
the confirmatory test is a Bayes factor: the power to return a decisive Bayes factor is not
the same quantity as the power to return a significant *p*-value.

A central feature of the design is that affectedness is a property of the verb. Each verb
carries a single affectedness value, held constant across participants and sentence types,
exactly as in the real materials. The number of verbs therefore determines the precision
with which the affectedness effect can be estimated. The verb count is treated as a design
dimension alongside the participant count.

The two analyses differ in where they take the assumed true effects and the affectedness
values. The data-grounded analysis, which is primary, fits the planned model to each
language's pilot data. It then takes the population effects, the random-effect covariance
and the ordinal thresholds from that fit, and it reads each verb's actual affectedness score
from the pilot data. The literature-anchored cross-check instead sets the population effects
to values from the cross-linguistic meta-analysis. On the standardised affectedness
predictor, the passive affectedness slope is set near the pooled value of about 0.23 logits
per standard deviation (a logit is a unit of the model's underlying scale, not a raw point
on the 1 to 7 rating), with a negative active-by-affectedness interaction of about $-0.15$
and a small positive pseudo-passive-by-affectedness interaction. In the cross-check,
affectedness is spread across a plausible range instead of read from the pilot, and the
by-participant and by-verb random-effect standard deviations are set to realistic
magnitudes. The referent-gender covariate is held out of both preliminary runs and is
examined separately.

The analysis model uses the maximal random-effects structure consistent with the design,
meaning the most complex structure the design supports, so any effect that can vary across
participants or verbs is allowed to do so. It has correlated by-participant slopes for
sentence type, affectedness and their interaction, together with by-verb intercepts and
sentence-type slopes [@barrLevyScheepersTily2013]. Because affectedness does not vary within
a verb, a by-verb affectedness slope cannot be estimated and is omitted, so affectedness
enters as a population-level effect, one estimated for the sample as a whole. The priors are
weakly to moderately informative and centred at zero on the focal slopes, so that the
directional tests are not pre-loaded. A normal prior of standard deviation 0.5 is placed on
the affectedness slope and on the active interaction, and a slightly wider normal prior of
0.6 on the pseudo-passive interaction. The sentence-type main effects take a wider normal
prior. Student-*t* priors are placed on the ordinal thresholds and on the by-participant and
by-verb standard deviations, and an LKJ prior, the standard prior for a correlation matrix,
on the correlations among those random effects [@gelman2006; @lewandowskiKurowickaJoe2009].

Evidence for each prediction is summarised by a directional Savage-Dickey Bayes factor, a
number that weighs how strongly the data favour the predicted direction of the effect over
there being no effect, obtained by sampling the prior alongside the posterior
[@wagenmakersLodewyckxKuriyalGrasman2010]. A simulated study counts as a detection when its
Bayes factor reaches 10, the threshold the project initially set for strong evidence. Under
the zero-centred prior, this is not a stringent bar. The Bayes factor used here is
directional, meaning the posterior odds that the coefficient carries the predicted sign
divided by the same odds under the prior, so under a symmetric prior a value of 10 is odds
of 10 to 1 on the direction. That is a posterior probability of 0.91, or about 1.34 standard
errors from zero, whereas a one-sided *p* < .05 requires 1.64, so the Bayesian criterion is
if anything the less demanding of the two. It is not the point-null Bayes factor of a
default two-sided test, which is a different quantity on a different scale. Power is the
proportion of simulated studies that reach it. Results are reported at Bayes factors of 10,
6 and 3 throughout, since the stored per-cell Bayes factors allow any threshold to be scored
without re-running anything. The recommendation, set out below, is judged on the interaction
alone at a Bayes factor of 6, and fixes a single sample size in advance instead of taking
the smallest one that meets a power target.

Each replicate is fitted with four Markov chains of 3,000 iterations, 1,000 of them warm-up,
at the same target acceptance rate of 0.99 and maximum tree depth of 12 that the
confirmatory analysis will use, which leaves 8,000 post-warm-up draws. The proposal
specifies a heavier run for that single confirmatory fit, sixteen chains of 5,000 iterations
with 2,000 of warm-up, or 48,000 draws. The shorter chains add Monte-Carlo noise to each
simulated Bayes factor without shifting it systematically, so the power figures below are,
if anything, mildly conservative.

The literature-anchored grid crosses the three languages with participant counts from 50 to
120. The output read here covers `r lit_language_text` from `r lit_n_min` to `r lit_n_max`
participants at verb counts of 40 and 72. Each completed cell holds 50 replicates, giving a
Monte-Carlo standard error of about five to six percentage points near a power of 80%, so
estimates close to the threshold are not perfectly ordered across adjacent sample sizes. The
central code for the data-generating process, the model and the Bayes factor is reproduced
in Appendix B. The full operating characteristics of this run, cell by cell, are reported in
@tbl-opchar in Appendix C, and Appendix D maps the whole repository, including which file
implements each of the two simulation engines described above and how the pooled analysis
reuses one of them.

# Results

## Data-grounded results

The data-grounded analysis is the primary basis for the recommendation, and it is now
substantially complete. In its assurance form it draws the focal effects afresh from each
language's pilot posterior on every simulated study, so the reported power carries the
pilot's own uncertainty about the size of those effects. @tbl-pilot-joint reports the joint
power, the proportion of simulated studies in which both focal predictions clear the
Bayes-factor threshold together, at the pilot's real verb count.

```{r tbl-pilot-joint}
#| tbl-cap: "Preliminary Data-Grounded Joint Power by Participant Count and Language, Assurance Variant"

pj <- tidyr::crossing(n_participants = sort(unique(jp_a$n_participants)),
                      language = c("English", "Turkish", "Norwegian")) |>
  left_join(select(jp_a, language, n_participants, p_joint),
            by = c("language", "n_participants")) |>
  mutate(cell = ifelse(is.na(p_joint), "-", pct1(p_joint))) |>
  select(-p_joint) |>
  pivot_wider(names_from = language, values_from = cell)
kable(pj[, c("n_participants", "English", "Turkish", "Norwegian")],
      format = "latex", booktabs = TRUE, longtable = TRUE,
      align = "rrrr",
      col.names = c("Participants", "English", "Turkish", "Norwegian")) |>
  kableExtra::kable_styling(
    latex_options = "repeat_header",
    repeat_header_method = "replace",
    repeat_header_text = "\\textbf{Table 1 (continued)}"
  )
```

`\vspace{-8pt}`{=latex}

*Note.* Joint power is the proportion of studies in which both focal predictions clear a
Bayes factor of 10 together, at 72 verbs and 66 for Norwegian. Cells holding fewer than
`r MIN_REPS_V2` replicates are excluded, since they cannot support the precision reported
here. The rest hold `r min(jp_a$reps)` to `r max(jp_a$reps)` studies each. Percentages are
shown to one decimal place to preserve the exact Monte-Carlo summaries. Norwegian at 130
participants sits a little above the approximate infinite-participant ceiling in
@tbl-ceilings, at `r jratio("Norwegian", 130)` = `r pct1(jval("Norwegian", 130, "p_joint"))`
against a Monte-Carlo standard error of `r pct1(jmcse("Norwegian", 130))`, so the difference
is sampling variation and the two agree.

`\vspace{12pt}`{=latex}

The pattern differs sharply by language. Norwegian reaches the 80% target at about
`r nor_cross` participants and is comfortably powered beyond it. English and Turkish do not
reach it at any simulated sample size. For English, the analytic ceiling is
`r pct0(ppv("English", "ceiling_joint_bf10"))`, but that is the limiting value with
unlimited participants, not the power of a finite study; its simulated joint power remains
below 80%. It is `r pct1(jval("English", 300, "p_joint"))` by 300 participants and
`r pct1(jval("English", 400, "p_joint"))` at 400. The reason is
structural. The joint requirement is held back by the affectedness main effect (H1a), whose
detection rate levels off at roughly three in four, while the interaction (H1b) climbs
comfortably past it. Because affectedness is a property of the verb, its main effect is
estimated from the 72 verbs, so adding participants does not sharpen it. The precision of
that effect, and with it the power to detect it, is set by the number of verbs. For Turkish,
both focal effects are weak in the pilot. The joint power never exceeds
`r pct1(tur_max_joint)` at any simulated sample size, and the present design does not bring
it to target.

A conservative safeguard variant, which fixes each focal effect at a cautious lower bound of
its pilot posterior, is lower again. Its highest joint power over the samples run is
`r pct0(sg_max("Norwegian"))` for Norwegian, `r pct0(sg_max("English"))` for English and
`r pct0(sg_max("Turkish"))` for Turkish. This is the design's power under a deliberately
pessimistic reading of the pilot.

The pilot posterior itself puts the probability that both focal effects have their predicted sign at `r pct1(ppv("English", "sign_prob_joint"))` for English,
`r pct1(ppv("Norwegian", "sign_prob_joint"))` for Norwegian and
`r pct1(ppv("Turkish", "sign_prob_joint"))` for Turkish, so the pilot points the right way
in all three languages. The shortfall
for English and Turkish is therefore a matter of precision rather than direction. Because
affectedness varies only
between verbs, the standard error of its effect cannot fall below the by-verb variability
divided by the spread of the affectedness scores, however many participants respond. This
is the stimulus-sampling limit long recognised in psycholinguistics [@clark1973] and
quantified for designs with crossed random effects by Westfall et al. [-@westfallKennyJudd2014;
see also @juddWestfallKenny2017; @brysbaertStevens2018]. The limit can be computed directly
from the fitted pilot models, with no simulation involved. @tbl-ceilings reports the
resulting upper bounds, the detection rates that would remain even with unlimited
participants at the current verb counts.

```{r tbl-ceilings}
#| tbl-cap: "Analytic Upper Bounds on Detection with Unlimited Participants, at the Current Verb Counts"

fmt_verbs <- function(v) ifelse(is.na(v), "> 1,200", format(v, big.mark = ","))
ce <- pp |> transmute(
  Language        = as.character(language),
  `H1a ceiling`   = pct0(ceiling_h1a_bf10),
  `H1b ceiling`   = pct0(ceiling_h1b_bf10),
  `Joint ceiling` = pct0(ceiling_joint_bf10),
  `Verbs for 80%` = fmt_verbs(verbs_for_80),
  `Verbs for 90%` = fmt_verbs(verbs_for_90)
)
kable(ce, align = "lrrrrr")
```

`\vspace{-8pt}`{=latex}

*Note.* Ceilings are the probability of a Bayes factor of at least 10 with unlimited
participants, from the `r format(ppv("English", "ndraws"), big.mark = ",")` pilot posterior
draws and the fitted by-verb covariance at 72 verbs and 66 for Norwegian. The last columns
give the verbs needed for an 80% and 90% joint ceiling. A ceiling just above target, as for
English, still leaves it out of reach, since any finite sample sits below its ceiling.

`\vspace{12pt}`{=latex}

These bounds agree closely with the simulations. English's simulated joint power climbs to
`r pct1(jval("English", 300, "p_joint"))` by 300 participants against a ceiling of
`r pct0(ppv("English", "ceiling_joint_bf10"))`. Turkish never exceeds
`r pct1(tur_max_joint)` against a ceiling of `r pct0(ppv("Turkish", "ceiling_joint_bf10"))`.
Norwegian's simulated `r pct1(jval("Norwegian", 100, "p_joint"))` at 100 participants
matches its ceiling of `r pct0(ppv("Norwegian", "ceiling_joint_bf10"))` within Monte-Carlo
error. The agreement between two entirely different routes to the same quantities,
stochastic simulation and direct calculation, is a strong check that the pipeline is sound.
A further check makes the point starkly for English. The pilot's own posterior standard
deviation for the affectedness slope, a measure of how precisely the pilot pins that slope
down, is `r num2(ppv("English", "slope_post_sd"))`, already close to the verb-count floor of
`r num2(ppv("English", "se_floor_h1a"))`. A replication with the same 72 verbs would
therefore have little room to estimate that slope more precisely than the pilot already has,
whatever its sample size. The residual gap reflects the floor formula's own approximation,
since it ignores some of the covariance in the by-verb random effects, rather than a real
target that more participants could close.

The ceilings also show what would and would not help. Relaxing the evidence criterion from a
Bayes factor of 10 to 6 or 3 lifts the English joint ceiling to
`r pct0(ppv("English", "ceiling_joint_bf6"))` and
`r pct0(ppv("English", "ceiling_joint_bf3"))`, but lifts Turkish only to
`r pct0(ppv("Turkish", "ceiling_joint_bf6"))` and
`r pct0(ppv("Turkish", "ceiling_joint_bf3"))`, so a weaker threshold cannot rescue the
Turkish test. Adding verbs raises every ceiling, because the precision floor shrinks with
the square root of the verb count. English would reach a joint ceiling of 90% at about
`r ppv("English", "verbs_for_90")` verbs, and Norwegian at about
`r ppv("Norwegian", "verbs_for_90")`. Turkish would need about
`r ppv("Turkish", "verbs_for_80")` verbs merely to reach a ceiling of 80%, beyond what a
single language can supply, so for Turkish the remedy lies elsewhere, as taken up under
Paths Forward below.

These results are preliminary in one remaining respect, since each condition rests on
`r min(jp_a$reps)` to `r max(jp_a$reps)` simulated studies and so carries a Monte-Carlo
error of a few percentage points at the better-populated sample sizes and more at the
sparser ones. The
English plateau itself, though, is now confirmed rather than extrapolated. At 400
participants, with `r jval("English", 400, "reps")` replicates in, the joint power sits at
`r pct1(jval("English", 400, "p_joint"))` (`r jratio("English", 400)`), compared with
`r pct1(jval("English", 300, "p_joint"))` (`r jratio("English", 300)`) at 300 participants.
These sparse estimates are not literally flat, but their decline is compatible with
Monte-Carlo variation and, critically, there is no sustained rise towards the target.
Overturning the plateau would
have required a sustained rise at 400 and 500, and the 400-participant data already rule
that out. No further cells are being simulated, the project having
capped recruitment at 100 participants, and by that test the sample sizes already run settle
the plateau.

## Literature-anchored cross-check

The grid is read as a surface over participant count and verb count, summarised by the
binding focal power, the smaller of the two focal detection probabilities, since the design
as first specified required both to clear 80%. The recommendation now gates on the
interaction alone, so this surface records the earlier criterion. This cross-check is the
more optimistic of the two analyses, because it fixes the effects at the pooled literature
values and carries no uncertainty about them. @fig-surface shows this surface for all three
languages. Cells outlined in gold meet the target, and the blank cells are Turkish at 40
verbs, never run at 120 participants and holding too few replicates at 50 to report. The
pattern is clear. Power increases steeply with the number of verbs and only weakly with the
number of participants.

```{r fig-surface}
#| fig-cap: "Binding Focal Power by Participant Count and Verb Count, for Each Language"
#| fig-width: 7.0
#| fig-height: 2.5
#| fig-pos: "H"

# Languages and sample sizes are taken from the data rather than listed here,
# so the panel follows whatever the aggregation covers.
grid_df <- tidyr::crossing(
  language = sort(unique(foc_full$language)),
  n_participants = sort(unique(foc_full$n_participants)),
  n_verbs = c(40, 72)
) |>
  left_join(select(foc_full, language, n_participants, n_verbs, binding),
            by = c("language", "n_participants", "n_verbs")) |>
  mutate(
    nv = factor(n_verbs, levels = c(40, 72)),
    np = factor(n_participants, levels = c(50, 60, 70, 80, 90, 100, 120)),
    lab = ifelse(is.na(binding), "", sprintf("%.0f", 100 * binding)),
    txtcol = ifelse(!is.na(binding) & binding >= 0.55, "white", "grey15")
  )

ggplot(grid_df, aes(np, nv, fill = binding)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_tile(data = filter(grid_df, !is.na(binding), binding >= 0.80),
            color = "#E6A100", linewidth = 1.1, fill = NA) +
  geom_text(aes(label = lab, color = txtcol), size = 3.1, fontface = "bold") +
  scale_color_identity() +
  scale_fill_gradientn(
    colours = c("#f7fbff", "#c6dbef", "#6baed6", "#2171b5", "#08306b"),
    limits = c(0, 1), na.value = "grey92",
    name = "Binding focal\npower",
    labels = scales::percent_format(accuracy = 1)
  ) +
  facet_wrap(~ language) +
  labs(x = expression(paste("Participants per language (", italic(N), ")")),
       y = "Number of verbs") +
  guides(fill = guide_colourbar(barheight = unit(2.8, "cm"),
                                barwidth = unit(0.4, "cm"))) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.margin = margin(t = 8, r = 4, b = 4, l = 4),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

`\vspace{-8pt}`{=latex}

*Note.* Each cell gives the binding focal power as a percentage, with darker shading for
higher power. Cells outlined in gold meet the 80% target. Blank cells were either not run or
hold too few replicates to report.

`\vspace{12pt}`{=latex}

The two axes are not equally important. At 40 verbs in English, the binding power is
`r pct0(lit40_50)` at 50 participants and reaches the 80% target only between about 60 and
90 participants, within Monte-Carlo noise. At 72 verbs, it is already
`r pct0(min(lit50$binding))` to `r pct0(max(lit50$binding))` at 50 participants. Increasing
the number of verbs from 40 to 72 raises power more than increasing the number of
participants does, as follows from affectedness being a verb-level property. @fig-curves
separates the two focal predictions and shows how each approaches the target as the sample
grows, at each verb count.

```{r fig-curves}
#| fig-cap: "Detection Probability for the Focal Predictions by Verb Count and Language"
#| fig-width: 6.8
#| fig-height: 2.8
#| fig-pos: "H"

z <- 1.96
curve_df <- foc_full |>
  select(language, n_participants, n_verbs, reps, H1a, H1b) |>
  tidyr::pivot_longer(c(H1a, H1b), names_to = "Prediction", values_to = "p") |>
  mutate(
    den = 1 + z^2 / reps,
    ctr = (p + z^2 / (2 * reps)) / den,
    hw  = z * sqrt(p * (1 - p) / reps + z^2 / (4 * reps^2)) / den,
    lo  = pmax(0, ctr - hw), hi = pmin(1, ctr + hw),
    nv  = factor(paste(n_verbs, "verbs"), levels = paste(c(40, 72), "verbs"))
  )

pd <- position_dodge(width = 4)
ggplot(curve_df, aes(n_participants, p, colour = Prediction, shape = Prediction)) +
  geom_hline(yintercept = 0.80, linetype = "dashed", colour = "grey55") +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 3, linewidth = 0.4, position = pd) +
  geom_line(linewidth = 0.7, position = pd) +
  geom_point(size = 2, position = pd, fill = "white", stroke = 0.7) +
  facet_grid(nv ~ language) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format(accuracy = 1)) +
  scale_colour_manual(values = c(H1a = "#0072B2", H1b = "#D55E00")) +
  scale_shape_manual(values = c(H1a = 21, H1b = 24)) +
  labs(x = expression(paste("Participants per language (", italic(N), ")")),
       y = "Detection probability", colour = "Prediction", shape = "Prediction") +
  theme_bw(base_size = 10) +
  theme(legend.position = "top", panel.grid.minor = element_blank(),
        strip.text = element_text(face = "bold"))
```

`\vspace{-8pt}`{=latex}

*Note.* Points are detection probabilities over the simulated studies in each complete cell,
with Wilson 95% intervals reflecting Monte-Carlo uncertainty. The dashed line marks the 80%
target. Points are dodged so the two predictions do not overlap.

`\vspace{12pt}`{=latex}

In the literature-anchored output, which covers `r lit_language_text`, the binding
focal power at the full 72-verb set is already between
`r pct0(min(lit50$binding))` and `r pct0(max(lit50$binding))` at 50 participants, the
smallest sample simulated, in every language included in this cross-check. On the
literature effects, the adequate per-language sample is therefore at most 50 participants at
the full verb set. This is the optimistic reading. The data-grounded results above, which
condition on the weaker pilot effects, do not reach the target for English or Turkish at any
sample size. Because 50 is the smallest sample in the main grid, the analysis places
the adequate sample at or below 50 without determining it more precisely. Pinning it down further, below 50 participants, is not pursued here, since the
recommendation rests on the data-grounded analysis rather than on this cross-check.

## Why the Two Analyses Disagree

The gap between the two analyses calls for an explanation, because the literature-anchored
grid reports `r pct0(min(lit50$binding))` to `r pct0(max(lit50$binding))` at 50
participants on the same 72 verbs at which the data-grounded joint power never reaches 80%.
Expressed per standard deviation of affectedness, the pilot's passive slope is
`r num2(ppv("English", "slope_per_sd"))` logits for English,
`r num2(ppv("Norwegian", "slope_per_sd"))` for Norwegian and
`r num2(ppv("Turkish", "slope_per_sd"))` for Turkish, against a pooled literature value of
about 0.23. The Turkish active interaction, at
`r mnum2(ppv("Turkish", "interaction_per_sd"))`, matches the literature's $-0.15$ almost
exactly, and the English interaction, at `r mnum2(ppv("English", "interaction_per_sd"))`, is
almost three times the literature value, so the CLAPS pilot effects are not unusually
small.

The disagreement comes instead from the noise side of the simulation, in two specific
assumptions. First, the literature-anchored simulator assumed a by-verb intercept standard
deviation of 0.5, a measure of how much verbs differ in their baseline acceptability, where the fitted pilot models put it at
`r num2(ppv("English", "tau_verb_intercept"))` for English,
`r num2(ppv("Turkish", "tau_verb_intercept"))` for Turkish and
`r num2(ppv("Norwegian", "tau_verb_intercept"))` for Norwegian. The precision floor on the
affectedness slope scales directly with this quantity, so assuming 0.5 where the fitted
value is `r num2(ppv("English", "tau_verb_intercept"))` overstates the attainable evidence
considerably. Second, the literature-anchored simulator included no by-verb sentence-type
slopes at all, allowing no verb-to-verb variation in how sentence type shifts the ratings, where the fitted pilots put their standard deviation at
`r num2(ppv("English", "tau_verb_active"))` for English,
`r num2(ppv("Turkish", "tau_verb_active"))` for Turkish and
`r num2(ppv("Norwegian", "tau_verb_active"))` for Norwegian. That omission left the
interaction test with no verb-level noise floor in the simulation, which is why the earlier
grid found the interaction so easy to detect. Both assumptions were reasonable defaults
before the pilot models existed. The pilot has now measured the quantities they stood in
for, and the data-grounded analysis uses the measured values. The earlier optimism, in
short, came less from the assumed effect sizes than from understating how much verbs vary
[@clark1973; @westfallKennyJudd2014].

## Effect-Size Qualification

One assumption deserves particular care, because it bears directly on the absolute sample
size. The numbers shown here condition on assumed effects set to the cross-linguistic pooled
posterior means from a Bayesian meta-analytic synthesis of the passive-affectedness
literature, which is a central estimate, not an optimistic one [@ambridgeArnonBekman2023].
On the simulator's internal predictor scale, these appear as larger nominal values, but
expressed per standard deviation of affectedness, they are about 0.23 and $-0.15$ logits,
matching the pooled literature values. The larger nominal numbers therefore reflect the
scaling alone. Two cautions nonetheless apply. The first is the one the previous section
sets out. The fitted pilot models put the verb-level variability well above the values this
grid assumed, and that difference, more than any difference in effect size, made the grid
optimistic. The pilot effects themselves are broadly in line with the pooled literature
values once both are expressed per standard deviation of affectedness, though the Turkish
terms sit at the weak end, with the interaction at about
`r mnum2(ppv("Turkish", "interaction_per_sd"))` in those units. The second is that the
literature is itself subject to publication bias and to the winner's curse, the tendency of
published estimates, especially from small and low-powered studies, to overstate true
magnitudes [@gelmanCarlin2014; @ioannidis2008; @button2013]. Design analyses anchored to
such estimates can therefore be optimistic about power [@vasishth2018; @albersLakens2018].

These cautions are the reason the data-grounded analysis is the primary basis for the
recommendation. It draws the focal effects from each language's own pilot posterior, and it
also evaluates a conservative lower bound. It therefore conditions on the effects observed
in CLAPS and carries their uncertainty through to the reported power. Its results are
reported above. They agree with this cross-check for Norwegian, powered at about 100
participants, and they diverge from it for English and Turkish, where the weaker pilot
effects hold the joint power below target at every sample size run.

A separate point concerns the scale on which the prior sits, not the size of the assumed
effect. The literature-anchored grid behind the figures expresses affectedness on the
simulator's internal predictor, whose standard deviation is about 0.29. The real
confirmatory analysis and the data-grounded analysis both place it on the analysis
pipeline's Gelman scaling, which divides a predictor by twice its standard deviation and so
leaves it with a standard deviation of 0.5, and the same weakly informative prior is applied
throughout. A Bayes factor reflects the width of the prior relative to the coefficient it
constrains. Holding the prior fixed while the predictor scale differs therefore leaves the
literature-anchored grid mildly optimistic about the focal-slope Bayes factor, by a factor
of about 1.7, relative to the analysis that will actually be run. The qualitative conclusion
is unaffected, because the verb count dominates the result, the focal Bayes factors at the
full verb set lie far above the threshold, and the data-grounded analysis already uses the
correct scale. The cross-check grid was not re-run on the corrected scale and will not be,
since the recommendation rests on the data-grounded analysis alone. The figures here should
therefore be read as a mild upper bound on the Bayes-factor evidence.

For an effect that lives at the level of the participant, a weaker true effect simply needs
a larger sample, and the rough guide is that a true effect around two-thirds of the assumed
magnitude roughly doubles the sample needed. The affectedness main effect lives at the level
of the verb, however, where its precision is set by the number of verbs, so a weaker true
effect cannot be recovered by recruiting more participants. This is exactly what the
data-grounded results show. At the full verb count and the pilot's own effects, the
affectedness power for English settles below the target and does not climb with sample size,
and for Turkish neither focal effect is strong enough to bring the joint power to target.
The effective levers are therefore the number of verbs, and for Turkish the strength of the
interaction. Where feasible, the design should still be planned against a deliberately
conservative, or safeguard, effect size rather than a point estimate [@perugini2014], and
the data-grounded analysis builds this in directly through its assurance and safeguard
variants. Discounting the literature-anchored grid towards weaker effects makes the same
point: re-evaluated at 75% and 60% of the assumed effect magnitude, with the 60% level
approximating the weaker effects seen in the pilot, it needs larger samples. The
data-grounded analysis conditions on those pilot effects directly, so the primary results
above already reflect that discounting.

## Convergence and Computation

Convergence is sound across the grid, with no divergent transitions anywhere, which are the
warning sign of an unreliable fit. At the smaller verb count, essentially every fit meets
the strict criteria: an R-hat below 1.01, the check that the separate chains have settled on
the same answer, together with bulk and tail effective sample sizes of at least 400, which
measure how much independent information the draws carry. At 72 verbs, the maximal model
occasionally falls just short of these criteria, in about one fit in ten at the smaller
samples and up to one in five at the larger ones. The failures show up on the R-hat and
effective-sample-size thresholds while divergences stay at zero, so they reflect how
efficiently the sampler explores the posterior, not a fault in the specification.
Restricting the power calculation to the fully converged fits leaves the estimates
essentially unchanged. For English at 72 verbs and 50 participants, the binding focal power
is `r pct(cc$all)` over all 50 fits and `r pct(cc$conv)` over the `r cc$n_conv` converged
fits, so the recommendation does not depend on the difference.

## Secondary Prediction

The pseudo-passive-by-affectedness interaction reaches a detection probability of about
`r pct0(min(h2_72))` to `r pct0(max(h2_72))` at 72 verbs. This remains below the focal
predictions and reflects its smaller
assumed magnitude, and the opposite-direction test is near zero throughout, as expected.
This prediction is secondary and two-tailed, and it does not constrain the sample-size
recommendation. Full operating characteristics for all four predictions, including the
convergence rates, are reported in Appendix C.

# Paths Forward

The analytic bounds turn the shortfall into a set of concrete design choices. Four options
follow, in roughly increasing order of departure from the current plan. Two of them have
since been closed by the project, and are kept here because the reasoning that ruled them
out is part of the record.

The design is now capped at 80 participants per language, with 100 the absolute maximum any
team would be asked for. The verb pool cannot grow beyond 72, because suitable verbs are
scarce in every language and session length was never the constraint. The third option below
is therefore unavailable rather than merely expensive. Collection must be at a single sample
size fixed and announced in advance, because the teams have to take a stated number to their
ethics committees, which rules out the fourth option. Many of the teams have not run a study
of this type before and could not defend a more elaborate protocol to their committees.

The first option changes nothing about the materials and revisits the adequacy rule. The
design's primary prediction is H1b, the active-by-affectedness interaction, with H1a as a
supporting prediction. Gating each language on the primary prediction alone, with the
passive slope reported as estimation, brings English to 80% on the interaction at
`r h1b_cross("English")` participants and Norwegian at `r h1b_cross("Norwegian")`, at the
current verb counts and at the strong threshold. Those crossing points are the smallest
samples at which the *lower bound* of the simulated rate reaches 80%, not the first cell
whose point estimate happens to. With a few dozen replicates per cell the two can differ by
a whole sample size. No feasible English sample can satisfy the joint gate on both
predictions. This option does not help Turkish, whose interaction ceiling is
`r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` even on its own.

The second option makes the cross-linguistic claim the primary confirmatory test. The
theoretical claim CLAPS addresses is explicitly cross-linguistic, that affectedness
structures passive acceptability across languages, and the meta-analytic synthesis of the
earlier studies frames the evidence in exactly those terms [@ambridgeArnonBekman2023]. A
hierarchical model fitted to all three languages jointly tests that claim directly. It also
multiplies the verb-level information, roughly 210 verb-by-language units in place of 72,
which divides the precision floor by about the square root of three. On the pilot effects,
every language's focal terms point the predicted way, so the pooled test starts from a
favourable position. Turkish then contributes fully to the confirmatory result, and its
language-specific effects are reported as estimates with credible intervals, which reflects
what a 72-verb single-language design can support. The pooled design analysis runs on the same machinery used here, and it is now built. Its rationale and
the first results are set out below.

The third option expands the verb pool without lengthening any participant's session. Verbs,
not participants, are the scarce resource, and a planned-missingness design assigns each
participant a random subset of a larger pool [@grahamTaylorOlchowskiCumsille2006;
@littleRhemtulla2013]. Each volunteer still rates about 72 items, so the 45-minute session
is unchanged, while the pool grows to, say, 120 to 150 verbs. At about
`r ppv("English", "verbs_for_90")` verbs, the English joint ceiling reaches 90%, and the
pooled cross-linguistic test gains in proportion. The cost falls on materials, namely
affectedness norms and translations for the new verbs. This option is closed. Planned
missingness would work if the 72-verb limit constrained the session, but it constrains the
pool: the teams are already struggling to find 72 suitable verbs in most languages, so there
is no larger set to sample from. That matters beyond this one option, because every ceiling
in this report is set by the verb count. With the pool fixed, those ceilings are now
permanent features of the design that no materials budget could move, and Turkish, which
would need about `r ppv("Turkish", "verbs_for_80")` verbs to reach 80% on its own, is out of
reach for good.

The fourth option spends volunteer capacity more efficiently instead of demanding more of
it. A sequential design analyses the data in waves and stops as soon as the Bayes factor
reaches a preset bound or the sample reaches an agreed maximum. This typically needs
markedly fewer participants than a fixed-sample design with the same operating
characteristics, meaning the same chances of reaching a decisive result
[@schonbrodtWagenmakersZehetleitnerPerugini2017; @stefanGronauSchonbrodtWagenmakers2019].
Optional stopping of this kind poses no problem for Bayes-factor inference [@rouder2014].

This option is also closed, though not for the reason usually given, since the statistical
objection to it does not apply. Nothing in the Bayes-factor framework penalises
optional stopping. The obstacle is operational: the collecting teams have to put a specific
number of participants in front of an ethics committee, and a protocol whose sample size
depends on interim results is one that many of them could not explain or defend, however
sound it is. Operating characteristics count for nothing if the study cannot be administered
by the people running it. The sample size below is therefore a single fixed number, announced
in advance.

The recommendation is the first two options taken together, and nothing else: the pooled
cross-linguistic analysis is the confirmatory test, it is gated on the interaction alone, its
threshold is a Bayes factor of 6, and every team collects 80 participants.

Gating on the interaction alone makes the design work, and the effects themselves were never
the problem. When a study has to satisfy both predictions at once, its chances are the two
ceilings multiplied together, and the passive-only slope is the one that holds them down. It
tops out at `r pct0(ppv("English", "ceiling_h1a_bf10"))` in English and
`r pct0(ppv("Norwegian", "ceiling_h1a_bf10"))` in Norwegian however many participants are
recruited, while the interaction reaches `r pct1(ppv("English", "ceiling_h1b_bf10"))` and
`r pct1(ppv("Norwegian", "ceiling_h1b_bf10"))`. Dropping the conjunction lifts the English
ceiling from `r pct0(ppv("English", "ceiling_joint_bf10"))` to
`r pct1(ppv("English", "ceiling_h1b_bf10"))` and the Norwegian ceiling from
`r pct0(ppv("Norwegian", "ceiling_joint_bf10"))` to
`r pct1(ppv("Norwegian", "ceiling_h1b_bf10"))`. There was little theoretical reason for the
conjunction either. The claim CLAPS tests is that affectedness matters *more* for passives
than for actives, which is the interaction. The passive-only slope checks that the
interaction means what it should, and that belongs in the report as an estimate and not in
the gate as a second hurdle. It is reported in full below, with its own detection rate, as a
prespecified positive control.

The threshold needs justifying in plainer terms than its name suggests, because the Bayes
factor used here is directional and does not mean what the conventional verbal labels mean.
A directional Bayes factor of $k$ is reached exactly when the posterior probability of the
predicted sign reaches $k/(k+1)$, so a Bayes factor of 3 is a posterior sign probability of
`r num2(bf_sign(3))`, which is `r sprintf("%.2f", bf_z(3))` standard errors from zero and
corresponds to a one-sided tail probability of `r sprintf("%.2f", bf_onep(3))`. A criterion
that lenient will not carry a confirmatory claim through review, whatever it is called, and
an earlier draft of this report recommended it. A Bayes factor of 6 is
`r sprintf("%.2f", bf_onep(6))` on the same scale and 10 is
`r sprintf("%.3f", bf_onep(10))`; matching a one-sided 5% test would need about
`r sprintf("%.0f", bf_for_p(0.05))`.

The project's own starting point was a Bayes factor of 10, which the internal record of
decisions carries as the primary criterion for the focal interaction, with 3 as a secondary
one. Those figures were entered as a working placeholder pending this design analysis, not
as a settled choice, and this is the analysis they were waiting for. It does not support 10.
At the feasible cap the pooled test detects the interaction in
`r pooled_combined("p_h1b")$txt` of simulated studies at that threshold, against
`r pooled_combined("p_h1b_bf6")$txt` at a Bayes factor of 6. No amount of recruiting closes
that gap, for the reason the rest of this report sets out: the affectedness effect is
bounded by the number of verbs. Holding to 10 would mean preregistering a confirmatory test
this design cannot pass, which is the outcome a design analysis exists to prevent. The
revision to 6 is therefore forced by the evidence.

The threshold is set at 6 because it is the strongest criterion the pooled test comes close
to meeting at the cap, and because quoting its tail equivalent alongside it is clearer than
borrowing a label from a different quantity. How close is a question the simulations cannot
yet settle, and the next paragraph gives the figure with the interval it carries.

At that threshold the pooled test detects the interaction in
`r pooled_combined("p_h1b_bf6")$txt` of simulated studies, combining the pooled cells at or
below the cap. Cell by cell the pooled arm carries only
`r min(pooledp$reps[pooledp$n_participants <= 100])` to
`r max(pooledp$reps[pooledp$n_participants <= 100])` replicates and no single
cell settles anything on its own. Since power here is flat in sample size, combining them
gives the best estimate available. At a Bayes factor of 10 the same figure is
`r pooled_combined("p_h1b")$txt`, which is why 10 is not the recommendation.

Each language is then reported against the same criterion, as an estimate with its full
posterior. English reaches `r pct0(h1bv("English", 80, "p_h1b_bf6"))`
`r h1bci("English", 80, "p_h1b_bf6")` on the interaction at 80 participants and Norwegian
`r pct0(h1bv("Norwegian", 80, "p_h1b_bf6"))` `r h1bci("Norwegian", 80, "p_h1b_bf6")`, so a
team publishing a single-language study in either can state a conventional power figure for
the confirmatory prediction and defend it. Turkish reaches
`r pct0(h1bv("Turkish", 80, "p_h1b_bf6"))` `r h1bci("Turkish", 80, "p_h1b_bf6")`, and its
ceiling on the interaction is `r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` at a Bayes factor
of 10 however many participants are recruited. Turkish contributes fully to the pooled
confirmatory result and is reported as estimation on its own. That is not a shortfall to be
fixed by recruitment; it is what 72 verbs buy in a language whose interaction is the weakest
of the three. Turkish is also the one language whose passive-only slope is the
better-detected of its two focal predictions, `r pct0(jval("Turkish", 80, "p_h1a"))` against
`r pct0(h1bv("Turkish", 80))` at a Bayes factor of 10. Anyone reporting Turkish on its own
should lead on the slope rather than the interaction.

Eighty participants is the recommendation, and 100 would not change it. Between those two
samples the simulations differ by less than their own resolution, and the analytic bounds say
the most that 20 further participants could add is a few points, because the binding
information is in the verbs. The case for 80 is that it is the number the teams can
collect.

## What Is Settled and What Is Still Open

The design analysis varies several things at once, and it is worth separating the questions
the results have answered from those that remain a matter of judgement. Three findings are
settled and no further simulation will overturn them. Power for the affectedness effect is
governed by the verb count, so recruitment alone cannot rescue a language that is short of
target. English and Turkish do not reach the conventional target on both focal predictions
at any feasible per-language sample, whereas Norwegian does. Pooling the three languages
multiplies the verb-level information the estimate rests on and is the only remedy that
stays inside the agreed cap without changing the materials.

The choices that follow from this are set out in @tbl-choices. Each row is a decision for the
collaborators, with what the evidence says about it and what, if anything, is still
outstanding.

<!-- This table is tall enough to break across a page, which strands a single row
     under the caption and repeats the header. The break below keeps it whole. If the
     text above is substantially rewritten, check the rendered page and adjust. -->

`\newpage`{=latex}

| Choice | Decision | What settled it |
|:-----------------|:----------------------|:-----------------------------------------------|
| Adequacy rule | The interaction alone | No feasible sample reaches the conjunction of both focal predictions; dropping it lifts the English ceiling from `r pct0(ppv("English", "ceiling_joint_bf10"))` to `r pct0(ppv("English", "ceiling_h1b_bf10"))`. The passive-only slope is reported as a positive control |
| Evidence threshold | Bayes factor of 6 | Revised from the 10 first intended, which leaves the pooled test at `r pct0(pooled_combined("p_h1b")$p)` and no sample can lift. A directional Bayes factor of 3 is a one-sided tail probability of `r sprintf("%.2f", bf_onep(3))`, too lenient to carry a confirmatory claim |
| Sample size | 80 per language | Between 80 and 100 the simulations differ by less than their own resolution, and the verb count binds. 80 is the number the teams can collect |
| Confirmatory test | Pooled cross-linguistic | The theoretical claim is cross-linguistic, and the verb-level information is greatest there |
| Turkish | Estimation only | Its ceiling on the interaction is `r pct0(ppv("Turkish", "ceiling_h1b_bf10"))` at a Bayes factor of 10 whatever the sample. Settled; further cells will not change it |
| Verb pool | `r ppv("English", "n_verbs")` verbs, fixed | Closed by the project: the limit is the supply of suitable verbs, not session length, so the pool cannot be enlarged |
| Stopping rule | Fixed sample, announced in advance | Closed by the project: teams must take a stated sample size to their ethics committees. The statistical objection to sequential testing does not apply, but the operational one does |

: Design decisions, with what settled each. {#tbl-choices}

`\vspace{12pt}`{=latex}

None of these now waits on further computation. The project closed two of them, and the
simulations the rest; both are recorded as such. The rest follow from results already in
hand, and the simulations have said everything they can about them.

One caveat applies to the whole table. Every rate in this report is computed over all fitted
cells, whether or not they pass the convergence screen, which is deliberate, because a real
analysis would not discard a study for imperfect sampling. The proportion passing that
screen falls as the sample grows, so rates at different sample sizes are averages over
somewhat different mixtures of clean and untidy fits. On the screened subset the ordering of
adjacent sample sizes sometimes reverses. This does not affect the recommendation, which
rests on a criterion the design clears with room to spare and on a sample size chosen for
feasibility, but it is a reason to treat small differences between adjacent cells as noise.

## The Pooled Analysis: Rationale and Expectations

Following the recommendation above, the pooled cross-linguistic design analysis has been
implemented and validated. Each simulated study draws every language's effects from that
language's own pilot posterior, exactly as in the per-language assurance runs, and simulates
the three datasets. It then analyses them together with the cross-language model used in the
earlier synthesis work, which carries correlated random slopes by participant, by verb and
by language, with one amendment to the by-verb term set out below. The evidence criteria are
unchanged, namely directional Savage-Dickey Bayes factors on the pooled focal coefficients
under the same zero-centred priors. The recommendation is 80 participants per language, with
100 the most any team would be asked for, so the estimate combines the pooled cells at or
below 100. The grid itself runs from 50 to 150 per language, and the cells above 100 are
retained for record interest. No further simulation is queued or running. The decision arm,
at 70, 80 and 100 participants, carries 120 per cell, which brings the Monte-Carlo error
near a power of 80% down to about four percentage points. The pooled analytic ceiling given
below is computed from the fitted per-language models, not from a single pooled fit. The
pooled model has separately been fitted to the combined real pilot data, and what that fit
shows is set out in its own subsection below, because it qualifies the figures reported
here.

The simulated pooled cells, by contrast, are complete. `r pooled_interim`

The rationale is theoretical before it is statistical. The hypothesis CLAPS tests is
cross-linguistic, and the evidence for it has always been read across languages, most
explicitly in the meta-analytic synthesis of the five earlier studies
[@ambridgeArnonBekman2023]. The per-language analyses
treat each language as a self-contained confirmatory study, which is exactly why each
inherits its own verb-count ceiling. The pooled model instead tests the general claim
directly. Its population-level affectedness slope and interaction are the cross-linguistic
effects, and its by-language random slopes let each language depart from them, so
generality is tested rather than assumed.

Statistically, pooling eases the binding constraint identified above. The precision floor on
the affectedness effect is set by the verb-level information, and the pooled analysis draws
on about 210 verb-by-language units against 72, which shrinks the floor by roughly the
square root of three. The pilot slopes point the predicted way in all three languages, at
`r num2(ppv("English", "slope_per_sd"))`, `r num2(ppv("Turkish", "slope_per_sd"))` and
`r num2(ppv("Norwegian", "slope_per_sd"))` logits per one within-language standard deviation
of affectedness for the main effect and `r mnum2(ppv("English", "interaction_per_sd"))`,
`r mnum2(ppv("Turkish", "interaction_per_sd"))` and
`r mnum2(ppv("Norwegian", "interaction_per_sd"))` on the same scale for the interaction, so
their average, which leans on the more precisely estimated languages, sits comfortably clear
of zero. Partial pooling also shares strength in the other direction, since each language's
estimate borrows from the others in proportion to its own uncertainty, and Turkish gains
most from that. The fitted coefficients on the model's Gelman-scaled predictor are about
twice these per-standard-deviation values because that predictor has a standard deviation of
approximately 0.5.

The expected differences from the per-language results follow from this. First, the ceilings
should effectively disappear as constraints. With the pooled effects near the averages
above, and a precision floor of about `r num2(pooled_floor_per_sd)` logits per standard
deviation, both pooled focal ceilings sit near 100%. The pooled test should therefore be
limited by ordinary sampling noise alone. That expectation was formed before the pooled
cells returned, and the results reported above are what it should be read against: on the
criterion now recommended, the pooled interaction reaches
`r pooled_combined("p_h1b_bf6")$txt` and is flat in sample size, as a verb-limited design
predicts. Second, the limiting quantity changes. It is no longer the verb count but the
variation between languages, especially in the interaction, whose pilot values span
`r mnum2(min(pp$interaction_per_sd))` to `r mnum2(max(pp$interaction_per_sd))` per standard
deviation. That variation does not shrink with more verbs or more participants, and with
only three languages its variance is only weakly pinned down by the data and rests partly on
the prior [@gelman2006], so it is the number to watch in the results and in sensitivity
checks. Under the pilot values, it still leaves the pooled ceilings near
`r pct0(pooled_het_ceiling)` or higher. That figure is a reference, not an attainable
ceiling, and the distinction matters: it holds the between-language variance fixed at the
spread of the three pilot values, as though it were known. It is not known, and with three
languages it is barely identified, so the achievable rate is materially lower. Comparing the
simulated pooled result against this number therefore overstates any shortfall, and the
pooled arm's observed rate should be read against a benchmark that carries the uncertainty
in that variance. Third, the interpretation changes. A strong pooled Bayes factor alongside
a wide Turkish-specific interval is a coherent and likely outcome, and it is the same
pattern the cross-linguistic literature shows. The per-language conclusions above stand
unchanged. Norwegian is confirmable on its own, and the English and Turkish per-language
joint tests remain capped whatever the pooled result.

The pooled cells have since returned, and they do sit below these expectations at the
strongest threshold. The likely cause lies in the between-language variance component, not
in the focal effects, and the pooled test's sensitivity to that component's prior will be
examined before any recommendation is drawn from it.



### The pooled model on real pilot data

The pooled analysis model was fitted once to the combined real pilot data of all three
languages, 75,528 ratings over 210 verbs, as a check that it behaves on real ratings before
thousands of simulated fits assume it does. The result is reported here in full, because it
bears on the figures above in both directions.

The effects come out as expected. The focal interaction is estimated at $-0.48$, with a 95%
credible interval from $-1.04$ to $+0.31$ and a posterior probability of 0.91 that it carries
the predicted sign, which is a directional Bayes factor of about 10. The unweighted average of
the three per-language estimates that the simulation draws from, $-0.57$, falls comfortably
inside that interval, so the data-generating process the design analysis uses is consistent
with what the pooled model recovers from real ratings.

The sampling was another matter. The fit did not meet the convergence criteria, saturating
the maximum treedepth on 5,968 of its 8,000 post-warm-up iterations, alongside one divergent
transition, an R-hat of 1.008 and a minimum bulk effective sample size of 453. Saturation on
that scale is not a marginal miss. It means the sampler could not traverse the posterior
within its step budget, which is the geometry problem the by-verb amendment below is meant
to remove, appearing here on real data and more severely than in simulation.

More consequential is the fit's implication for the variation between languages, the
quantity that limits the pooled test. Its standard deviation for the focal interaction is
estimated at 0.59, with an interval from 0.07 to 1.65, where the design analysis leans on
the spread of the three per-language point estimates, 0.30, about half as large. Were the
larger value nearer the truth, the pooled test would be less powerful than the figures above
report, since the precision of the cross-language average scales with it. Three things pull
the other way. The interval is very wide and easily contains the smaller value. The
simulation's own procedure induces more between-language spread than the raw 0.30, because
each replicate draws each language's effects from its own posterior. And the estimate comes
from the poorly sampled fit just described, so it is a flag rather than a correction.

A refit under the amended specification started on 20 August 2026 and should return in about
eight days, the time the original fit took. The amended model carries fewer parameters and
should sample more easily. That refit will settle both questions, and until it lands the
pooled figures above should be read as the design analysis's best estimate under a
data-generating process the real data supports, and not as a guarantee. None of it disturbs
the sample size, which follows from the verb count and the number of languages.

### The by-verb term, and why it is amended

The confirmatory model drops the by-verb random slope on affectedness, keeping
`(1 + S_Type | Verb)`, which is the term the single-language models already use. The reason is
that affectedness barely varies within a verb. Each verb carries two values, one for each agent
gender, and their spread is about 0.6% of the spread between verbs, an SD of 0.18 against 30.9.
A random slope estimated from that much variation is close to a second intercept, and the
sampler shows it.

The evidence is in the runs already completed, and it is one-sided. Across the 135 completed
pooled fits, which carry the by-verb slope, 59% produced at least one divergent transition
and 47% saturated the treedepth. Across 360 completed single-language fits at 80
participants, which use the term recommended here, both figures are zero. (These come from
the per-cell diagnostic records, not from the summary tables this report reads, so the text
quotes them without deriving them.) Divergences and treedepth saturation are signatures of
awkward posterior geometry rather than of a sampler given too little time, so they do not go
away with longer chains, and divergent transitions can bias the estimates they contribute
to. The amendment removes that geometry at the cost of a variance component the data cannot
inform.

Two qualifications. The operating characteristics reported above were computed under the
unamended model, so they describe a specification marginally different from the one
recommended. The dropped component is estimated from almost no information and the focal
interaction's precision is dominated by the by-language term, so the figures are expected to
carry over with little change, but that expectation has not been simulated and should not be
presented as though it had. Separately, the amendment addresses geometry, leaving efficiency
untouched, and the two failure modes call for different remedies. Most non-convergence in
both model families is a failure to reach the effective-sample-size and R-hat thresholds,
which affects the single-language models as much as the pooled one and follows from fitting
8,000 draws two thousand times over. The confirmatory analysis is fitted once and takes the
heavier sampler the proposal sets for confirmatory fits, the sixteen chains of 5,000
iterations given above, six times the post-warm-up draws of a replicate and enough to clear
those thresholds. Treedepth saturation is the other mode, and a longer run does nothing for
it, because the binding constraint is the depth of each trajectory, not how many there are.
The amendment exists to remove it, and the refit running since 20 August is the test of
whether it succeeds. That is the one question here still open. The proposal leaves no
discretion if the answer is no: zero divergences and zero treedepth saturations are locked
acceptance criteria for confirmatory fits, and a fit that misses them falls to the next
level of the model ladder, its result set aside. Raising the maximum tree depth is not an
available remedy, the configuration reserving it for diagnosis.

# Limitations and Status

The status differs across the two analyses. The literature-anchored cross-check is complete
for all three languages and points to a modest sample at the full verb set. The
data-grounded analysis, which is the primary basis, is substantially complete. Norwegian is
finished and is adequately powered at about 100 participants. English and Turkish fall short
of the joint power target at every sample size run. The finding that participants alone
cannot bring English and Turkish to target rests on the affectedness effect being
verb-limited. The 400-participant cells confirm that for English at
`r jval("English", 400, "reps")` replicates, and the analytic ceilings bound it from above.
No further simulation remains to run: recruitment is capped at 100 participants, and every
sample at or below that cap has been simulated to full replication. The one fit still
outstanding is the refit described above.

That status statement concerns the joint criterion, which this report no longer recommends.
Gated on the interaction alone, English and Norwegian both reach target within the cap; it is
Turkish that remains short, and permanently so.

Two aspects of the simulation record qualify how far these figures can be pressed. Each
assurance replicate reuses one of the first 24 to 40 saved posterior draws instead of draws
spread evenly across the whole posterior. Checked against the full posterior, those blocks
are representative for English and Norwegian and sit slightly below the mean for the Turkish
affectedness slope, so the simulated Turkish figures are, if anything, mildly pessimistic.
The analytic ceilings use all 8,000 draws and are unaffected, and the pooled
cross-linguistic design analysis and the decision arm sample their draw indices randomly
over the full pilot posteriors, so the limitation does not reach them. Replication itself is
complete, since the original single-language grid finished all 720 of its cells and the low
counts noticed mid-run were cells still computing.

The analysis rests on `r format(n_assurance + n_safeguard + sum(pooledp$reps) + n_lit, big.mark = ",")`
simulated studies in the cells reported here, and they are not evenly spread. The
per-language assurance grid holds `r format(n_assurance, big.mark = ",")` and the
literature-anchored cross-check `r format(n_lit, big.mark = ",")`, against
`r sum(pooledp$reps)` in the pooled arm, which is
`r sprintf("%.0f%%", 100 * sum(pooledp$reps) / (n_assurance + n_safeguard + sum(pooledp$reps) + n_lit))`
of the total. The recommendation rests on the smallest of them, and that has a consequence.
Combining the pooled cells at or below the cap gives
`r pooled_combined("p_h1b_bf6")$txt` for the interaction at a Bayes factor of 6, and the
lower end of that interval falls below 80%. The single-language figures are established at
their targets and the pooled one is not: it is the best estimate available, not a
demonstration that the confirmatory test reaches a particular power.

That estimate carries its own uncertainty, and the scale of it can be stated. Taking the
pooled cells at face value under a flat prior, the true rate is most likely near
`r pct0(pooled_post()$k / pooled_post()$n)`, with a probability of about
`r num2(pp_atleast(0.80))` that it is at least 80%, about `r num2(pp_atleast(0.90))` that it
reaches 90%, and about `r num2(pp_below(0.75))` that it is below 75%. A run of
`r N_NEW_SUBMITTED` further cells at 80 participants per language was prepared to tighten
this, and deliberately not pursued. Its arithmetic is the reason. It would have been
unlikely to disturb the recommendation, the chance of landing more than ten points from the
current figure being about `r num2(pooled_predictive$p_beyond_10)`, and it would have fallen
within five points roughly `r pct0(pooled_predictive$p_within_5)` of the time. It was also
unlikely to settle the question: because the rate sits so close to the target, the chance
that the combined interval would then clear 80% is about
`r num2(pooled_predictive$p_settles)`. Several weeks of computing for an even chance of
firming up a number that changes no decision is not a good trade.

The more substantial reason is that the pooled arm does not appear to have much room to
improve. Its rate should be read against a benchmark that treats the between-language
variance as estimated from three languages rather than known, and on that footing the
observed rate is close to what three languages permit. The constraint is the number of
languages, which the project cannot change, not a defect in the model that a better
specification would repair. Further simulation would therefore measure this design more
precisely without producing a better one, and the design analysis is closed on that basis.

Each data-grounded condition rests on `r min(jp_a$reps)` to `r max(jp_a$reps)` simulated
studies, and each literature-anchored cell on 50, so the figures carry a Monte-Carlo error
of several percentage points near a power of 80%. Individual cells close to the threshold
should be read as near target, and the conclusions rest on the pattern across sample sizes.
On the literature effects, the adequate sample at 72 verbs is reported as at most 50, since
the main sweep does not go below 50. It is not pinned down further, as the recommendation
rests on the data-grounded analysis.

This is an a-priori power analysis, a simulation under assumed effects to size the study,
and it is distinct from the confirmatory analysis of the real ratings that will be run once
data collection is complete. The analysis code and configuration are openly available, and
the report rebuilds from the committed summary tables without the raw data, which remain
private to CLAPS until the project reaches a more advanced stage. The simulation programme
is now closed, recruitment having been capped at 100 participants, so this report is
finalised on the evidence described, with no further cells pending.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix A}\\[3pt]
{\bfseries Composition of the Pilot Sample}
\end{center}
```

This appendix describes the pilot sample from which the affectedness ratings and the
descriptive trends in @fig-pilot are drawn, and against which the data-grounded analysis
fits its models. The counts are read directly from the harmonised pilot data.

```{r tbl-pilot}
#| tbl-cap: "Composition of the Harmonised Pilot Sample, by Language"

comp <- pilot_composition
kable(comp, align = "lrrrrr", format.args = list(big.mark = ","))
```

`\vspace{-8pt}`{=latex}

*Note.* Counts are read from the harmonised pilot data. Items are verb-by-agent-gender
combinations, so the item count is twice the verb count. Norwegian lacks a pseudo-passive,
hence two sentence types rather than three.

`\vspace{12pt}`{=latex}

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix B}\\[3pt]
{\bfseries Key Code for the Simulation and the Models}
\end{center}
```

This appendix reproduces the central code, lightly abridged, from the analysis scripts. The
code shown is the data-generating process for the literature-anchored cross-check, in
`R/06_simulate_design.R`, together with the analysis model in `R/04_model_formulas.R` and
the Bayes factor in `R/05_hypothesis_tests.R`. The data-grounded analysis uses the same
analysis model and Bayes factor, but draws its data-generating effects, random-effect
covariance and thresholds from the fitted pilot models, and reads each verb's affectedness
from the pilot data. Its assurance and safeguard variants are in
`R/10_simulate_from_pilot.R` and `R/10_simulate_from_pilot_v2.R`.

Each simulated dataset is drawn from the same cumulative-logit model that is later fitted.
In the cross-check, affectedness is drawn once per verb across a plausible range, which makes
it a verb-level predictor, and the ordinal response follows from the cumulative-logit linear
predictor.

```r
# DGP for the literature-anchored cross-check, abridged from R/06_simulate_design.R.
verb_semantics <- runif(n_verbs, min = -0.5, max = 0.5)   # one value per verb

# For participant pid, verb vid and sentence type st:
sem   <- verb_semantics[[vid]]
b_int <- if (st == "Active")              beta_active_interaction * sem
         else if (st == "Pseudo_Passive") beta_pseudo_interaction * sem
         else 0
eta   <- part_intercepts[pid] + part_slopes[pid] * sem +
         verb_intercepts[vid] + beta_semantics * sem + b_int

cum_probs <- plogis(thresholds - eta)        # seven ordered categories
probs     <- diff(c(0, cum_probs, 1))
resp      <- sample(seq_len(7), size = 1, prob = probs)
```

The analysis model is the maximal random-effects structure consistent with the design,
fitted with the cumulative-logit family.

```r
# Analysis model, from R/04_model_formulas.R.
brms::bf(
  Response ~ S_Type * Semantics_scaled +
    (1 + S_Type * Semantics_scaled | Participant) +
    (1 + S_Type | Verb),
  family = brms::cumulative(link = "logit", threshold = "flexible")
)
```

Evidence for each prediction is a directional Savage-Dickey Bayes factor, with the prior
sampled alongside the posterior and the posterior mass taken in the predicted direction.

```r
# Directional Savage-Dickey Bayes factor, from R/05_hypothesis_tests.R.
# post_vals and prior_vals are draws of the focal coefficient; the negative
# direction is shown, as for the primary prediction H1b.
post_prob  <- mean(post_vals  < 0)
prior_prob <- mean(prior_vals < 0)
bf_10      <- (post_prob / (1 - post_prob)) / (prior_prob / (1 - prior_prob))
```

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix C}\\[3pt]
{\bfseries Full Bayes-Factor Operating Characteristics}
\end{center}
```

This appendix reports the operating characteristics for all four predictions in every available
cell of the corrected literature-anchored run with at least ten replicates, namely the
probability of exceeding the primary and secondary Bayes-factor thresholds, the median Bayes
factor and the convergence rate, set out in @tbl-opchar, which covers `r lit_language_text`.

```{r tbl-opchar}
#| tbl-cap: "Operating Characteristics by Design Condition and Prediction"

hyp_lab <- c(H1a_semantics_positive = "H1a", H1b_active_interaction_negative = "H1b",
             H2a_pseudo_positive = "H2a", H2b_pseudo_negative = "H2b")
opc <- exc |>
  filter(reps >= 10) |>
  mutate(Hyp = factor(hyp_lab[hypothesis], levels = c("H1a", "H1b", "H2a", "H2b"))) |>
  transmute(Language = language, Verbs = n_verbs, N = n_participants, Hyp,
            `P(BF>=10)` = sprintf("%.0f%%", 100 * p_bf_primary),
            `P(BF>=3)`  = sprintf("%.0f%%", 100 * p_bf_secondary),
            # Formatted per value rather than left numeric: a numeric column here
            # prints as 2.8e+01 and 4.1e-02, which is not how anyone reads a Bayes factor.
            `Median BF` = vapply(median_bf, function(x)
              format(signif(x, 2), scientific = FALSE, trim = TRUE, big.mark = ","),
              character(1)),
            Conv        = sprintf("%.0f%%", 100 * p_convergence_ok),
            Reps        = reps) |>
  arrange(Language, Verbs, N, Hyp)
kable(opc, booktabs = TRUE, longtable = TRUE, align = "lrrlrrrrr",
      col.names = c("Language", "Verbs", "N", "Prediction", "P(BF>=10)",
                    "P(BF>=3)", "Median BF", "Convergence", "Reps")) |>
  kableExtra::kable_styling(font_size = 10, latex_options = c("repeat_header"))
```

`\vspace{-18pt}`{=latex}

*Note.* The probability columns give the proportion of simulated studies reaching a Bayes
factor of 10 and of 3, Median BF the median across them to two significant figures,
Convergence the proportion of fits meeting the criteria, and Reps the number of studies in
the cell. Cells holding fewer than ten replicates are excluded, and Norwegian has no
pseudo-passive, hence half as many rows.

`\vspace{12pt}`{=latex}

{{< pagebreak >}}

```{=latex}
\begin{center}
{\large\bfseries Appendix D}\\[3pt]
{\bfseries Guide to the Analysis Code}
\end{center}
```

This appendix maps the analysis code, which is openly available at
<https://github.com/pablobernabeu/CLAPS-analysis>. It is meant to make the repository
navigable for a reader who wants to check how a number in this report was produced, or to
reuse part of the machinery. The raw pilot data are not published, since they are
participant-level records held privately until the project is further advanced, so the
repository ships the simulation and analysis code together with the aggregated summaries
the report reads.

Three entry points cover most purposes. A reader checking a reported number wants
`outputs/`, which holds the aggregated summaries and is where every figure in this report
comes from. A reader checking how a simulated study is built and analysed wants
[`R/06_simulate_design.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design.R)
for the literature-anchored generator,
[`R/10_simulate_from_pilot_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot_v2.R)
for the pilot-grounded one, and
[`R/05_hypothesis_tests.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/05_hypothesis_tests.R)
for the Bayes factor. A reader intending to rerun the analysis wants `scripts/`, whose files
are numbered in the order they are meant to run, together with the matching submitters in
`hpc/`. @tbl-repo sets out what each top-level directory holds.

`\vspace{10pt}`{=latex}

| Location | What it holds |
|:------------------|:-----------------------------------------------------------|
| `R/` | The building blocks: data validation and preprocessing (`01`, `02`), priors and model formulas (`03`, `04`), the Bayes factor (`05`), the two simulation engines (`06`, and `10` with `11`), convergence diagnostics (`07`) and summarising (`08`) |
| `scripts/` | Runnable steps, numbered in running order, from harmonising the pilot (`00`) through fitting the pilot models, generating a design grid, running one cell, to aggregating results (`aggregate_pilot.R`, `aggregate_pooled.R`, `aggregate_literature.R`) |
| `hpc/` | One SLURM submitter per arm. Each takes a grid file and an output directory, and runs one grid row per array task |
| `config/` | The design grids, one row per simulated condition, plus the model and prior configuration |
| `outputs/` | The aggregated summaries this report reads, and the figure it displays |
| `tests/` | Unit tests for the validation, prior, diagnostic and Bayes-factor logic |
| `reports/` | This report, its bibliography and citation style |

: Layout of the public analysis repository. {#tbl-repo}

`\vspace{12pt}`{=latex}

The unit of work is the cell, meaning one simulated condition: a language, a participant
count, a verb count and a replicate index. A grid file in `config/` lists the cells, an
array task runs one row, and each cell is saved as a single `.rds` file holding its
summary, its Bayes factors and its convergence diagnostics. The aggregation scripts read a
directory of those files and reduce them to the tables in `outputs/`. Reading one saved
cell is therefore the quickest way to see exactly what the pipeline computes per simulated
study.

Two conventions are worth knowing before reading the code. Names carrying `pilot` refer to
the data-grounded analysis that conditions on the fitted pilot models and forms the primary
basis of this report, while names carrying `literature` refer to the cross-check that
conditions on the published effect sizes. Separately, the suffix `_v2` on a simulation
engine marks an extension of the original, never a replacement for it, so both files are
live and the pair should be read together.

Three files in `R/` generate simulated studies, `06`, `10` and `11`, and between them they
implement only two engines. The engines differ in where the true effects come from, and the
second of them is written at two scopes.

The literature-anchored engine is
[`R/06_simulate_design.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design.R).
It invents a study from assumed effect sizes taken from the published estimates, drawing
participants, verbs and affectedness values from scratch. Because its parameters are
stipulated, not estimated, extending it from one language to several is only a matter of
adding a language level to the same recipe, which is why this engine covers both scopes in
one file: `simulate_claps_data()` for a single language and
`simulate_claps_data_multilanguage()` beside it.

The pilot-grounded engine is
[`R/10_simulate_from_pilot.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot.R).
It takes the true effects from a model already fitted to the pilot data, so a simulated
study inherits that language's estimated slopes, thresholds and variance components together
with its real verbs and their measured affectedness.
[`R/10_simulate_from_pilot_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/10_simulate_from_pilot_v2.R)
loads that file and adds the assurance and safeguard variants, which sample the effects
afresh from the pilot posterior at each replicate, where the base version holds them at its
mean.

Extending this second engine across languages is not a matter of adding a level, because
each language carries its own fitted model. A pooled study has to be assembled from three
separate sets of parameters, simulated language by language and then combined before a
single cross-language model is fitted to the result. That orchestration is added by
[`R/11_simulate_pooled_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/11_simulate_pooled_v2.R)
adds. It loads the `_v2` file, calls its per-language simulator once for each language,
binds the three datasets and fits the pooled model, so it is the same engine at a wider
scope.

The files therefore stack rather than compete, each loading the one below it and adding to it:
`11` on `10_v2` on `10`, and separately
[`R/06_simulate_design_gelman.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/06_simulate_design_gelman.R)
on `06`. That last file is the narrowest of them. It changes only the scale on which the
analysed predictor is expressed, leaving the simulated responses identical, so that the
literature-anchored cells are analysed on the same footing as the confirmatory model.

Both engines end in the same place. Each defines a cell runner, `run_design_cell()` in `06`
and `run_databased_cell_v2()` and `run_pooled_cell_v2()` in `10_v2` and `11`, which
simulates one dataset, fits it with
[`R/04_model_formulas.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/04_model_formulas.R)
and the priors in
[`R/03_define_priors.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/03_define_priors.R),
computes the Bayes factors with
[`R/05_hypothesis_tests.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/05_hypothesis_tests.R)
and records the diagnostics from
[`R/07_extract_diagnostics.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/R/07_extract_diagnostics.R),
then writes the `.rds` for that cell. Only the source of the true effects differs; the
model, the evidence criterion and the convergence standard are shared, which makes the two
analyses comparable. @tbl-engines summarises where each engine's effects come from and which
file implements it at each scope.

`\vspace{10pt}`{=latex}

| Engine | Where the effects come from | Single language | All three pooled |
|:----------------------|:--------------------------------|:---------------------|:---------------------|
| Literature-anchored | Published effect sizes, stipulated | `06` | `06`, the same file |
| Pilot-grounded | Estimated from the fitted pilot models | `10`, extended by `10_v2` | `11`, which loads `10_v2` |

: The two simulation engines and the files implementing them. {#tbl-engines}

`\vspace{12pt}`{=latex}

Following one arm end to end is the quickest way to see how the pieces connect. For the
pooled analysis reported here, the grid is `config/design_grid_pooled_v2.csv`, one row per
simulated study;
[`hpc/submit_pooled_v2_array.sh`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/hpc/submit_pooled_v2_array.sh)
submits one array task per row;
[`scripts/07_pooled_cell_v2.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/scripts/07_pooled_cell_v2.R)
runs a single row by calling `run_pooled_cell_v2()`; and
[`scripts/aggregate_pooled.R`](https://github.com/pablobernabeu/CLAPS-analysis/blob/main/scripts/aggregate_pooled.R)
reduces the resulting `.rds` files to `pooled_power_pilot.csv` in
`outputs/design_summary_pilot/`, which is the table this report reads for every pooled
figure it quotes.
