---
title: "Preliminary Sample-Size Analysis for CLAPS"
author: "Pablo Bernabeu and CLAPS collaborators"
date: today
date-format: "D MMMM YYYY"
lang: en-GB
format:
  pdf:
    geometry: "margin=1in"
    linestretch: 1.25
    fig-cap-location: top
    fig-pos: "htbp"
    colorlinks: true
    code-overflow: wrap
    include-in-header:
      text: |
        \definecolor{subtitlegray}{gray}{0.35}
        \setkomafont{date}{\normalfont\normalsize\color{subtitlegray}}
        % Microtypographic character protrusion and font expansion for an
        % evener, less rivered right margin under justification.
        \usepackage{microtype}
        % Discourage widows and orphans (a lone line stranded at the top or
        % bottom of a page), which a technical report with many short
        % paragraphs is otherwise prone to.
        \widowpenalty=10000
        \clubpenalty=10000
        % Allow a little more stretch between words before over-full lines are
        % risked, so justification does not force awkward hyphenation.
        \setlength{\emergencystretch}{3em}
        % Wrap long lines inside code blocks at the margin (breakanywhere lets a
        % long path or token break even without a space) rather than letting
        % them run past the right edge.
        \usepackage{fvextra}
        \fvset{breaklines=true,breakanywhere=true}
        % Show the "Abstract" label, which KOMA suppresses by default. The
        % report follows APA 7 elsewhere, where the abstract is labelled.
        \KOMAoptions{abstract=on}
        % APA 7 caption style: bold label, italic title, applied to figures and tables.
        \usepackage{caption}
        \DeclareCaptionFormat{apa7}{#1\\[5pt]#3}
        \DeclareCaptionFormat{apa7par}{#1\par\vspace{5pt}#3}
        \captionsetup{format=apa7, labelsep=none, labelfont=bf, textfont=it,
          justification=raggedright, singlelinecheck=false, skip=6pt}
        \captionsetup[longtable]{format=apa7par, labelsep=none, labelfont=bf,
          textfont=it, justification=raggedright, singlelinecheck=false, skip=6pt}
        % Top-align figures that land on their own float page (avoids vertical centring).
        \makeatletter
        \setlength{\@fptop}{0pt}
        \setlength{\@fpsep}{12pt plus 1fil}
        \setlength{\@fpbot}{0pt plus 1fil}
        \makeatother
bibliography:
  - ../references.bib
csl: apa.csl
link-citations: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(kableExtra)
library(ggplot2)

# One language order for every table and figure, the order the pilot tables use. Naming
# it keeps any display from falling back on alphabetical sorting. union() keeps any
# language outside the list instead of turning it into NA.
LANG_LEVELS <- c("English", "Turkish", "Norwegian")
lang_factor <- function(x) factor(x, levels = union(LANG_LEVELS, sort(unique(x))))

# The public report is deliberately independent of participant-level pilot data.
# It uses only disclosure-reviewed derivatives that are already represented in
# the published PDF: a raster figure and aggregate sample-composition counts.
pilot_composition <- read_csv(
  "../outputs/design_summary_pilot/pilot_sample_composition.csv",
  show_col_types = FALSE
)
pilot_figure <- "../outputs/figures/pilot_acceptability_by_gender_report.png"

# Pilot-grounded (v2) summaries, aggregated on the cluster by
# scripts/aggregate_pilot.R: per-cell detection rates, the fitted pilot parameters
# and the analytic detection ceilings. Reading them here keeps every simulation
# result in the report programmatic.
jp <- read_csv("../outputs/design_summary_pilot/joint_power_pilot.csv",
               show_col_types = FALSE)
pp <- read_csv("../outputs/design_summary_pilot/pilot_params_ceilings.csv",
               show_col_types = FALSE) |>
  mutate(language = lang_factor(language)) |>
  arrange(language)

pct0  <- function(x) sprintf("%.0f%%", 100 * x)
num2  <- function(x) sprintf("%.2f", x)
num3  <- function(x) sprintf("%.3f", x)
mnum2 <- function(x) sprintf("$%.2f$", x)    # math mode, for a proper minus sign
snum2 <- function(x) sprintf("$%+.2f$", x)   # signed, in math mode
snum3 <- function(x) sprintf("$%+.3f$", x)
# Probabilities without the leading zero, as APA 7 asks for quantities that cannot
# exceed one.
prob2 <- function(x) sub("^0", "", sprintf("%.2f", x))
ppv   <- function(lang, col) pp[[col]][pp$language == lang]
and_list <- function(x) {
  if (length(x) > 1) paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1)) else
    as.character(x)
}

# Grid cells are cited only when they hold a minimum number of replicates.
MIN_REPS_V2 <- 10L
jp_a <- filter(jp, mode == "assurance", reps >= MIN_REPS_V2)
jval <- function(lang, N, col) jp_a[[col]][jp_a$language == lang & jp_a$n_participants == N]

# The decision arm: the simulated studies at 70, 80 and 100 participants per language on
# which the per-language results rest, each scored for every prediction and summarised by
# scripts/aggregate_pseudo_passive.R. Every per-language rate in the text and tables is
# read from it, so the rates at those samples all rest on the same replicates. The file
# stores integer counts, and a language without a pseudo-passive carries NA in every H2
# column, so it cannot print as a test that failed.
ppd <- read_csv("../outputs/design_summary_pilot/pseudo_passive_decision_arm.csv",
                show_col_types = FALSE)
ppp <- read_csv("../outputs/design_summary_pilot/pseudo_passive_pilot_posterior.csv",
                show_col_types = FALSE)
# The grid summary above counts the same replicates wherever its cell holds exactly the
# decision arm, as at 80 participants. This stops the render if a re-aggregation of
# either file leaves the two out of step.
ppd_same <- ppd |>
  inner_join(select(jp_a, language, n_participants, reps, p_h1b),
             by = c("language", "n_participants"), suffix = c("", "_jp")) |>
  filter(reps == reps_jp)
stopifnot(setequal(ppd_same$language, LANG_LEVELS),
          all(ppd_same$n_h1b == round(ppd_same$p_h1b * ppd_same$reps)))

da_reps <- unique(ppd$reps)
da_ns   <- sort(unique(ppd$n_participants))
stopifnot(length(da_reps) == 1)
da      <- function(lang, N, col) ppd[[col]][ppd$language == lang & ppd$n_participants == N]
da_rate <- function(lang, N, col = "n_h1b") da(lang, N, col) / da(lang, N, "reps")
# Rates come with an exact interval, since each is a proportion over a limited number of
# replicates and none is known to the precision a bare percentage implies. Intervals are
# printed in percentage points, the unit of the rates they belong to.
exact_ci <- function(k, n) stats::binom.test(k, n)$conf.int
pct_ci   <- function(k, n) {
  b <- exact_ci(k, n)
  sprintf("[%.0f, %.0f]", 100 * b[1], 100 * b[2])
}
da_ci <- function(lang, N, col = "n_h1b") pct_ci(da(lang, N, col), da(lang, N, "reps"))

# Pilot posterior of the pseudo-passive interaction, per SD of affectedness.
h2post  <- function(lang, col) ppp[[col]][ppp$language == lang]
# Every posterior draw on one side of zero is said in words, and anything short of that
# is given as a count, so the sentence stays true if the pilot models are refitted.
h2draws <- function(lang) {
  k <- h2post(lang, "n_negative")
  n <- h2post(lang, "ndraws")
  if (k == n) paste("all", format(n, big.mark = ",")) else
    paste(format(k, big.mark = ","), "of", format(n, big.mark = ","))
}

# A directional Bayes factor of k is met exactly when the posterior probability of the
# predicted sign reaches k/(k+1).
bf_sign <- function(k) k / (k + 1)

# The limit the verbs set on the interaction's detection rate, from an approximate formula
# applied to the fitted pilot models. It is quoted in whole percentages because the formula
# ignores part of the covariance among the by-verb effects.
ceil_txt <- function(lang) {
  x <- ppv(lang, "ceiling_h1b_bf10")
  if (x >= 0.99) "over 99%" else paste("about", pct0(x))
}

# The three-language pooled arm, aggregated by scripts/aggregate_pooled.R. Its cells are
# combined at or below 100 participants per language, since the rate shows no trend over
# that range and no single cell holds more than a few dozen replicates.
pooledp <- read_csv("../outputs/design_summary_pilot/pooled_power_pilot.csv",
                    show_col_types = FALSE)
POOLED_MAX_N <- 100
pooled_combined <- function(col = "p_h1b", max_n = POOLED_MAX_N) {
  d <- pooledp[pooledp$n_participants <= max_n, ]
  k <- sum(round(d[[col]] * d$reps))
  n <- sum(d$reps)
  list(p = k / n, k = k, n = n, ns = sort(unique(d$n_participants)),
       txt = sprintf("%s %s", pct0(k / n), pct_ci(k, n)))
}

# The repository's wider grid (joint_power_pilot.csv) holds more studies than the decision
# arm at some samples, scored for H1b alone. The excess is stated once, when it is the same at
# every such cell, so a reader who opens the repository can match the two sets of figures.
grid_extra <- jp_a |>
  inner_join(select(ppd, language, n_participants, reps_da = reps),
             by = c("language", "n_participants")) |>
  mutate(extra = reps - reps_da) |>
  filter(extra > 0)
grid_extra_txt <- if (nrow(grid_extra) > 0 && length(unique(grid_extra$extra)) == 1) {
  sprintf(paste("The repository's wider grid adds %d earlier studies at %s participants,",
                "scored for H1b alone."),
          unique(grid_extra$extra), and_list(sort(unique(grid_extra$n_participants))))
} else ""

# The pooled analysis with more languages, from scripts/run_klanguage_design_analysis.R
# on the engine in R/13_simulate_klanguage_pooled.R. In klanguage_known7.csv the seven
# languages measured so far take part with effects drawn from their estimates, and the rest
# are drawn
# from a population whose between-language spread is that of one of three sets of
# languages.
K_DIR <- "../outputs/design_summary_pilot/"
k7    <- read_csv(paste0(K_DIR, "klanguage_known7.csv"), show_col_types = FALSE)
kpopr <- read_csv(paste0(K_DIR, "klanguage_populations_by_reading.csv"),
                  show_col_types = FALSE)
K_POPS <- c("claps_protocol", "seven", "glossa_protocol")
# How much doubt about protocol transfer each regime is paired with. The driver script
# keeps every pairing and leaves the choice to the reporting. A regime in which languages
# measured under the CLAPS protocol spread by only about 0.155 attributes most of the
# meta-analysis languages' spread of about 0.53 to their protocol, which is the premise
# that their estimates transfer poorly, so that regime takes the 0.40 doubt. The other two
# do not attribute the spread to protocol and take the 0.15 doubt.
K7_SET <- c(claps_protocol  = "seven, drawn, 2023 widened 0.40",
            seven           = "seven, drawn, 2023 widened 0.15",
            glossa_protocol = "seven, drawn, 2023 widened 0.15")
kk <- function(pop, K, variant = "canonical", set = K7_SET[[pop]]) {
  x <- k7$power[k7$reading_variant == variant & k7$population == pop & k7$K == K &
                k7$known_set == set & k7$mode == "assurance"]
  stopifnot(length(x) == 1)
  x
}
kpop <- function(pop, col, variant = "canonical") {
  x <- kpopr[[col]][kpopr$reading_variant == variant & kpopr$population == pop]
  stopifnot(length(x) == 1)
  x
}
k_reps <- unique(k7$n_reps)
stopifnot(length(k_reps) == 1)
# Two decimals, rounding half up, since 0.155 is stored just below its decimal value.
num2h <- function(x) sprintf("%.2f", x + 1e-9)

# The same synthesis for the pseudo-passive interaction, from
# scripts/run_klanguage_pseudo_passive.R. The measured languages with a pseudo-passive
# take part at effects drawn from their estimates, and new languages have one at the rate
# among the measured languages. Its three sets of languages stand in for the three above.
kpp    <- read_csv(paste0(K_DIR, "klanguage_pseudo_passive.csv"), show_col_types = FALSE)
kpppop <- read_csv(paste0(K_DIR, "klanguage_pseudo_passive_populations.csv"),
                   show_col_types = FALSE)
KPP_POPS <- c(claps_protocol = "claps_protocol", seven = "five",
              glossa_protocol = "glossa_protocol")
KPP_SET  <- "five, drawn, Glossa widened 0.15"
n_measured    <- max(k7$n_known)
n_measured_pp <- kpppop$K[kpppop$population == "five"]
# Languages with a pseudo-passive among K, at the measured rate or with every new one.
kpp_n     <- function(K) n_measured_pp + as.integer(round((K - n_measured) * n_measured_pp / n_measured))
kpp_n_all <- function(K) K - (n_measured - n_measured_pp)
kpp_rate <- function(pop, K, n_fun = kpp_n) {
  x <- kpp$power[kpp$population == KPP_POPS[[pop]] & kpp$K_pp == n_fun(K) &
                 kpp$known_set == KPP_SET & kpp$mode == "assurance" &
                 kpp$prior_tau_scale == 1.1]
  stopifnot(length(x) == 1)
  x
}
kpp_tau <- function(pop) kpppop$tau_median[kpppop$population == KPP_POPS[[pop]]]
kpp_k   <- function(pop) kpppop$K[kpppop$population == KPP_POPS[[pop]]]
# Every row takes the 0.15 doubt, which is coherent only while the pilots spread at least
# as much as the meta-analysis languages (see the pairing note above).
stopifnot(kpp_tau("claps_protocol") >= kpp_tau("glossa_protocol"))
# How far giving every new language a pseudo-passive moves a rate, rounded up because the
# sentence states a maximum.
kpp_rule_gap_pts <- ceiling(round(100 * max(abs(unlist(lapply(K_POPS, function(p)
  vapply(c(10, 20), function(K) kpp_rate(p, K, kpp_n_all) - kpp_rate(p, K), numeric(1)))))), 6))
# Counts below ten in words, as APA 7 asks.
num_word <- function(n) c("one", "two", "three", "four", "five", "six", "seven", "eight",
                          "nine")[n]
# Percentages for a table typeset without escaping, where a bare % starts a comment.
pct0_tex <- function(x) sub("%", "\\\\%", pct0(x))

# The agent-gender versions, taken up in Appendix B. The gender terms come from the pilot
# re-analyses in scripts/assess_gender_versions.R, a linear mixed model of the 1-to-7
# ratings with agent gender coded plus or minus a half, so they are quoted beside that
# model's own estimate of the interaction.
gve <- read_csv("../outputs/design_summary_pilot/gender_versions_gender_effects.csv",
                show_col_types = FALSE)
gvev <- function(lang, term, col = "estimate") {
  gve[[col]][gve$language == lang & gve$term == term]
}
# The gender script scales affectedness by twice its verb-level standard deviation, so a
# coefficient per standard deviation of affectedness is exactly half the fitted one.
gv_per_sd <- function(x) x / 2
GV_TERM_H1B <- "S_TypeActive:Semantics_scaled:Gender_c"
# What one agent gender for every item would do to the interaction: with the code at
# plus or minus a half, a design whose participants all see one gender shifts the slope by
# half the gender term, reported as a fraction of the slope itself.
gv_shift <- function(lang, g = 0.5) {
  abs(g * gvev(lang, GV_TERM_H1B) / gvev(lang, GV_TERM_H1B, "h1b_estimate"))
}
# Sentences in one pilot session: every verb in three sentence types and in both
# agent-gender versions. Only the verb count is read from the data.
GV_S_TYPES_ADMINISTERED <- 3L
GV_VERSIONS             <- 2L
gv_trials <- function(lang) GV_VERSIONS * GV_S_TYPES_ADMINISTERED * ppv(lang, "n_verbs")
```

`\begin{abstract}\normalsize\noindent`{=latex} CLAPS asks whether affectedness, how strongly
an event affects the person or thing it happens to, matters more for the acceptability of
passive sentences than of active ones, and whether that pattern holds across languages. This
report evaluates by simulation what the planned sample of 80 participants per language can be
expected to deliver. Studies were simulated from the planned ordinal model fitted to the pilot
data in English, Turkish and Norwegian, with the effects drawn from each language's pilot
posterior, and each was analysed as the real data will be. The design has about 72 verbs per
language, with agent gender counterbalanced across participants. At 80 participants, the
primary prediction, that the affectedness slope is steeper for passives than for actives,
reaches a directional Bayes factor of 10 in `r pct0(da_rate("English", 80))` of simulated
English studies and `r pct0(da_rate("Norwegian", 80))` of Norwegian ones. Turkish, whose pilot
effect is weakest, reaches it in `r pct0(da_rate("Turkish", 80))`. The secondary prediction,
that the slope is also steeper for passives than for pseudo-passives, reaches a Bayes factor of
10 in `r pct0(da_rate("English", 80, "n_h2b"))` of English studies, while the Turkish pilot shows
the reverse pattern and the simulations inherit it. Affectedness is a property of the verb, so
the verbs bound the precision of every prediction, and beyond 80 participants the rates barely
change. The pooled cross-linguistic analysis, planned as a synthesis with no confirmatory claim,
reaches a Bayes factor of 10 in `r pct0(pooled_combined()$p)` of simulated studies with the three
pilot languages, at `r min(pooled_combined()$ns)` to `r max(pooled_combined()$ns)` participants
per language. If the seven languages measured so far take part and further languages differ as
much as those seven, it does so in `r pct0(kk("seven", 10))` of studies with ten languages and
`r pct0(kk("seven", 20))` with twenty. On the spread across the
`r num_word(n_measured_pp)` measured languages with a pseudo-passive, the pooled test of the secondary
prediction does so in `r pct0(kpp_rate("seven", 10))` and `r pct0(kpp_rate("seven", 20))`.
`\end{abstract}`{=latex}

<!-- The body starts on its own page. Left to flow, it followed the abstract down the
     title page and finished hard against the bottom margin. -->

`\newpage`{=latex}

# Background

CLAPS (A Crosslinguistic Approach to Passive Semantics) asks whether the acceptability of
passive and related sentence types tracks the semantic affectedness of the patient, whether that
relationship differs by sentence type and whether it holds across languages. Affectedness, the
degree to which a patient or theme undergoes a change of state or is causally impinged upon by
the action, is a core dimension of argument structure [@dowty1991; @beavers2011]. It has long
been argued to be the semantic heart of the passive, which carries the meaning that the surface
subject is in a state or circumstance characterised by the agent having acted upon it
[@pinkerLebeauxFrost1987]. Acceptability-judgement and comprehension studies with adults in
several languages support the view that passive acceptability is semantically structured
[@ambridgeBidgoodPineRowlandFreudenthal2016; @aryawibawaAmbridge2018;
@darmasetiyawanAmbridge2022; @liuAmbridge2021; @bidgoodPineRowlandAmbridge2020;
@ambridgeArnonBekman2023].

The evidence to date spans English, Indonesian, Balinese, Hebrew and Mandarin. A Bayesian
meta-analytic synthesis of those studies finds that affectedness raises the acceptability of
passives but not of pseudo-passives [@ambridgeArnonBekman2023]. In most of the five languages,
it raises passive acceptability more than active acceptability, although the size of that
advantage varies considerably and in Balinese it runs the other way. CLAPS adds Turkish and
Norwegian, for which its pilot supplies estimates.

Each language's data will be collected by its own team, at a sample size fixed in advance. This
report, prepared for the study's Stage 1 Registered Report, evaluates what the planned sample can
be expected to deliver. It does so through a design analysis [@schonbrodtWagenmakers2018] whose
simulated effects come from the CLAPS pilot.

# The Model and the Predictions

Each acceptability judgement is a rating from 1 to 7. The ratings are modelled with a Bayesian
cumulative-logit mixed-effects model, an ordinal model that respects the discreteness, ordering
and bounded range of rating data [@burknerVuorre2019]. Treating the same ratings as continuous
biases the regression coefficients and inflates error rates where floor or ceiling effects are
present [@liddellKruschke2018], and modelling them cumulatively recovers psycholinguistic effects
more accurately [@verissimo2021; @taylorRousseletScheepersSereno2023].

The model predicts acceptability from sentence type, affectedness and their interaction. Each
stimulus appears as an active, a passive and a pseudo-passive. In English, for example, these are
"The woman pushed the man", "The man was pushed by the woman" and "It was the man that the woman
pushed", the last a cleft that puts the patient first without the passive's morphology. The
passive is the reference level, and the active and the pseudo-passive are compared with it.

The pilot, whose sample Appendix A describes, shows the expected pattern in English and
Norwegian (@fig-pilot). In both, the affectedness slope is positive and steepest for the passive,
and flat or slightly negative for the active, as it is for the English pseudo-passive. In
Turkish, the passive slope is also steeper than the active slope, though by less, while the
pseudo-passive slope is at least as steep as the passive slope. The trends look much the same for
male and female agents, although in English the active-by-affectedness interaction differs
between agent genders. The design therefore balances the two agent genders across participants
(Appendix B).

```{r fig-pilot}
#| fig-cap: "Raw Pilot Acceptability by Affectedness, Sentence Type, Agent Gender and Language"
#| fig-width: 6.5
#| fig-pos: "H"
#| fig-align: center
#| out-width: "6.1in"

knitr::include_graphics(pilot_figure)
```

`\vspace{-8pt}`{=latex}

*Note.* Panels are by language across rows and by agent gender across columns. Points are
individual pilot ratings, jittered, and lines are per-sentence-type linear trends with 95%
confidence bands, which ignore the clustering of ratings within participants and verbs and so
understate the uncertainty. Affectedness is the score normed for each agent-gender version,
z-scored within language, and the model uses one score per verb, the mean of the two. The
Norwegian materials include no pseudo-passive.

`\vspace{12pt}`{=latex}

H1a is the passive affectedness slope, the effect of affectedness on the acceptability of
passives. It is a planned estimate, reported with its 95% credible interval but with no
prediction or test of its own. The theory's claim concerns the passive in particular, and
affectedness could raise the acceptability of sentences in general, so the tests compare the
passive slope with the active and pseudo-passive slopes. H1b, the primary prediction, holds that
the affectedness slope is steeper for passives than for actives, so that the active-by-affectedness
interaction is negative. H2b, the secondary prediction, holds that the slope is also steeper for
passives than for pseudo-passives, so that the pseudo-passive-by-affectedness interaction is
negative. The meta-analysis [@ambridgeArnonBekman2023] and the English pilot both put that
interaction below zero. H2a, the opposite direction, is not proposed as a hypothesis, although a
Bayes factor for H2b of 1/10 or less is evidence for it. H1b is tested in every language, and H2b
in every language with a pseudo-passive, which the Norwegian materials do not include.

# Design-Analysis Method

The design analysis simulates many studies from the cumulative-logit model planned for the real
data and analyses each as the real data will be analysed. The procedure stays within the
Bayesian framework throughout, because the confirmatory test is a Bayes factor. Its result is the
probability that a study reaches a Bayes factor of 10, a different quantity from the power to
return a significant *p*-value.

The per-language simulations take their effects from the pilot. For each language, the planned
model is fitted to the pilot ratings. On every simulated study, the fixed effects and ordinal
thresholds are drawn together from one draw of that fit's posterior, and the random-effect
covariance is held at its posterior mean. New participants and new verb deviations are then drawn
around the pilot's own verbs and their measured affectedness scores. Each reported rate is an
assurance, a detection rate averaged over the effect sizes the pilot leaves plausible, which
allows for the uncertainty of a single pilot's point estimate [@albersLakens2018]. Every pilot
language goes forward whatever its pilot showed, so the follow-up bias that Albers and Lakens
describe does not arise.

Published effect sizes are subject to publication bias and tend to overstate true effects
[@gelmanCarlin2014; @button2013], so design analyses anchored to them can be optimistic
[@vasishth2018]. In this report, published estimates enter only the multi-language synthesis, for
the four languages the pilot does not cover, and there they carry extra uncertainty to allow for
the change of protocol. Their average, `r mnum2(kpop("glossa_protocol", "mean_effect"))` per
standard deviation of affectedness, is close to the three pilots'
`r mnum2(kpop("claps_protocol", "mean_effect"))`, so they show no sign of inflation relative to
the CLAPS protocol.

Affectedness is a property of the verb. In the study, it will be rated in a separate norming task,
in which twenty further speakers of each language rate the affectedness of the patient in each
event, once for each agent-gender version, and a verb's score is the mean of its two version
scores. Each verb therefore carries one affectedness score, the same across participants and
sentence types, and every affectedness effect is estimated from differences between verbs. The
verbs and the spread of their scores bound the precision of those effects, however many
participants take part. This is the stimulus-sampling limit long recognised in psycholinguistics
[@clark1973] and quantified for designs with crossed random effects by Westfall et al.
[-@westfallKennyJudd2014; see also @brysbaertStevens2018; @juddWestfallKenny2017]. With unlimited
participants, the standard error of such an effect cannot fall below the by-verb standard
deviation of that effect divided by the square root of the verbs' summed squared affectedness
deviations, a limit that can be read from the fitted pilot models without simulation.

The analysis model has the maximal random-effects structure the design supports
[@barrLevyScheepersTily2013], with correlated by-participant slopes for sentence type,
affectedness and their interaction, and by-verb intercepts and sentence-type slopes. Because
affectedness does not vary within a verb, a by-verb affectedness slope cannot be estimated and is
omitted. Affectedness is centred and divided by twice its standard deviation within each language,
and effects are reported per standard deviation of affectedness. The priors are centred at zero on
the focal slopes, so they favour neither sign. The affectedness slope and the active interaction
take normal priors with a standard deviation of 0.5 on the model's scale, or 0.25 per standard
deviation of affectedness, and the pseudo-passive interaction takes 0.6, or 0.30 per standard
deviation. The other fixed effects take normal priors with a standard deviation of 1.5. The
ordinal thresholds take Student-*t* priors with three degrees of freedom and a scale of 2.5, and
the random-effect standard deviations Student-*t* priors with three degrees of freedom and a scale
of 1. Their correlations take an LKJ prior with a shape of 2 [@gelman2006;
@lewandowskiKurowickaJoe2009].

Evidence for each tested prediction is a directional Bayes factor, the posterior odds that the
coefficient carries the predicted sign divided by the same odds under the prior
[@moreyWagenmakers2014]. It is estimated from posterior draws and from prior draws sampled
alongside them. The two signs it compares are each represented by the zero-centred prior
restricted to that side, which is a half-normal. Under these symmetric priors, a Bayes factor of
10 corresponds to a posterior probability of `r prob2(bf_sign(10))` that the effect has the
predicted sign, and a Bayes factor of 1/10 or less is the same weight of evidence for the opposite
sign. A simulated study counts as a detection when its Bayes factor reaches 10.

Each stimulus exists in two versions, one with a woman agent and one with a man agent. Each
participant will see one version of each verb, so an English or Turkish session has
`r gv_trials("English") / 2` sentences, and the simulations describe that session. The versions
are balanced across participants by two complementary lists (Appendix B).

The per-language results rest on a dedicated set of `r da_reps` simulated studies at each of
`r and_list(da_ns)` participants per language, each scored for every prediction, at the pilot's
verb counts of `r ppv("English", "n_verbs")` in English and Turkish and
`r ppv("Norwegian", "n_verbs")` in Norwegian. `r grid_extra_txt` Each study is fitted with four
chains of 3,000 iterations at a target acceptance rate of .99 and a maximum treedepth of 12. After
1,000 warm-up iterations per chain, 8,000 post-warm-up draws remain. The confirmatory analysis will
use sixteen chains of 5,000 iterations with 2,000 of warm-up. The shorter chains add Monte-Carlo
noise to each simulated Bayes factor without shifting it systematically. Each per-language rate in
@tbl-power carries an exact 95% interval. Over `r da_reps` studies, a rate near 90% has a
Monte-Carlo standard error of about three percentage points, so differences of a few points
between neighbouring sample sizes are noise.

The simulations ran on the University of Oxford's Advanced Research Computing service, under R
4.4.2 with brms 2.21.0 and the cmdstanr backend on CmdStan 2.39.0, and the lockfile `renv.lock`
fixes the package versions. The analysis code, the simulation settings and the aggregated results
are openly available at <https://github.com/pablobernabeu/CLAPS-analysis>, and this report rebuilds
from those results without the raw pilot data, which are not yet public.

# Results

@tbl-power gives the detection rate for each prediction by language and sample size.

```{r tbl-power}
#| tbl-cap: "Detection Rates by Language and Participant Count"

# One entry per prediction and cell, as a rate with its exact interval. A language
# without a pseudo-passive carries NA in the H2 counts and prints as a dash.
rate_cell <- function(k, n) {
  if (is.na(k)) return("--")
  sprintf("%s %s", pct0(k / n), pct_ci(k, n))
}
tp <- ppd |>
  mutate(language = lang_factor(language)) |>
  arrange(language, n_participants)
tp <- tibble(
  Language       = ifelse(duplicated(tp$language), "", as.character(tp$language)),
  Participants   = tp$n_participants,
  H1b            = mapply(rate_cell, tp$n_h1b, tp$reps),
  H2b            = mapply(rate_cell, tp$n_h2b, tp$reps),
  `H2b reversed` = mapply(rate_cell, tp$n_h2a, tp$reps))
# The format is named because, left to choose it, kable under Quarto drops linesep, and
# the gap that closes each language would fall every fifth row instead.
kable(tp, format = "latex", booktabs = TRUE, longtable = TRUE, align = "lrrrr",
      linesep = c("", "", "\\addlinespace")) |>
  kableExtra::kable_styling(latex_options = c("repeat_header"))
```

`\vspace{-8pt}`{=latex}

*Note.* Each row rests on `r da_reps` simulated studies. Entries are the percentage of studies
reaching a directional Bayes factor of 10, with an exact 95% interval in brackets. H1b, the primary
prediction, is the active-by-affectedness interaction, and H2b, the secondary, is the
pseudo-passive-by-affectedness interaction, both predicted negative. The last column gives the
percentage of studies in which H2b's Bayes factor is 1/10 or less, evidence for the opposite
direction. The Norwegian materials include no pseudo-passive.

`\vspace{12pt}`{=latex}

## The Primary Prediction

At 80 participants, H1b reaches a Bayes factor of 10 in `r pct0(da_rate("English", 80))`
`r da_ci("English", 80)` of simulated English studies, `r pct0(da_rate("Norwegian", 80))`
`r da_ci("Norwegian", 80)` of Norwegian ones and `r pct0(da_rate("Turkish", 80))`
`r da_ci("Turkish", 80)` of Turkish ones. The Turkish pilot interaction is the weakest of the
three, at `r mnum2(ppv("Turkish", "interaction_per_sd"))` logits per standard deviation of
affectedness against `r mnum2(ppv("English", "interaction_per_sd"))` in English and
`r mnum2(ppv("Norwegian", "interaction_per_sd"))` in Norwegian.

The rates change little between 80 and 100 participants, as the verb limit predicts, so recruiting
beyond 80 would add little in any language. At that limit, averaged over the pilot posterior, the
interaction would reach a Bayes factor of 10 in `r ceil_txt("English")` of English studies,
`r ceil_txt("Norwegian")` of Norwegian ones and `r ceil_txt("Turkish")` of Turkish ones.

## The Secondary Prediction

In English, the pilot's pseudo-passive-by-affectedness interaction is
`r snum2(h2post("English", "per_sd_mean"))` logits per standard deviation of affectedness, with a
95% credible interval from `r snum2(h2post("English", "per_sd_lo95"))` to
`r snum2(h2post("English", "per_sd_hi95"))` and `r h2draws("English")` posterior draws below zero.
At 80 participants, H2b reaches a Bayes factor of 10 in `r pct0(da_rate("English", 80, "n_h2b"))`
`r da_ci("English", 80, "n_h2b")` of simulated studies, more often than H1b.

In Turkish, the pilot interaction runs the other way, at
`r snum2(h2post("Turkish", "per_sd_mean"))` per standard deviation, with a 95% credible interval
from `r snum2(h2post("Turkish", "per_sd_lo95"))` to `r snum2(h2post("Turkish", "per_sd_hi95"))`,
so the affectedness slope is steeper for Turkish pseudo-passives than for passives. The
simulations draw their effects from that posterior and inherit the reversal. At 80 participants,
`r pct0(da_rate("Turkish", 80, "n_h2a"))` of simulated Turkish studies give a Bayes factor for H2b
of 1/10 or less, and the remaining `r da("Turkish", 80, "n_h2_inconclusive")` of `r da_reps` are
inconclusive. A reversal of this kind would be a finding in its own right, showing that the
constraint on pseudo-passives differs between languages. The Stage 1 protocol will state in advance
how it will be interpreted.

The negative pseudo-passive interaction also appears in Balinese, Indonesian and Mandarin in the
per-language models of the meta-analysis. Those studies inform the planning, but the confirmatory
data will be collected under the CLAPS protocol alone.

## The Pooled Synthesis

The pooled model tests the primary prediction across all the languages together, and the secondary
prediction across those with a pseudo-passive. It has the per-language random-effects structure
with random slopes by language added, so its population-level interactions estimate the
cross-linguistic effects and each language can depart from them. It is planned as a synthesis with
no confirmatory claim, reported at whatever number of languages takes part, and the confirmatory
claims rest on the per-language tests.

With the three pilot languages alone, the pooled model reaches a Bayes factor of 10 on the active
interaction in `r pooled_combined()$txt` of `r pooled_combined()$n` simulated studies at
`r min(pooled_combined()$ns)` to `r max(pooled_combined()$ns)` participants per language. The rate
shows no trend over that range. It is held down by the variation of the effect between languages,
which neither verbs nor participants reduce and which three languages estimate poorly. These
simulations used the structure of the published meta-analysis model, which also carries a by-verb
affectedness slope.

Because more teams are expected to take part, the synthesis is also simulated with ten and twenty
languages. A two-stage version simulates each language's estimate of an interaction and combines
the estimates in a Bayesian random-effects meta-analysis. Its one free parameter, the prior that
the full model in effect places on the between-language spread, is calibrated against the full
model's three-language simulations. In these simulations, the seven languages measured so far take
part as CLAPS samples, with effects drawn from their estimates. They are the three CLAPS pilots,
and Balinese, Hebrew, Indonesian and Mandarin, the four languages of the meta-analysis other than
English. The remaining languages, up to ten or twenty, are drawn from a population of languages.
The pseudo-passive interaction is simulated in the same way in the languages that have a
pseudo-passive, `r num_word(n_measured_pp)` of the seven, and @tbl-klang gives both results.

```{r tbl-klang}
#| tbl-cap: "Pooled Detection Rates with the Measured Languages and Further Languages Drawn from a Population"

K_POP_LAB <- c(claps_protocol  = "The CLAPS pilots",
               seven           = "All measured languages",
               glossa_protocol = "The other meta-analysis languages")
kt <- tibble(
  set      = unname(K_POP_LAB[K_POPS]),
  tau_a    = num2h(vapply(K_POPS, kpop, numeric(1), col = "tau_median")),
  ten_a    = pct0_tex(vapply(K_POPS, kk, numeric(1), K = 10)),
  twenty_a = pct0_tex(vapply(K_POPS, kk, numeric(1), K = 20)),
  tau_p    = num2h(vapply(K_POPS, kpp_tau, numeric(1))),
  ten_p    = pct0_tex(vapply(K_POPS, kpp_rate, numeric(1), K = 10)),
  twenty_p = pct0_tex(vapply(K_POPS, kpp_rate, numeric(1), K = 20)))
# Typeset as LaTeX, like the table above, so that the two interactions can share a header
# row. Escaping is off for the tau symbol, which is why the percentages are escaped by hand.
kable(kt, format = "latex", booktabs = TRUE, longtable = TRUE, escape = FALSE,
      align = "lrrrrrr",
      col.names = c("New languages as different as", "Median $\\tau$", "Ten", "Twenty",
                    "Median $\\tau$", "Ten", "Twenty")) |>
  kableExtra::add_header_above(c(" " = 1, "Active" = 3, "Pseudo-passive" = 3))
```

`\vspace{-8pt}`{=latex}

*Note.* Each rate is the percentage of `r format(k_reps, big.mark = ",")` simulated studies in
which the pooled directional Bayes factor reaches 10, with ten or twenty languages in all and 80
participants and 72 verbs per language. $\tau$ is the between-language standard deviation of the
interaction per standard deviation of affectedness, given as its posterior median, and new
languages are drawn from its posterior. The active columns rest on the three pilots, the seven
measured languages and the four other meta-analysis languages, and the pseudo-passive columns on
the `r num_word(kpp_k("claps_protocol"))` pilots, `r num_word(n_measured_pp)` measured languages
and `r num_word(kpp_k("glossa_protocol"))` meta-analysis languages that have a pseudo-passive. Each
meta-analysis language carries an extra standard deviation of 0.15 for the change of protocol, or
0.40 in the first active row, whose small spread implies that the earlier estimates carry over
poorly. New languages have a pseudo-passive at the measured rate, so that `r kpp_n(10)` of 10 and
`r kpp_n(20)` of 20 carry the test, and giving every new language one moves no rate by more than
`r kpp_rule_gap_pts` points.

`\vspace{12pt}`{=latex}

The spread of the effect between languages decides the result, and it is not yet known. For the
active interaction, the three CLAPS pilots differ little, the four other meta-analysis languages
differ a great deal, and all seven together lie between. The larger spread may reflect real
differences between the languages, or the earlier studies' procedures, which differed in scale
length, verb count and affectedness norming. The spread across all seven languages is the planning
value used here, since three languages are too few to estimate a spread well and the larger spread
may be partly procedural. On that value, the pooled test reaches a Bayes factor of 10 in
`r pct0(kk("seven", 10))` of simulated studies with ten languages and in `r pct0(kk("seven", 20))`
with twenty. If new languages differ as little as the pilots, the rate falls to
`r pct0(kk("claps_protocol", 10))` at ten languages, because the four meta-analysis estimates then
carry over poorly, and rises to `r pct0(kk("claps_protocol", 20))` at twenty.

For the pseudo-passive interaction, the spread across the `r num_word(n_measured_pp)` measured
languages with a pseudo-passive is again the planning value. On it, the pooled test reaches a Bayes
factor of 10 in `r pct0(kpp_rate("seven", 10))` of simulated studies with ten languages and
`r pct0(kpp_rate("seven", 20))` with twenty. Here the pilots are the unfavourable set. The Turkish
reversal offsets the English effect, so new languages drawn around the two average close to zero,
and on their spread the rate is `r pct0(kpp_rate("claps_protocol", 10))` and
`r pct0(kpp_rate("claps_protocol", 20))`.

# The Sampling Plan

Every team will collect usable data from 80 participants, replacing any participant who is
excluded, a number fixed in advance because each team must state its sample size to its ethics
committee. Each language will use as close to 72 verbs as its supply of suitable verbs allows.

*Nature Human Behaviour* sets its 0.95 power requirement for frequentist analysis plans. For a
Bayesian plan, it asks that data collection continue until the Bayes factor is "at least 10 times in
favour of the experimental hypothesis over the null hypothesis (or vice versa)". Authors with
limited resources may instead state a maximum feasible sample, "sufficiently large that
inconclusive results at this sample size would nevertheless be an important message for the
field". CLAPS would take the second route for every hypothesis, with a Bayes factor of 10 as the
threshold in every language. Eighty is the most the collecting teams can provide. At that sample,
each language is close to the limit its verbs set, so further participants would add little to the
chance of reaching the threshold. An inconclusive result at 80 would still reflect an estimate with
nearly the precision the language's verbs allow. @tbl-design summarises the planned analyses and
their detection rates.

<!-- The table starts on a fresh page, after the paragraph that introduces it, so that the
     text after it fills the rest of that page. Left to flow, it began at the foot of a page
     with its header and one row. Check the rendered pages whenever the text above changes. -->

`\newpage`{=latex}

| Analysis | Coefficient and direction | Test | Detection rate |
|:---------------|:------------------------|:------------------------------|:------------------------------|
| H1b, primary | Active by affectedness, negative | Directional Bayes factor of 10, in each language | English `r pct0(da_rate("English", 80))`, Norwegian `r pct0(da_rate("Norwegian", 80))` and Turkish `r pct0(da_rate("Turkish", 80))` |
| H2b, secondary | Pseudo-passive by affectedness, negative | Directional Bayes factor of 10, in each language with a pseudo-passive | English `r pct0(da_rate("English", 80, "n_h2b"))`. Turkish is expected to reverse, with a Bayes factor of 1/10 or less in `r pct0(da_rate("Turkish", 80, "n_h2a"))` |
| H1a | Passive affectedness slope | Planned estimate, with its 95% credible interval | Not tested |
| Pooled synthesis | Cross-linguistic active-by-affectedness interaction, negative | Reported with its directional Bayes factor, as a synthesis with no confirmatory claim | `r pct0(pooled_combined()$p)` with the three pilot languages. Including the seven measured languages, `r pct0(kk("seven", 10))` with ten and `r pct0(kk("seven", 20))` with twenty |
| Pooled synthesis | Cross-linguistic pseudo-passive-by-affectedness interaction, negative | As above, in the languages with a pseudo-passive | Including the `r num_word(n_measured_pp)` measured languages with one, `r pct0(kpp_rate("seven", 10))` with ten and `r pct0(kpp_rate("seven", 20))` with twenty |

: Planned Analyses and Their Detection Rates {#tbl-design}

`\vspace{-8pt}`{=latex}

*Note.* The per-language rates are at 80 participants. The three-language pooled rate combines
simulations at `r min(pooled_combined()$ns)` to `r max(pooled_combined()$ns)` participants per
language. The ten- and twenty-language rates are at 80 participants per language, on the spread
across the measured languages that carry the interaction.

`\vspace{12pt}`{=latex}

The Bayes factor to be registered is directional [@moreyWagenmakers2014]. Each prediction is
represented by the half-normal given in the Design-Analysis Method, one of the forms the journal
names for representing a prediction, and its competitor is the reversed effect, represented by the
same half-normal on the other side. A Bayes factor of 10 is therefore strong evidence for the
prediction over its reversal, and one of 1/10 or less is equally strong evidence the other way.

Turkish would be registered by the same route. Its interaction reaches a Bayes factor of 10 in
`r pct0(da_rate("Turkish", 80))` of simulated studies and in at most `r ceil_txt("Turkish")` with
unlimited participants, so an inconclusive Turkish result would be informative for the reason given
above. Its pseudo-passive interaction will probably run the other way. Both Turkish predictions
would be registered as hypotheses, because the theory makes the same predictions for every
language, and leaving a language out on the strength of its pilot would amount to choosing
languages by their results.

# Limitations and Open Decisions

At 80 participants, no simulated fit has a divergent transition or a saturated treedepth.
Confirmatory fits must also reach an R-hat below 1.01 and bulk and tail effective sample sizes of at
least 400. With the shorter runs used for simulation, `r da("English", 80, "n_converged")` of
`r da_reps` English fits meet those thresholds, as do `r da("Norwegian", 80, "n_converged")`
Norwegian and `r da("Turkish", 80, "n_converged")` Turkish fits. Every rate counts all fits. Over
the passing fits alone, the rates are `r pct0(jval("English", 80, "p_h1b_converged"))` in English,
`r pct0(jval("Norwegian", 80, "p_h1b_converged"))` in Norwegian and
`r pct0(jval("Turkish", 80, "p_h1b_converged"))` in Turkish, each within what a random subset of
that size would show. The confirmatory chains will also be half as long again, and sixteen of them
will give six times the post-warm-up draws.

Fitted to the combined pilot data, 75,528 ratings over 210 verbs, the pooled model estimated the
interaction at about $-0.24$ logits per standard deviation of affectedness, with a 95% credible
interval from $-0.51$ to $+0.15$. That estimate is close to the average of the three pilot
estimates, `r mnum2(mean(pp$interaction_per_sd))`. The pooled model is harder to fit than the
per-language ones, however. It saturated the maximum treedepth on three quarters of its iterations,
and confirmatory fits must show none. A pooled fit to the collected data that misses the criteria
falls back to the two-stage synthesis of the per-language estimates, the analysis behind the ten-
and twenty-language rates above.

The multi-language figures rest on an assumed population of languages. The seven measured so far
are a convenience sample from five families in two macroareas, and which further languages will
join is not yet known.

Several decisions remain open. Turkish needs confirmation that its pseudo-passive is the same
construction as the English one. It has yet to be settled which teams will collect data, and
whether the Stage 2 paper waits for a minimum number of languages or the pooled synthesis is
reported whatever the number. Which teams take part matters most for the pooled pseudo-passive
test, since `r num_word(kpp_k("glossa_protocol"))` of the `r num_word(n_measured_pp)` languages
behind it come from the meta-analysis. The pooled model also has to specify whether a verb and its
translations share one by-verb random effect. The maximum-feasible route for every hypothesis, the
treatment of H1a as a planned estimate, registering only the negative direction of H2b and the
registration of both Turkish predictions as hypotheses await the collaborators' agreement. None of
these decisions changes the per-language sample size.

{{< pagebreak >}}

# References {.unnumbered}

::: {#refs}
:::

{{< pagebreak >}}

```{=latex}
% Appendix tables carry their own letter, as APA 7 asks, so the body
% sequence is not interrupted by tables the reader meets later.
\setcounter{table}{0}
\renewcommand{\thetable}{A\arabic{table}}
\begin{center}
{\large\bfseries Appendix A}\\[3pt]
{\bfseries Composition of the Pilot Sample}
\end{center}
```

This appendix describes the pilot sample whose ratings give the descriptive trends in @fig-pilot
and to which the planned model is fitted for the simulations. @tbl-pilot gives the counts, read
directly from the harmonised pilot data.

```{r tbl-pilot}
#| tbl-cap: "Composition of the Harmonised Pilot Sample, by Language"

comp <- pilot_composition
kable(comp, align = "lrrrrr", format.args = list(big.mark = ","))
```

`\vspace{-8pt}`{=latex}

*Note.* Items are verb-by-agent-gender combinations, so the item count is twice the verb count.
Every pilot participant saw both versions of every verb, whereas in the study itself each
participant will see one version of each verb (Appendix B). The Norwegian materials include no
pseudo-passive, hence only two sentence types. Norwegian sessions also included a synthetic
passive, which the preregistered analysis excludes, so a Norwegian participant rated one and a half
times as many sentences as this table counts.

`\vspace{12pt}`{=latex}

{{< pagebreak >}}

```{=latex}
% Appendix tables carry their own letter, as APA 7 asks, so the body
% sequence is not interrupted by tables the reader meets later.
\setcounter{table}{0}
\renewcommand{\thetable}{B\arabic{table}}
\begin{center}
{\large\bfseries Appendix B}\\[3pt]
{\bfseries The Agent-Gender Lists}
\end{center}
```

This appendix sets out how the two agent-gender versions of each stimulus are balanced across
participants, and why. Each participant will see one version of each verb, so the balance has to
come from the assignment of versions to participants.

Balance matters because agent gender moves the interaction in English. In a linear mixed model of
the 1-to-7 pilot ratings, used here only to gauge the relative size of the gender term, with agent
gender coded plus or minus a half, the English interaction is
`r mnum2(gv_per_sd(gvev("English", GV_TERM_H1B, "h1b_estimate")))` rating points per standard
deviation of affectedness and its gender-by-active-by-affectedness term is
`r snum3(gv_per_sd(gvev("English", GV_TERM_H1B)))` (*SE* =
`r num3(gv_per_sd(gvev("English", GV_TERM_H1B, "se")))`). The interaction is therefore markedly
stronger for man-agent items than for woman-agent ones. A design in which every participant saw
only woman-agent versions, or only man-agent ones, would move the English interaction by about
`r pct0(gv_shift("English"))`, against `r pct0(gv_shift("Turkish"))` in Turkish and
`r pct0(gv_shift("Norwegian"))` in Norwegian.

Two complementary lists remove that shift without lengthening the session. The verbs are split into
two sets of `r ppv("English", "n_verbs") / 2`, or `r ppv("Norwegian", "n_verbs") / 2` in Norwegian,
matched on affectedness and verb class, and the participants into two lists. List 1 shows the first
set with woman agents and the second with man agents, and list 2 the reverse. Each participant
therefore sees both agent genders equally often, each verb is rated in each version by half the
sample, and the estimand remains the slope averaged over agent gender. Because both lists show both
agent genders on matched verb sets, an unequal split between the lists leaves the interaction
essentially unchanged. Even so, assigning participants to the lists alternately, and replacing
exclusions within the same list, keeps each verb's two versions evenly rated. Splitting the
agent-gender versions across participants in this way follows Ambridge et al.
[-@ambridgeBidgoodPineRowlandFreudenthal2016, Study 3] and the Hebrew study of Ambridge et al.
[-@ambridgeArnonBekman2023].

Both versions stay in the affectedness rating task, which asks for one judgement per verb and agent
gender and is short. Halving it would leave each agent-specific score resting on half as many
raters and would attenuate every slope estimated from those scores.
